---
title: "Differential_Expression_Analysis_C1"
author: "Hamish Williams"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 8,  # Set default plot width
  fig.height = 6,  # Set default plot height
  dev = "png",
  echo = T,
  warning = F,
  error = F# Set default device to PNG for Word compatibility
)

# Specify the default figure size for the whole document
options(
  repr.plot.width = 6,  # Set default plot width
  repr.plot.height = 4.5  # Set default plot height
)
```

# Outline of Document

This document will serve as a culmination of all of the relevant methods and results relating specifically to the RNA-seq and DESeq2 analyses performed on both experiment 1 & 2 data. I aim to write this document as a type of "scientific journal", which will be more formal and written as a draft type of word document that will be integrated into the main chapter(s) document(s).

I will write each section as it appears, explaining in more detail the steps taken and the thought processes behind it. Initially I'll write the sections as they appear in the initial markdown script, but may gradually change and edit the sections to be split more clearly into an overarching methods and results section.

# Set-up

There are several steps necessary before starting the analyses from my script. This includes:

-   Installing and Loading packages
-   Loading Required Data

## Installing & Loading Packages

There are over a dozen packages used for this stage of the analyses, and as such, it can be a little tedious to have to manually install and load all of the required packages needed. Here, I have created a custom function which will check if the packages in the required list are installed or not, install them through the normal means (or via Biocmanager where applicable), and the subsequently load th packages into R. We repeat the function on our package list to ensure that all of the packages are installed and loaded correctly.

```{r Install_and_Load_Packages, warning=F, message=F}
# Create function to check a package list
  # This function will install bioconductor packages also without needing to do the whole BiocManager::install(...), which is nice
check_packages <- function(pkg_list) {
  for (pkg in pkg_list) {
    if (!require(pkg, character.only = TRUE)) {
      print(paste(pkg, "is not installed, installing package..."))
      
      # Check if the package requires BiocManager
      if (pkg %in% c("gage", "RUVSeq", "gprofiler2","DESeq2","PCAtools", "ComplexHeatmap")) {
        if (!require("BiocManager", quietly = TRUE)) {
          install.packages("BiocManager")
          BiocManager::install(version = "3.17")
        }
        BiocManager::install(pkg, force = TRUE)
      } else {
        install.packages(pkg)
      }
      
    } else {
      suppressMessages(library(pkg, character.only = TRUE))
      print(paste(pkg, "is installed and loaded"))
    }
  }
}

# Generate a package list: 
pkg_list<-c("tidyverse", #manipulation
            "DESeq2", # main differential expression analysis tool
            "tibble", # tables
            "stats", # stats
            "ggplot2", # plotting
            "pheatmap", # plotting
            "dplyr", # manipulation 
            "tidyr", # manipulation
            "ggfortify", #plotting
            "corrplot", # plotting
            "gprofiler2", #
            "knitr", # manipulation
            "ggforce", # pltting
            "gridExtra", # plotting
            "ggrepel", # plotting
            "VennDiagram", # plotting
            "ggVennDiagram", # plotting
            "PCAtools", # plotting
            "cowplot", # plotting
            "apeglm", # LFC shrinking method option
            "EnhancedVolcano", # plotting
            "ashr", # LFC shrinking method option
            "ComplexHeatmap"
            ) # plotting

# Execute function:
check_packages(pkg_list)
```
# Import Data

We need A couple of different Data files to begin our analyses. These data files include:

1.  Our Count Data

-   This Data is the raw output data from our featureCounts script from the HPC UNIX system. It is essentially a very large file which contains count data for genes in the *Actinia equina* genome where they are aligned to a high proportion.
-   We separate this data into their experimental groups, as the analyses needs to be completed separately. I have perhaps performed this separation early, which bulks the code slightly, however I think having a separated working script makes things easier to keep track of.

2.  Our ColData file

-   This is a=our experimental conditions file. It is a small file which should primarily contain the sample IDs (consistent with the Count Data), and also the categorical variables in which the experiment was completed (*i.e. Control vs Treatment, Experiment 1 vs. Experiment 2*).
-   Again, separating this file so that we can work on our two experimental data sets separately.

```{r Import_Data}
# Import Count Data
counts <- as.matrix(read.table("./Chapter 1/Data/A_Equina_Counts_redo.tsv", header = T, sep = '\t'))
# Extract Counts only data
Counts_only <- subset(counts, select = c(-Chr,-Start,-End,-Strand,-Length))
# Subset counts data into each experiment group:
counts_exp1 <- subset(counts[,c(1:5,6:12,27:33)])
counts_exp2<- subset(counts[,c(1:5,13:26)])

# Import colData
colData <- read.table("data/colData.tsv", header = T, sep = '\t', 
                      stringsAsFactors = TRUE)
# Subset into Experiment groups:
colData_exp1 <- subset(colData[c(1:7,22:28),])
colData_exp2<- subset(colData[8:21,])
```
