---
title: "Differential_Expression_Analysis_C1"
author: "Hamish Williams"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 8,  # Set default plot width
  fig.height = 6,  # Set default plot height
  dev = "png",
  echo = T,
  warning = F,
  error = F# Set default device to PNG for Word compatibility
)

# Specify the default figure size for the whole document
options(
  repr.plot.width = 6,  # Set default plot width
  repr.plot.height = 4.5  # Set default plot height
)
```

# Outline of Document

This document will serve as a culmination of all of the relevant methods and results relating specifically to the RNA-seq and DESeq2 analyses performed on both experiment 1 & 2 data. I aim to write this document as a type of "scientific journal", which will be more formal and written as a draft type of word document that will be integrated into the main chapter(s) document(s).

I will write each section as it appears, explaining in more detail the steps taken and the thought processes behind it. Initially I'll write the sections as they appear in the initial markdown script, but may gradually change and edit the sections to be split more clearly into an overarching methods and results section.

# Set-up

There are several steps necessary before starting the analyses from my script. This includes:

-   Installing and Loading packages
-   Loading Required Data

## Installing & Loading Packages

There are over a dozen packages used for this stage of the analyses, and as such, it can be a little tedious to have to manually install and load all of the required packages needed. Here, I have created a custom function which will check if the packages in the required list are installed or not, install them through the normal means (or via Biocmanager where applicable), and the subsequently load th packages into R. We repeat the function on our package list to ensure that all of the packages are installed and loaded correctly.

```{r Install_and_Load_Packages, warning=F, message=F}
# Create function to check a package list
  # This function will install bioconductor packages also without needing to do the whole BiocManager::install(...), which is nice
check_packages <- function(pkg_list) {
  for (pkg in pkg_list) {
    if (!require(pkg, character.only = TRUE)) {
      print(paste(pkg, "is not installed, installing package..."))
      
      # Check if the package requires BiocManager
      if (pkg %in% c("gage", "RUVSeq", "gprofiler2","DESeq2","PCAtools", "ComplexHeatmap")) {
        if (!require("BiocManager", quietly = TRUE)) {
          install.packages("BiocManager")
          BiocManager::install(version = "3.17")
        }
        BiocManager::install(pkg, force = TRUE)
      } else {
        install.packages(pkg)
      }
      
    } else {
      suppressMessages(library(pkg, character.only = TRUE))
      print(paste(pkg, "is installed and loaded"))
    }
  }
}

# Generate a package list: 
pkg_list<-c("tidyverse", #manipulation
            "DESeq2", # main differential expression analysis tool
            "tibble", # tables
            "stats", # stats
            "ggplot2", # plotting
            "pheatmap", # plotting
            "dplyr", # manipulation 
            "tidyr", # manipulation
            "ggfortify", #plotting
            "corrplot", # plotting
            "gprofiler2", #
            "knitr", # manipulation
            "ggforce", # pltting
            "gridExtra", # plotting
            "ggrepel", # plotting
            "VennDiagram", # plotting
            "ggVennDiagram", # plotting
            "PCAtools", # plotting
            "cowplot", # plotting
            "apeglm", # LFC shrinking method option
            "EnhancedVolcano", # plotting
            "ashr", # LFC shrinking method option
            "ComplexHeatmap",
            "RColorBrewer"
            ) # plotting

# Execute function:
check_packages(pkg_list)
```
# Import Data

We need A couple of different Data files to begin our analyses. These data files include:

## 1. Count Data

-   This Data is the raw output data from our featureCounts script from the HPC UNIX system. It is essentially a very large file which contains count data for genes in the *Actinia equina* genome where they are aligned to a high proportion.
-   We separate this data into their experimental groups, as the analyses needs to be completed separately. I have perhaps performed this separation early, which bulks the code slightly, however I think having a separated working script makes things easier to keep track of.

## 2. ColData file

-   This is a=our experimental conditions file. It is a small file which should primarily contain the sample IDs (consistent with the Count Data), and also the categorical variables in which the experiment was completed (*i.e. Control vs Treatment, Experiment 1 vs. Experiment 2*).
-   Again, separating this file so that we can work on our two experimental data sets separately.

```{r Import_Data}
# Import Count Data
getwd()
  # .rmd file is in .../Chapter 1/Scripts/
counts <- as.matrix(read.table("../Data/A_Equina_Counts_redo.tsv", header = T, sep = '\t'))
# Extract Counts only data
Counts_only <- subset(counts, select = c(-Chr,-Start,-End,-Strand,-Length))
# Subset counts data into each experiment group:
counts_exp1 <- subset(counts[,c(1:5,6:12,27:33)])
counts_exp2<- subset(counts[,c(1:5,13:26)])

# Import colData
colData <- read.table("../Data/colData.tsv", header = T, sep = '\t', 
                      stringsAsFactors = TRUE)
# Subset into Experiment groups:
colData_exp1 <- subset(colData[c(1:7,22:28),])
colData_exp2<- subset(colData[8:21,])
```

# Processing the Raw Data

Before any analyses is performed from our data, we need to process it to an extent so that we can extract useful information from our Count data, as well as performing some diagnostic analyses on our dataset.

We are going to generate several new data frames and vectors in order to do this, which includes:

## 1.  CPM - Counts per Million

-   A normalization method used to scale the raw counts to account for differences in library size or sequencing depth between samples.
-   Calculated by taking raw counts for each gene, dividing by the total number of counts in the sample. then multiplying by one million.

$$
    \text{CPM}_{ij} = \left( \frac{\text{Count}_{ij}}{\text{Library size}_j} \right) \times 10^6
$$

```{r CPM}
# Apply CPM Formula
cpm <- {apply(subset(counts, select = c(-Chr,-Start,-End,-Strand,-Length)), 2, 
                function(x) as.numeric(x)/sum(as.numeric(x)) * 10^6)}
cpm_exp1 <- {apply(subset(counts_exp1, select = c(-Chr,-Start,-End,-Strand,-Length)), 2, 
              function(x) as.numeric(x)/sum(as.numeric(x)) * 10^6)}
cpm_exp2 <- {apply(subset(counts_exp2, select = c(-Chr,-Start,-End,-Strand,-Length)), 2, 
              function(x) as.numeric(x)/sum(as.numeric(x)) * 10^6)}

# add back row.names information
counts_rownames <- as.vector(row.names(counts))
rownames(cpm) <- counts_rownames
rownames(cpm_exp1) <- counts_rownames
rownames(cpm_exp2) <- counts_rownames

# Check Values
colSums(cpm) # = 10^6 for ALL
colSums(cpm_exp1) # = 10^6 for ALL
colSums(cpm_exp2) # = 10^6 for ALL
```

## 2.  Gene Lengths

-   A simple extraction where we obtain the lengths of the genes.

```{r GeneLengths}
geneLengths <- as.vector(subset(counts, select = c(Length)))
geneLengths <- as.vector(as.numeric(geneLengths))
```

## 3.  RPKM - Reads Per Kilobase Million (mapped reads)

-   A normalization method used to normalize gene expression levels based on both the length of the gene and the total number of mapped reads in the sample.

$$
RPKM_{ij} = \frac{10^9 \times \text{Read count}_{ij}}{\text{Gene length}_i \times \text{Total mapped reads}_j} 
$$

```{r RPKM}
rpkm <- apply(X = subset(counts, select = c(-Chr,-Start,-End,-Strand,-Length)),
                 MARGIN = 2, 
                 FUN = function(x) {
                   10^9 * as.numeric(x) / geneLengths / sum(as.numeric(x))
                 })
rpkm_exp1 <- apply(X = subset(counts_exp1, select = c(-Chr,-Start,-End,-Strand,-Length)),
               MARGIN = 2, 
               FUN = function(x) {
                 10^9 * as.numeric(x) / geneLengths / sum(as.numeric(x))
               })
rpkm_exp2 <- apply(X = subset(counts_exp2, select = c(-Chr,-Start,-End,-Strand,-Length)),
               MARGIN = 2, 
               FUN = function(x) {
                 10^9 * as.numeric(x) / geneLengths / sum(as.numeric(x))
               })

# Add back row names
rownames(rpkm) <- counts_rownames
rownames(rpkm_exp1) <- counts_rownames
rownames(rpkm_exp2) <- counts_rownames

# Check values
colSums(rpkm) # Different values
colSums(rpkm_exp1) # Different values
colSums(rpkm_exp2) # Different values
```

## 4.  RPK - Reads Per Kilobase

-   A non-standard term used in RNA-seq. Mainly used to calculate the TPM after.

$$
RPK_{ij} = \frac{\text{Read count}_{ij}}{(\frac{\text{Gene length}_i}{1000})}
$$

```{r RPK}
rpk <- apply(subset(counts, select = c(-Chr,-Start,-End,-Strand,-Length)), 2, 
                 function(x) as.numeric(x)/(geneLengths/1000))

rpk_exp1 <- apply(subset(counts_exp1, select = c(-Chr,-Start,-End,-Strand,-Length)), 2, 
              function(x) as.numeric(x)/(geneLengths/1000))

rpk_exp2 <- apply(subset(counts_exp2, select = c(-Chr,-Start,-End,-Strand,-Length)), 2, 
              function(x) as.numeric(x)/(geneLengths/1000))

# Add back row names
rownames(rpk) <- counts_rownames
rownames(rpk_exp1) <- counts_rownames
rownames(rpk_exp2) <- counts_rownames
```

## 5.  TPM - Transcripts per Million

-   Normalization method designed to provide a measure of gene expression that accounts for both the length of the gene and the total number of transcripts in the sample.\
-   mainly useful for comparing gene expression levels between samples or conditions

$$
\text{TPM}_{ij} = \left( \frac{\text{Transcript count}_{ij}}{\text{Gene length}_i} \right) \times \left( \sum_{k=1}^n \frac{\text{Transcript count}_{kj}}{\text{Gene length}_k} \right)^{-1} \times 10^6
$$

```{r TPM}
tpm <- apply(rpk, 2, function(x) as.numeric(x) / sum(as.numeric(x)) * 10^6)
tpm_exp1 <- apply(rpk_exp1, 2, function(x) as.numeric(x) / sum(as.numeric(x)) * 10^6)
tpm_exp2 <- apply(rpk_exp2, 2, function(x) as.numeric(x) / sum(as.numeric(x)) * 10^6)

# Row names
rownames(tpm) <- counts_rownames
rownames(tpm_exp1) <- counts_rownames
rownames(tpm_exp2) <- counts_rownames

# Check Values
colSums(tpm) # = 10^6 for ALL
colSums(tpm_exp1) # = 10^6 for ALL
colSums(tpm_exp2) # = 10^6 for ALL
```

## 6.  Variance

-   A simple calculation of the variance of a given dataframe or set of values. - in this instance, we use the TPM values to get an idea of our variance.

$$
V_i = \text{var}(\text{tpm}_{i1}, \text{tpm}_{i2}, \ldots, \text{tpm}_{in})
$$

```{r Variance}
V <- apply(tpm, 1, var)
V_exp1 <- apply(tpm_exp1, 1, var)
write.csv(V_exp1, file = "../Results/Intermediary Files/variance_exp1.csv")
V_exp2 <- apply(tpm_exp2, 1, var)
write.csv(V_exp2, file = "../Results/Intermediary Files/variance_exp2.csv")
```

## 7.  Subset data frames for Diagnostics

-   We use the top 100 most variable genes in the genome to run some diagnostic plots to get an idea of how our data is structured and how the raw data is presented.
-   From this we can determine some initial thoughts about what patterns seems to emerge, identify potential covariance with treatments or variables, and draw PCA plots of our pre-Differential-expression-data.

```{r top_100_genes}
# Subset the top 100 most variable genes for generating diagnostic plots
selectedGenes <- names(V[order(V, decreasing = T)][1:100])
selectedGenes_exp1 <- names(V_exp1[order(V_exp1, decreasing = T)][1:100])
selectedGenes_exp2 <- names(V_exp2[order(V_exp2, decreasing = T)][1:100])
```

# Diagnostic Plots

Now that we have the initial data processed and calculated some parameters, we can begin to explore the Count and colData files to identify related-ness and co-linearlity from our samples and treatment groups.

## Heatmap

We can plot a heatmap of our top 100 most variable genes annotated with our **colData**, using this we can start to explore potential patterns in our samples between treatments.

```{r Heatmap}
# remove redundant columns from colData
colData_heatmap <- subset(colData[,-c(1,2,4,5,6)])
colData_heatmap_exp1 <- subset(colData_heatmap[c(1:7,22:28),])
colData_heatmap_exp2<- subset(colData_heatmap[8:21,])
# annotation colours
ann_colours <- list(Group = c("X" = "grey",
                              "A" = "red",
                              "B" = "blue",
                              "C" = "green"),
                    Clone = c("Clone10" = "pink",
                              "Clone11" = "red",
                              "Clone15" = "orange",
                              "Clone16" = "yellow2",
                              "Clone2" = "yellow4",
                              "Clone5" = "lightgreen",
                              "Clone9" = "green",
                              "Clone6" = "green4",
                              "Clone12" = "turquoise",
                              "Clone14" = "lightblue4",
                              "Clone20" = "lightblue",
                              "Clone3" = "blue",
                              "Clone19" = "violet",
                              "Clone8" = "purple"))

# plot
pheatmap(tpm[selectedGenes,], scale = 'row', 
         show_rownames = FALSE, 
         annotation_col = colData_heatmap,
         clustering_method = 'ward.D',
         clustering_distance_cols = 'euclidean',
         clustering_distance_rows = 'euclidean',
         border_color = "black",
         number_color = "black",
         fontsize_number = 8,
         col = brewer.pal(10, 'RdYlGn'),
         cutree_rows = 2, cutree_cols = 4,
         annotation_colors = ann_colours,
         legend_breaks = c(-4, 0, 4), # legend customisation
         legend_labels = c("Low", "Medium", "High") # legend customisation
         )

# save
out_dir <- "../Results/Figures/Heatmaps"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
dev.copy(png, file = "../Results/Figures/Heatmaps/heatmap_all_samples.png")
dev.off()

# heatmap of acute experiment only
pheatmap(tpm_exp1[selectedGenes_exp1,], scale = 'row', 
         show_rownames = FALSE, 
         annotation_col = colData_heatmap_exp1,
         clustering_method = 'ward.D',
         clustering_distance_cols = 'euclidean',
         clustering_distance_rows = 'euclidean',
         border_color = "black",
         number_color = "black",
         fontsize_number = 8,
         col = brewer.pal(10, 'RdYlGn'),
         cutree_rows = 2, cutree_cols = 4,
         annotation_colors = ann_colours,
         legend_breaks = c(-4, 0, 4), # legend customisation
         legend_labels = c("Low", "Medium", "High") # legend customisation
         )
dev.copy(png, file = "../Results/Figures/Heatmaps/heatmap_acute_only.png")
dev.off()

# heatmap of primed xperiment
pheatmap(tpm_exp2[selectedGenes_exp2,], scale = 'row', 
         show_rownames = FALSE, 
         annotation_col = colData_heatmap_exp2,
         clustering_method = 'ward.D',
         clustering_distance_cols = 'euclidean',
         clustering_distance_rows = 'euclidean',
         border_color = "black",
         number_color = "black",
         fontsize_number = 8,
         col = brewer.pal(10, 'RdYlGn'),
         cutree_rows = 2, cutree_cols = 4,
         annotation_colors = ann_colours,
         legend_breaks = c(-4, 0, 4), # legend customisation
         legend_labels = c("Low", "Medium", "High") # legend customisation
         )

# Save the Image
dev.copy(png, file = "../Results/Figures/Heatmaps/heatmap_primed_only.png")
dev.off()
```