---
title: "Methylation Enrichment Analysis - Final Methods"
author: "Hamish Williams"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

Performing over representation analysis on the DMR gene lists for both experiments to determine if there is some statistcially enriched terms and functions. This analysis is only going to contribute to highlighting if particular functions are being enriched, and not to remove genes associated to DMRs away from the results and discussion. The results of the DMR anlaysis have already gone through a lengthy and rigouros bout of statistical analyses to get to a final list of DMRs, hence this analysis is only to highlight results which may be of greater importance to the physiological response of actinia equina to their experimental contexts.

# Installing and Loading required packages

```{r check_packages_ClusterProfilerR, message=FALSE, warning=FALSE, results='hide'}
# Build a function which will automatically detect installed packages and install required packages
check_packages_ClusterProfileR <- function(pkg_list) {
  for (pkg in pkg_list) {
    if (!require(pkg, character.only = TRUE)) {
      print(paste(pkg, "is not installed, installing package..."))
      install.packages(pkg) # installs regular R packages
      BiocManager::install(pkg) # installs packages requiring BiocManager
    } else {
      suppressMessages(library(pkg, character.only = TRUE))
      print(paste(pkg, "is installed and loaded"))
    }
  }
}

# Generate a dataframe with a list of packages required

pkg_list_ClusterProfileR<-c("BiocManager", # Have this first to ensure BiocManager packages will be installed
                            "tidyverse", 
                            "tibble", 
                            "stats",
                            "ggplot2", 
                            "dplyr", 
                            "tidyr", 
                            "ggfortify", 
                            "knitr",
                            "clusterProfiler",
                            "AnnotationForge",
                            "enrichplot",
                            "data.table",
                            "stats",
                            "plotly",
                            "ggVennDiagram",
                            "VennDiagram",
                            "limma",
                            "ggridges",
                            "pathview",
                            "KEGGREST",
                            "stringr",
                            "gridExtra",
                            "ggside",
                            "gghalves",
                            "flextable",
                            "officer",
                            "ggh4x"
                            )


# Run the code
  # NOTE: this code should be run twice, double checking is always good!

# Run Function::
check_packages_ClusterProfileR(pkg_list_ClusterProfileR)
# You will need to answer "no" when asked if you want to update packages, otherwise it will take a while to load

# Double Check everything has loaded:
check_packages_ClusterProfileR(pkg_list_ClusterProfileR)
```

# convertFractionToNumeric

Used to convert gene ratios in final analyses to numerical values for generating figures

```{r convertFractionToNumeric, message=FALSE, warning=FALSE}
convertFractionToNumeric <- function(fraction) {
  fraction_parts <- strsplit(fraction, "/")[[1]]
  numerator <- as.numeric(fraction_parts[1])
  denominator <- as.numeric(fraction_parts[2])
  numeric_value <- numerator / denominator
  return(numeric_value)
}


# Make function to capture methylation data for each gene in the GSEA results
get_methDiff_data <- function(core_enrichment, meth_data) {
  genes <- unlist(strsplit(core_enrichment, "/"))
  lfc_values <- meth_data$meth.diff[match(genes, meth_data$gene_id_labelling)]
  paste(lfc_values, collapse = "/")
}
```

# Importing data:

```{r import_data}
# load in the gene lists for experiments 1 & 2;
exp1_DMR_list <- read.table("C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_2/data/temp_storage/exp1_DMRs_gene_id_matched.txt", header = T, sep = '\t')
exp1_DMR_list <- subset(exp1_DMR_list[exp1_DMR_list$p_fdr <= 0.1,])
exp1_DMR_list<- as.vector(exp1_DMR_list[,19])
exp2_DMR_list <- read.table("C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_2/data/temp_storage/exp2_DMRs_gene_id_matched.txt", header = T, sep = '\t')
exp2_DMR_list <- subset(exp2_DMR_list[exp2_DMR_list$p_fdr <= 0.1,])
exp2_DMR_list<- as.vector(exp2_DMR_list[,19])

## Merge Blastp and GO terms files together
BP_MF_Terms <- read.table(
  "C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_2/data/temp_storage/UNIPROT_results.txt",
  header = TRUE,
  sep = "\t",
  stringsAsFactors = FALSE,
  quote = ""
)

matching_file <- read.table(
  "C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_2/data/temp_storage/genome_matching_file.txt",
  header = FALSE,
  sep = "\t",
  stringsAsFactors = FALSE
)

# Change the name of the columns to be identical ----
{
  names(matching_file)[names(matching_file) == "V13"] <- "ID"
  names(matching_file)[names(matching_file) == "V1"] <- "GENE"
  names(BP_MF_Terms)[names(BP_MF_Terms) == "From"] <- "ID"
  names(BP_MF_Terms)[names(BP_MF_Terms) == "Gene.Names"] <- "GENENAMES"
  names(BP_MF_Terms)[names(BP_MF_Terms) == "Gene.Ontology..biological.process."] <- "GO:BP"
  names(BP_MF_Terms)[names(BP_MF_Terms) == "Gene.Ontology..molecular.function."] <- "GO:MF"
}

GO_lookup <- BP_MF_Terms %>%
  group_by(ID) %>%
  summarise(
    has_GO = any(
      (!is.na(`GO:BP`) & trimws(`GO:BP`) != "") |
      (!is.na(`GO:MF`) & trimws(`GO:MF`) != "")
    ),
    .groups = "drop"
  )

matching_file <- matching_file %>%
  left_join(GO_lookup, by = "ID") %>%
  filter(has_GO) %>%
  group_by(GENE) %>%
  slice_min(order_by = as.numeric(V11), n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(-has_GO)

matching_file <- unique(matching_file)
  
# Merge all matched BLAST rows with UniProt GO annotations
merged_df <- merge(matching_file, BP_MF_Terms, by = "ID") # genrates duplicates when matching
merged_df <- unique(merged_df)

# -------------------------------
# Helper function:
# expands one GO column so that each GO term becomes its own row
# -------------------------------
expand_go_column_mem <- function(
  df,
  go_col,
  gene_col = "GENE",
  id_col = "ID",
  deduplicate = FALSE
) {
  empty_out <- data.frame(
    GENE = character(),
    ID   = character(),
    GO   = character(),
    TERM = character(),
    stringsAsFactors = FALSE
  )

  # Check required columns exist
  if (!all(c(go_col, gene_col, id_col) %in% names(df))) {
    stop("One or more specified columns do not exist in the data frame.")
  }

  go_text  <- as.character(df[[go_col]])
  gene_vec <- as.character(df[[gene_col]])
  id_vec   <- as.character(df[[id_col]])

  # Keep only usable rows up front
  keep <- !is.na(go_text) &
          trimws(go_text) != "" &
          !is.na(gene_vec) &
          !is.na(id_vec)

  if (!any(keep)) {
    return(empty_out)
  }

  go_text  <- go_text[keep]
  gene_vec <- gene_vec[keep]
  id_vec   <- id_vec[keep]

  # Clean up extracted GO term descriptions
  clean_term <- function(x) {
    x <- trimws(x)

    # Normalise curly quotes to straight quotes
    x <- gsub("[\u2018\u2019]", "'", x, perl = TRUE)
    x <- gsub("[\u201C\u201D]", "\"", x, perl = TRUE)

    # Remove surrounding quotes/backticks
    x <- gsub("^[[:space:]]*[\"'`]+", "", x, perl = TRUE)
    x <- gsub("[\"'`]+[[:space:]]*$", "", x, perl = TRUE)

    # Remove leading/trailing separators left over from splitting
    x <- gsub("^[[:space:]]*[,;:]+[[:space:]]*", "", x, perl = TRUE)
    x <- gsub("[[:space:]]*[,;:]+[[:space:]]*$", "", x, perl = TRUE)

    # Final trim
    x <- trimws(x)

    x
  }

  # Extract GO-term pairs from one cell
  extract_go_pairs <- function(x) {
    x <- as.character(x)
    if (is.na(x) || trimws(x) == "") return(NULL)

    # Normalise whitespace and quotes
    x <- gsub("[\r\n\t]+", " ", x, perl = TRUE)
    x <- gsub("\\s+", " ", x, perl = TRUE)
    x <- gsub("[\u2018\u2019]", "'", x, perl = TRUE)
    x <- gsub("[\u201C\u201D]", "\"", x, perl = TRUE)
    x <- trimws(x)

    # Split first on semicolons, then extract every GO match within each chunk
    chunks <- trimws(unlist(strsplit(x, ";", fixed = TRUE)))
    chunks <- chunks[chunks != ""]

    if (length(chunks) == 0L) return(NULL)

    go_ids   <- character()
    go_terms <- character()

    for (chunk in chunks) {
      # Match each "description [GO:0000000]" unit
      matches <- regmatches(
        chunk,
        gregexpr("[^\\[]*\\[GO:[0-9]{7}\\]", chunk, perl = TRUE)
      )[[1]]

      if (length(matches) == 0L) next

      for (m in matches) {
        this_go <- sub(".*\\[(GO:[0-9]{7})\\].*", "\\1", m, perl = TRUE)
        this_term <- sub("\\s*\\[GO:[0-9]{7}\\].*", "", m, perl = TRUE)
        this_term <- clean_term(this_term)

        go_ids   <- c(go_ids, this_go)
        go_terms <- c(go_terms, if (nzchar(this_term)) this_term else NA_character_)
      }
    }

    if (length(go_ids) == 0L) return(NULL)

    data.frame(
      GO   = go_ids,
      TERM = go_terms,
      stringsAsFactors = FALSE
    )
  }

  # First pass: extract all GO-term pairs and count output rows
  parsed_list <- lapply(go_text, extract_go_pairs)
  counts <- vapply(parsed_list, function(x) {
    if (is.null(x)) 0L else nrow(x)
  }, integer(1))

  total_rows <- sum(counts)

  if (total_rows == 0L) {
    return(empty_out)
  }

  # Preallocate final vectors once
  GENE <- character(total_rows)
  ID   <- character(total_rows)
  GO   <- character(total_rows)
  TERM <- character(total_rows)

  pos <- 1L

  # Second pass: fill preallocated vectors
  for (i in seq_along(parsed_list)) {
    parsed <- parsed_list[[i]]
    if (is.null(parsed) || nrow(parsed) == 0L) next

    n_terms <- nrow(parsed)
    idx <- pos:(pos + n_terms - 1L)

    GENE[idx] <- gene_vec[i]
    ID[idx]   <- id_vec[i]
    GO[idx]   <- parsed$GO
    TERM[idx] <- parsed$TERM

    pos <- pos + n_terms
  }

  out <- data.frame(
    GENE = GENE,
    ID   = ID,
    GO   = GO,
    TERM = TERM,
    stringsAsFactors = FALSE
  )

  rownames(out) <- NULL

  if (deduplicate) {
    out <- unique(out)
    rownames(out) <- NULL
  }

  out
}

# -------------------------------
# Expand BP and MF properly
# -------------------------------
# remove files for space
rm(BP_MF_Terms,
   matching_file)
  
  # Still having memory issue, need to split file in half and run separately
# -------------------------------
# Split merged_df into thirds
# -------------------------------
n <- nrow(merged_df)

idx1 <- 1:floor(n / 3)
idx2 <- (floor(n / 3) + 1):floor(2 * n / 3)
idx3 <- (floor(2 * n / 3) + 1):n

part1 <- merged_df[idx1, , drop = FALSE]
part2 <- merged_df[idx2, , drop = FALSE]
part3 <- merged_df[idx3, , drop = FALSE]

rm(merged_df,
   idx1,
   idx2,
   idx3,
   n)

# -------------------------------
# Expand GO:BP in chunks
# -------------------------------
BP_long_1 <- expand_go_column_mem(part1, "GO:BP", deduplicate = FALSE)
BP_long_2 <- expand_go_column_mem(part2, "GO:BP", deduplicate = FALSE)
BP_long_3 <- expand_go_column_mem(part3, "GO:BP", deduplicate = FALSE)

BP_long <- unique(rbind(BP_long_1, BP_long_2, BP_long_3))
rm(BP_long_1, BP_long_2, BP_long_3)
BP_long <- unique(BP_long) # no duplicated detected

# -------------------------------
# Expand GO:MF in chunks
# -------------------------------
MF_long_1 <- expand_go_column_mem(part1, "GO:MF", deduplicate = FALSE)
MF_long_2 <- expand_go_column_mem(part2, "GO:MF", deduplicate = FALSE)
MF_long_3 <- expand_go_column_mem(part3, "GO:MF", deduplicate = FALSE)

MF_long <- unique(rbind(MF_long_1, MF_long_2, MF_long_3))
rm(MF_long_1, MF_long_2, MF_long_3)
MF_long <- unique(MF_long) # no duplicates detected

# Only make downstream objects after BP_long / MF_long are finished
GENES_BP <- unique(BP_long[, c("GO", "GENE")])
GENES_MF <- unique(MF_long[, c("GO", "GENE")])

TERMs_BP <- unique(BP_long[, c("GO", "TERM")])
TERMs_MF <- unique(MF_long[, c("GO", "TERM")])

# remove redundant dfs now
rm(part1,
   part2,
   part3,
   BP_long,
   MF_long)

save.image(file = "C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_2/data/temp_storage/BP_MF_long_extractions.Rdata")

# load when needed
  # load("C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/data/temp_storage/BP_MF_long_extractions.Rdata")
```

# Acute Experiment
## ORA
```{r stratified_ORA_exp1}
# Load DMR data
dmr_data <- read.table("C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_2/data/temp_storage/exp1_DMRs_gene_id_matched.txt", header = TRUE, sep = '\t', stringsAsFactors = FALSE)

# Set all region types to analyze
region_types <- c(
  "promoter",
  "three_prime_UTR",
  "exon",
  "gene",
  "five_prime_UTR",
  "downstream_region",
  "dispersed_repeat",
  "ncRNA_gene"
)

# Helper function to convert x/y strings to numeric
convertFractionToNumeric <- function(x) {
  sapply(as.character(x), function(val) {
    parts <- strsplit(val, "/")[[1]]
    if (length(parts) == 2) {
      as.numeric(parts[1]) / as.numeric(parts[2])
    } else {
      NA_real_
    }
  })
}

# Loop through each region
for (target_region in region_types) {
  
  message("Processing region: ", target_region)
  
  # Subset data to selected region
  region_data <- dmr_data[dmr_data$region == target_region, ]
  
  if (nrow(region_data) == 0) {
    message("  No rows found for region: ", target_region)
    next
  }
  
  # Filter for significant DMRs
  sig_data <- region_data[region_data$p_fdr <= 0.1, ]
  
  if (nrow(sig_data) == 0) {
    message("  No significant DMRs for region: ", target_region)
    next
  }
  
  # Keep the gene-context mapping from the filtered DMR data
  gene_context_map <- sig_data[
    !is.na(sig_data$gene_id_labelling) & sig_data$gene_id_labelling != "",
    c("gene_id_labelling", "context")
  ]
  
  # Extract gene IDs, removing NA or blank
  gene_ids <- as.character(sig_data$gene_id_labelling)
  gene_ids <- gene_ids[!is.na(gene_ids) & gene_ids != ""]
  
  if (length(gene_ids) == 0) {
    message("  No significant gene IDs found for region: ", target_region)
    next
  }
  
  # Run ORA safely
  ORA_BP <- tryCatch(
    enricher(
      gene = gene_ids,
      TERM2GENE = GENES_BP,
      TERM2NAME = TERMs_BP,
      minGSSize = 1,
      pvalueCutoff = 1,
      qvalueCutoff = 1
    ),
    error = function(e) NULL
  )
  
  ORA_MF <- tryCatch(
    enricher(
      gene = gene_ids,
      TERM2GENE = GENES_MF,
      TERM2NAME = TERMs_MF,
      minGSSize = 1,
      pvalueCutoff = 1,
      qvalueCutoff = 1
    ),
    error = function(e) NULL
  )
  
  result_list <- list()
  
  # Process BP results if present
  if (!is.null(ORA_BP) && nrow(ORA_BP@result) > 0) {
    bp_results <- ORA_BP@result
    
    bp_expanded <- bp_results %>%
      mutate(geneID = strsplit(as.character(geneID), "/")) %>%
      unnest(geneID) %>%
      left_join(
        gene_context_map,
        by = c("geneID" = "gene_id_labelling"),
        relationship = "many-to-many"
      ) %>%
      mutate(GO = "Biological Processes")
    
    result_list[["BP"]] <- bp_expanded
  }
  
  # Process MF results if present
  if (!is.null(ORA_MF) && nrow(ORA_MF@result) > 0) {
    mf_results <- ORA_MF@result
    
    mf_expanded <- mf_results %>%
      mutate(geneID = strsplit(as.character(geneID), "/")) %>%
      unnest(geneID) %>%
      left_join(
        gene_context_map,
        by = c("geneID" = "gene_id_labelling"),
        relationship = "many-to-many"
      ) %>%
      mutate(GO = "Molecular Functions")
    
    result_list[["MF"]] <- mf_expanded
  }
  
  # Skip if neither BP nor MF returned results
  if (length(result_list) == 0) {
    message("  No ORA results found for region: ", target_region)
    next
  }
  
  # Combine and annotate results
  all_results <- bind_rows(result_list) %>%
    mutate(
      region = target_region,
      experiment = "Acute Experiment",
      GeneRatio = convertFractionToNumeric(GeneRatio),
      BgRatio = convertFractionToNumeric(BgRatio)
    )
  
  # Save to environment and CSV
  assign(paste0("ORA_results_", target_region, "_exp1"), all_results)
  output_file <- paste0("ORA_results_", target_region, ".csv")
  write.csv(all_results, output_file, row.names = FALSE)
  
  message("  Analysis complete and results saved for region: ", target_region)
}

# Combined all results together
ORA_results_exp1_combined <- rbind(ORA_results_dispersed_repeat_exp1,
                                   ORA_results_exon_exp1,
                                   ORA_results_gene_exp1,
                                   ORA_results_three_prime_UTR_exp1,
                                   ORA_results_promoter_exp1)

# Plotting data
ORA_results_exp1_combined <- ORA_results_exp1_combined[order(ORA_results_exp1_combined$GeneRatio), ]

ORA_results_exp1_combined_clean <- unique(ORA_results_exp1_combined)


ORA_exp1_dotplot <- ggplot(data = ORA_results_exp1_combined_clean, 
                              aes(y = reorder(Description, 
                                              Count, 
                                              increasing = T), 
                                  x = GeneRatio, 
                                  color = region)) + 
                       geom_point(aes(size = Count), 
                                  stroke = 4) +
                       labs(title = "Biological Processes", y = "GO Description", x = "Gene Ratio") +
                       theme_light() + 
                       #scale_color_gradient(low = "red", high = "blue") + 
                       scale_size_continuous(breaks = seq(min(ORA_results_exp1_combined$Count), max(ORA_results_exp1_combined$Count), by = 1)) +
                       theme(
                         legend.spacing = unit(0.5, "cm"),         # Adjust overall legend spacing
                         legend.key.height = unit(3, "lines"), # Adjust the height between size examples
                         axis.text.y = element_text(size = 11)
                         ) + 
                       guides(size = guide_legend(title = "Count"))  + 
                       scale_y_discrete(labels = function(x) str_wrap(x, width = 28)) + # Adjust width for line breaks 
                       facet_wrap(~GO, scales = "free")

# Merge ORA_exp1_BP results and dotplot together, and rm dotplot
plot(ORA_exp1_dotplot)

## Export dataframe results
write.csv(ORA_results_exp1_combined_clean, "C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_2/Results/ORA_results_exp1_combined.csv", row.names = FALSE)

# Save current Environemnt:
save.image(file = "GSEA_exp1_prep.Rdata")
```


## GSEA

```{r looping_GSEA_by_region_exp1}
library(clusterProfiler)
library(dplyr)
library(ggplot2)
library(ggforce)
library(tidyr)
library(stringr)

# Read and prep data
exp1_DM_region_data <- read.table("C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_2/data/temp_storage/exp1_DMRs_gene_id_matched.txt", header = TRUE, sep = '\t')
exp1_DM_region_data$region <- as.factor(exp1_DM_region_data$region)

region_col <- "region"
gene_col <- 19

gene_col_name <- names(exp1_DM_region_data)[gene_col]

exp1_gene_duplicates_by_region <- exp1_DM_region_data %>%
  mutate(
    region_clean = trimws(as.character(.data[[region_col]])),
    gene_clean = trimws(as.character(.data[[gene_col_name]]))
  ) %>%
  filter(
    !is.na(region_clean), region_clean != "",
    !is.na(gene_clean), gene_clean != ""
  ) %>%
  count(region_clean, gene_clean, name = "n_instances") %>%
  filter(n_instances > 1) %>%
  arrange(region_clean, desc(n_instances))

exp1_gene_duplicates_by_region

exp1_duplicate_summary <- exp1_DM_region_data %>%
  mutate(
    region_clean = trimws(as.character(.data[[region_col]])),
    gene_clean = trimws(as.character(.data[[gene_col_name]]))
  ) %>%
  filter(
    !is.na(region_clean), region_clean != "",
    !is.na(gene_clean), gene_clean != ""
  ) %>%
  group_by(region_clean) %>%
  summarise(
    total_rows = n(),
    unique_genes = n_distinct(gene_clean),
    duplicate_gene_rows = total_rows - unique_genes,
    has_duplicates = duplicate_gene_rows > 0,
    .groups = "drop"
  ) %>%
  arrange(desc(duplicate_gene_rows))

exp1_duplicate_summary

library(dplyr)
library(tibble)
library(stringr)
library(purrr)

run_region_gsea <- function(
  dm_region_data,
  GENES_BP,
  TERMs_BP,
  GENES_MF,
  TERMs_MF,
  region_col = "region",
  score_col = 7,
  gene_col = 19,
  
  # Region-level filter:
  # minimum number of usable ranked genes required to run GSEA for a region
  min_genes = 10,
  
  # Significant-result filter after GSEA
  padj_cutoff = 0.1,
  
  # clusterProfiler::GSEA options
  exponent = 1,
  minGSSize = 10,
  maxGSSize = 500,
  pvalueCutoff = 1,
  pAdjustMethod = "BH",
  gson_BP = NULL,
  gson_MF = NULL,
  verbose = TRUE,
  
  # Reproducibility
  seed = 100,
  
  # Downstream annotation
  add_methdiff = TRUE) {
  
  empty_gsea_tbl <- tibble(
    ID = character(),
    Description = character(),
    setSize = integer(),
    enrichmentScore = numeric(),
    NES = numeric(),
    pvalue = numeric(),
    p.adjust = numeric(),
    qvalue = numeric(),
    rank = numeric(),
    leading_edge = character(),
    core_enrichment = character(),
    GO = character(),
    region = character(),
    methDiff_combined = character(),
    Gene_IDs_list = list(),
    MethDiff_list = list()
  )
  
  empty_audit_tbl <- tibble(
    region = character(),
    ontology = character(),
    n_region_rows = integer(),
    n_usable_genes = integer(),
    n_gsea_rows = integer(),
    n_sig_rows = integer(),
    status = character()
  )
  
  if (!region_col %in% names(dm_region_data)) {
    stop("region_col not found in dm_region_data")
  }
  
  if (!is.null(seed)) {
    set.seed(seed)
  }
  
  region_values <- dm_region_data[[region_col]] %>%
    as.character() %>%
    trimws()
  
  region_values <- sort(unique(region_values[!is.na(region_values) & region_values != ""]))
  
  if (length(region_values) == 0) {
    return(list(
      all_results = empty_gsea_tbl,
      sig_results = empty_gsea_tbl,
      audit = empty_audit_tbl
    ))
  }
  
  make_feature_TERM2GENE <- function(TERM2GENE, feature_lookup) {
  
  term_col <- names(TERM2GENE)[1]
  gene_col_term <- names(TERM2GENE)[2]
  
  TERM2GENE %>%
    transmute(
      term_id = .data[[term_col]],
      gene_id = trimws(as.character(.data[[gene_col_term]]))
    ) %>%
    inner_join(
      feature_lookup %>%
        select(gene_id, feature_id),
      by = "gene_id",
      relationship = "many-to-many"
    ) %>%
    transmute(
      term_id = term_id,
      feature_id = feature_id
    ) %>%
    distinct()
}
  
  run_one_gsea <- function(
    geneList,
    TERM2GENE,
    TERM2NAME,
    ontology_label,
    region_type,
    gson = NULL
  ) {
    
    gsea_args <- list(
      geneList = geneList,
      exponent = exponent,
      minGSSize = minGSSize,
      maxGSSize = maxGSSize,
      pvalueCutoff = pvalueCutoff,
      pAdjustMethod = pAdjustMethod,
      TERM2GENE = TERM2GENE,
      TERM2NAME = TERM2NAME,
      verbose = verbose
    )
    
    res <- tryCatch(
      {
        do.call(clusterProfiler::GSEA, gsea_args)
      },
      error = function(e) {
        message("  ", ontology_label, " failed for ", region_type, ": ", conditionMessage(e))
        NULL
      }
    )
    
    if (is.null(res) || is.null(res@result) || nrow(res@result) == 0) {
      return(NULL)
    }
    
    as_tibble(res@result) %>%
      mutate(
        GO = ontology_label,
        region = region_type
      )
  }
  
  all_results_list <- list()
  sig_results_list <- list()
  audit_list <- list()
  
  idx <- 1L
  
  for (region_type in region_values) {
    
    message("Processing region: ", region_type)
    
    region_data <- dm_region_data %>%
      filter(trimws(as.character(.data[[region_col]])) == region_type)
    
    n_region_rows <- nrow(region_data)
    
    if (n_region_rows == 0) {
      audit_list[[idx]] <- tibble(
        region = region_type,
        ontology = c("biological_processes", "molecular_function"),
        n_region_rows = 0L,
        n_usable_genes = 0L,
        n_gsea_rows = 0L,
        n_sig_rows = 0L,
        status = "no_region_rows"
      )
      idx <- idx + 1L
      next
    }
    
    gene_tbl <- tibble(
      row_index = seq_len(nrow(region_data)),
      gene_id = trimws(as.character(region_data[[gene_col]])),
      score = suppressWarnings(as.numeric(region_data[[score_col]]))
    ) %>%
      filter(!is.na(gene_id), gene_id != "", !is.na(score)) %>%
      group_by(gene_id) %>%
      mutate(
        duplicate_index = row_number(),
        feature_id = paste0(gene_id, "__feature", duplicate_index)
      ) %>%
      ungroup() %>%
      arrange(desc(score))
    
    n_usable_genes <- nrow(gene_tbl)
    
    feature_lookup <- gene_tbl %>%
      select(gene_id, feature_id) %>%
      distinct()
    
    GENES_BP_region <- make_feature_TERM2GENE(GENES_BP, feature_lookup)
    GENES_MF_region <- make_feature_TERM2GENE(GENES_MF, feature_lookup)
    
    if (n_usable_genes < min_genes) {
      
      message("  Skipping ", region_type, " — only ", n_usable_genes, " usable genes")
      
      audit_list[[idx]] <- tibble(
        region = region_type,
        ontology = c("biological_processes", "molecular_function"),
        n_region_rows = n_region_rows,
        n_usable_genes = n_usable_genes,
        n_gsea_rows = 0L,
        n_sig_rows = 0L,
        status = "too_few_genes"
      )
      
      idx <- idx + 1L
      next
    }
    
    geneList <- gene_tbl$score
    names(geneList) <- gene_tbl$feature_id
    geneList <- sort(geneList, decreasing = TRUE)
    
    bp_all <- run_one_gsea(
      geneList = geneList,
      TERM2GENE = GENES_BP_region,
      TERM2NAME = TERMs_BP,
      ontology_label = "biological_processes",
      region_type = region_type
    )
    
    mf_all <- run_one_gsea(
      geneList = geneList,
      TERM2GENE = GENES_MF_region,
      TERM2NAME = TERMs_MF,
      ontology_label = "molecular_function",
      region_type = region_type
    )
    
    region_all <- bind_rows(bp_all, mf_all)
    
    if (nrow(region_all) == 0) {
      
      message("  No GSEA results returned for ", region_type)
      
      audit_list[[idx]] <- tibble(
        region = region_type,
        ontology = c("biological_processes", "molecular_function"),
        n_region_rows = n_region_rows,
        n_usable_genes = n_usable_genes,
        n_gsea_rows = 0L,
        n_sig_rows = 0L,
        status = "no_gsea_results"
      )
      
      idx <- idx + 1L
      next
    }
    
    if (!all(region_all$region == region_type)) {
      stop("Region assignment failed for ", region_type)
    }
    
    region_sig <- region_all %>%
      filter(!is.na(p.adjust), p.adjust < padj_cutoff)
    
    if (
      nrow(region_sig) > 0 &&
      isTRUE(add_methdiff) &&
      exists("get_methDiff_data", mode = "function", inherits = TRUE) &&
      "core_enrichment" %in% names(region_sig)
    ) {
      
      region_sig <- region_sig %>%
        rowwise() %>%
        mutate(
          methDiff_combined = get_methDiff_data(core_enrichment, region_data)
        ) %>%
        ungroup() %>%
        mutate(
          Gene_IDs_list = strsplit(as.character(core_enrichment), "/"),
          MethDiff_list = strsplit(as.character(methDiff_combined), "/")
        )
      
    } else if (nrow(region_sig) > 0) {
      
      region_sig <- region_sig %>%
        mutate(
          methDiff_combined = NA_character_,
          Gene_IDs_list = strsplit(as.character(core_enrichment), "/"),
          MethDiff_list = vector("list", n())
        )
    }
    
    all_results_list[[region_type]] <- region_all
    
    if (nrow(region_sig) > 0) {
      sig_results_list[[region_type]] <- region_sig
      message("  Saved ", nrow(region_sig), " significant rows for ", region_type)
    } else {
      message("  No results pass p.adjust cutoff for ", region_type)
    }
    
    bp_n_all <- if (is.null(bp_all)) 0L else nrow(bp_all)
    mf_n_all <- if (is.null(mf_all)) 0L else nrow(mf_all)
    
    bp_n_sig <- region_sig %>%
      filter(GO == "biological_processes") %>%
      nrow()
    
    mf_n_sig <- region_sig %>%
      filter(GO == "molecular_function") %>%
      nrow()
    
    audit_list[[idx]] <- tibble(
      region = region_type,
      ontology = c("biological_processes", "molecular_function"),
      n_region_rows = n_region_rows,
      n_usable_genes = n_usable_genes,
      n_gsea_rows = c(bp_n_all, mf_n_all),
      n_sig_rows = c(bp_n_sig, mf_n_sig),
      status = c(
        ifelse(bp_n_all > 0, "gsea_run", "no_gsea_results"),
        ifelse(mf_n_all > 0, "gsea_run", "no_gsea_results")
      )
    )
    
    idx <- idx + 1L
  }
  
  all_results <- if (length(all_results_list) > 0) {
    bind_rows(all_results_list)
  } else {
    empty_gsea_tbl %>%
      select(-methDiff_combined, -Gene_IDs_list, -MethDiff_list)
  }
  
  sig_results <- if (length(sig_results_list) > 0) {
    bind_rows(sig_results_list)
  } else {
    empty_gsea_tbl
  }
  
  audit <- if (length(audit_list) > 0) {
    bind_rows(audit_list)
  } else {
    empty_audit_tbl
  }
  
  list(
    all_results = all_results,
    sig_results = sig_results,
    audit = audit
  )
}

gsea_exp1 <- run_region_gsea(
  dm_region_data = exp1_DM_region_data,
  GENES_BP = GENES_BP,
  TERMs_BP = TERMs_BP,
  GENES_MF = GENES_MF,
  TERMs_MF = TERMs_MF,
  min_genes = 5, # "number of genes in region type threshold"
  minGSSize = 5, # "number of testable genes in the set"
  maxGSSize = 1000,
  pvalueCutoff = 1,
  padj_cutoff = 0.1,
  seed = 100
)


GSEA_exp1_data <- gsea_exp1$sig_results
GSEA_exp1_all  <- gsea_exp1$all_results
GSEA_exp1_audit <- gsea_exp1$audit

# -----------------------------
# Quick audit
# -----------------------------
GSEA_exp1_data %>%
  count(region, GO, sort = TRUE)

# Final global validation
if (nrow(GSEA_exp1_data) > 0) {
  bad_region_rows <- GSEA_exp1_data %>%
    filter(is.na(region) | trimws(as.character(region)) == "")
  
  if (nrow(bad_region_rows) > 0) {
    warning("Some rows in GSEA_exp1_data have missing/blank region values.")
  }
}

# Continue to plotting
{
  expanded_exp1 <- GSEA_exp1_data %>%
    tidyr::unnest(c(Gene_IDs_list, MethDiff_list))
  expanded_exp1$MethDiff_list <- as.numeric(expanded_exp1$MethDiff_list)
  expanded_exp1 <- expanded_exp1 %>% distinct()
  expanded_exp1$WrappedDescription <- str_wrap(expanded_exp1$Description, width = 20)
}

# Plot
g1 <- ggplot(expanded_exp1, aes(
  x = reorder(WrappedDescription, MethDiff_list),
  y = MethDiff_list
)) +
  geom_boxplot(
    aes(group = interaction(WrappedDescription, ID)),
    fill = "lightgrey",
    #fill = "enrichmentScore",
   # side = "l",
    alpha = 1
  ) +
  geom_jitter(
    aes(colour = region),
   # side = "r",
    alpha = 0.7,
    size = 3,
  #  position = position_nudge(x = 0),
   width = 0.2
  ) +
  coord_flip() +
  theme_classic() +
  facet_wrap(~GO, scales = "free",
             labeller = as_labeller(
               c("biological_processes" = "Biological Processes",
                 "molecular_function" = "Molecular Functions")
             )
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.7) +
  labs(
    y = "Methylation Difference (%)",
    x = "Gene Ontology (GO)",
    #fill = "Enrichment Score",
    title ="GSEA for Acute Experiment"
  ) +
scale_color_manual(
  name = "Region",
  values = c("#0072B2", "#CC79A7", "red", "#a1d61a"),
  labels = c("Gene", "Promoter", "Repeat","Exon")
)

g1
```

# Primed Experiment
## ORA
```{r stratified_ORA_exp2}
# Load DMR data
dmr_data <- read.table("C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_2/data/temp_storage/exp2_DMRs_gene_id_matched.txt", header = TRUE, sep = '\t', stringsAsFactors = FALSE)

# Set all region types to analyze
region_types <- c(
  "promoter",
  "three_prime_UTR",
  "exon",
  "gene",
  "five_prime_UTR",
  "downstream_region",
  "dispersed_repeat",
  "ncRNA_gene"
)

# Helper function to convert x/y strings to numeric
convertFractionToNumeric <- function(x) {
  sapply(as.character(x), function(val) {
    parts <- strsplit(val, "/")[[1]]
    if (length(parts) == 2) {
      as.numeric(parts[1]) / as.numeric(parts[2])
    } else {
      NA_real_
    }
  })
}

# Loop through each region
for (target_region in region_types) {
  
  message("Processing region: ", target_region)
  
  # Subset data to selected region
  region_data <- dmr_data[dmr_data$region == target_region, ]
  
  if (nrow(region_data) == 0) {
    message("  No rows found for region: ", target_region)
    next
  }
  
  # Filter for significant DMRs
  sig_data <- region_data[region_data$p_fdr <= 0.1, ]
  
  if (nrow(sig_data) == 0) {
    message("  No significant DMRs for region: ", target_region)
    next
  }
  
  # Keep the gene-context mapping from the filtered DMR data
  gene_context_map <- sig_data[
    !is.na(sig_data$gene_id_labelling) & sig_data$gene_id_labelling != "",
    c("gene_id_labelling", "context")
  ]
  
  # Extract gene IDs, removing NA or blank
  gene_ids <- as.character(sig_data$gene_id_labelling)
  gene_ids <- gene_ids[!is.na(gene_ids) & gene_ids != ""]
  
  if (length(gene_ids) == 0) {
    message("  No significant gene IDs found for region: ", target_region)
    next
  }
  
  # Run ORA safely
  ORA_BP <- tryCatch(
    enricher(
      gene = gene_ids,
      TERM2GENE = GENES_BP,
      TERM2NAME = TERMs_BP,
      minGSSize = 1,
      pvalueCutoff = 1,
      qvalueCutoff = 1
    ),
    error = function(e) NULL
  )
  
  ORA_MF <- tryCatch(
    enricher(
      gene = gene_ids,
      TERM2GENE = GENES_MF,
      TERM2NAME = TERMs_MF,
      minGSSize = 1,
      pvalueCutoff = 1,
      qvalueCutoff = 1
    ),
    error = function(e) NULL
  )
  
  result_list <- list()
  
  # Process BP results if present
  if (!is.null(ORA_BP) && nrow(ORA_BP@result) > 0) {
    bp_results <- ORA_BP@result
    
    bp_expanded <- bp_results %>%
      mutate(geneID = strsplit(as.character(geneID), "/")) %>%
      unnest(geneID) %>%
      left_join(
        gene_context_map,
        by = c("geneID" = "gene_id_labelling"),
        relationship = "many-to-many"
      ) %>%
      mutate(GO = "Biological Processes")
    
    result_list[["BP"]] <- bp_expanded
  }
  
  # Process MF results if present
  if (!is.null(ORA_MF) && nrow(ORA_MF@result) > 0) {
    mf_results <- ORA_MF@result
    
    mf_expanded <- mf_results %>%
      mutate(geneID = strsplit(as.character(geneID), "/")) %>%
      unnest(geneID) %>%
      left_join(
        gene_context_map,
        by = c("geneID" = "gene_id_labelling"),
        relationship = "many-to-many"
      ) %>%
      mutate(GO = "Molecular Functions")
    
    result_list[["MF"]] <- mf_expanded
  }
  
  # Skip if neither BP nor MF returned results
  if (length(result_list) == 0) {
    message("  No ORA results found for region: ", target_region)
    next
  }
  
  # Combine and annotate results
  all_results <- bind_rows(result_list) %>%
    mutate(
      region = target_region,
      experiment = "Primed Experiment",
      GeneRatio = convertFractionToNumeric(GeneRatio),
      BgRatio = convertFractionToNumeric(BgRatio)
    )
  
  # Save to environment and CSV
  assign(paste0("ORA_results_", target_region, "_exp2"), all_results)
  output_file <- paste0("ORA_results_", target_region, ".csv")
  write.csv(all_results, output_file, row.names = FALSE)
  
  message("  Analysis complete and results saved for region: ", target_region)
}

# Combined all results together
ORA_results_exp2_combined <- rbind(ORA_results_dispersed_repeat_exp2,
                                   ORA_results_downstream_region_exp2,
                                   ORA_results_exon_exp2,
                                   ORA_results_gene_exp2,
                                   ORA_results_three_prime_UTR_exp2,
                                   ORA_results_promoter_exp2)

# Plotting data
ORA_results_exp2_combined <- ORA_results_exp2_combined[order(ORA_results_exp2_combined$GeneRatio), ]

ORA_results_exp2_combined_clean <- unique(ORA_results_exp2_combined)


ORA_exp2_dotplot <- ggplot(data = ORA_results_exp2_combined_clean, 
                              aes(y = reorder(Description, 
                                              Count, 
                                              increasing = T), 
                                  x = GeneRatio, 
                                  color = region)) + 
                       geom_point(aes(size = Count), 
                                  stroke = 4) +
                       labs(title = "Biological Processes", y = "GO Description", x = "Gene Ratio") +
                       theme_light() + 
                       #scale_color_gradient(low = "red", high = "blue") + 
                       scale_size_continuous(breaks = seq(min(ORA_results_exp2_combined$Count), max(ORA_results_exp2_combined$Count), by = 1)) +
                       theme(
                         legend.spacing = unit(0.5, "cm"),         # Adjust overall legend spacing
                         legend.key.height = unit(3, "lines"), # Adjust the height between size examples
                         axis.text.y = element_text(size = 11)
                         ) + 
                       guides(size = guide_legend(title = "Count"))  + 
                       scale_y_discrete(labels = function(x) str_wrap(x, width = 28)) + # Adjust width for line breaks 
                       facet_wrap(~GO, scales = "free")

# Merge ORA_exp2_BP results and dotplot together, and rm dotplot
plot(ORA_exp2_dotplot)

## Export dataframe results
write.csv(ORA_results_exp2_combined_clean, "C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_2/Results/ORA_results_exp2_combined.csv", row.names = FALSE)

# Save current Environemnt:
save.image(file = "GSEA_exp2_prep.Rdata")
```

## GSEA

```{r looping_GSEA_by_region_exp2}
library(clusterProfiler)
library(dplyr)
library(ggplot2)
library(ggforce)
library(tidyr)
library(stringr)

# Read and prep data
exp2_DM_region_data <- read.table("C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_2/data/temp_storage/exp2_DMRs_gene_id_matched.txt", header = TRUE, sep = '\t')
exp2_DM_region_data$region <- as.factor(exp2_DM_region_data$region)

# Make df to store data
GSEA_exp2_data <- data.frame()

# Continue to plotting
gsea_exp2 <- run_region_gsea(
  dm_region_data = exp2_DM_region_data,
  GENES_BP = GENES_BP,
  TERMs_BP = TERMs_BP,
  GENES_MF = GENES_MF,
  TERMs_MF = TERMs_MF,
  min_genes = 5, # "number of genes in region type threshold"
  minGSSize = 5, # "number of testable genes in the set"
  maxGSSize = 1000,
  pvalueCutoff = 1,
  padj_cutoff = 0.1,
  seed = 100
)

GSEA_exp2_data <- gsea_exp2$sig_results
GSEA_exp2_all  <- gsea_exp2$all_results
GSEA_exp2_audit <- gsea_exp2$audit

# -----------------------------
# Quick audit
# -----------------------------
GSEA_exp2_data %>%
  count(region, GO, sort = TRUE)

# Final global validation
if (nrow(GSEA_exp2_data) > 0) {
  bad_region_rows <- GSEA_exp2_data %>%
    filter(is.na(region) | trimws(as.character(region)) == "")
  
  if (nrow(bad_region_rows) > 0) {
    warning("Some rows in GSEA_exp2_data have missing/blank region values.")
  }
}

# Continue to plotting
{
  expanded_exp2 <- GSEA_exp2_data %>%
    tidyr::unnest(c(Gene_IDs_list, MethDiff_list))
  expanded_exp2$MethDiff_list <- as.numeric(expanded_exp2$MethDiff_list)
  expanded_exp2 <- expanded_exp2 %>% distinct()
  expanded_exp2$WrappedDescription <- str_wrap(expanded_exp2$Description, width = 20)
}

# Plot
g2 <- ggplot(expanded_exp2, aes(
  x = reorder(WrappedDescription, MethDiff_list),
  y = MethDiff_list
)) +
  geom_boxplot(
    aes(group = interaction(WrappedDescription, ID)),
    fill = "lightgrey",
    #fill = "enrichmentScore",
   # side = "l",
    alpha = 1
  ) +
  geom_jitter(
    aes(colour = region),
   # side = "r",
    alpha = 0.7,
    size = 3,
  #  position = position_nudge(x = 0),
   width = 0.2
  ) +
  coord_flip() +
  theme_classic() +
  facet_wrap(~GO, scales = "free",
             labeller = as_labeller(
               c("biological_processes" = "Biological Processes",
                 "molecular_function" = "Molecular Functions")
             )
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.7) +
  labs(
    y = "Methylation Difference (%)",
    x = "Gene Ontology (GO)",
    #fill = "Enrichment Score",
    title ="GSEA for Primed Experiment"
  ) +
scale_color_manual(
  name = "Region",
  values = c("#a1d61a","orange"),
  labels = c("Exon", "3` UTR")
)

g2
g1
```

# Semantic Clustering of GO terms

New analysis, combining all enrichment analysis results from methylation analysis together, and then finding the
semantically clustered terms which most represents the specific GO terms, allowing more clear overlap of 
biological processes and molecular functions. 

Pulling the code from Chapter 1's semantic clustering, and will need to make a lot of changes
to it so that it functions on the methylation data

## Make OrgDB

```{r make_a_equina_orgdb}
# save image, trying to make new orgDB
save.image(file = "C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_2/data/temp_storage/semantic_clustering_prep.Rdata")

load(file = "C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_2/data/temp_storage/semantic_clustering_prep.Rdata")

library(AnnotationForge)
library(AnnotationDbi)
library(GO.db)
library(GOSemSim)

# If you use GOSemSim in published research, please cite:
# - Guangchuang Yu. Gene Ontology Semantic Similarity Analysis Using GOSemSim. In: Kidder B. (eds) Stem Cell Transcriptional Networks. Methods in Molecular Biology, 2020, 2117:207-215. Humana, New York, NY.  doi:10.1007/978-1-0716-0301-7_11
# - Guangchuang Yu, Fei Li, Yide Qin, Xiaochen Bo, Yibo Wu, Shengqi Wang. GOSemSim: an R package formeasuring semantic similarity among GO terms and gene products Bioinformatics 2010, 26(7):976-978.  doi:10.1093/bioinformatics/btq064

# Need to make a dataframe with the following columns:
  # GID = the gene ID, e.g., ev,.utg4.1
  # Symbol = optional, dont need
  # GENENAME = Description of GO, e.g., DNA transcription factor ...

# GENES_BP and TERMS_BP contain the data I need

# Close any open connections that might keep the SQLite file locked
try(closeAllConnections(), silent = TRUE)

# Remove old SQLite file and old package build directory
unlink("C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_2/data/temp_storage/org.Aequina.eg.sqlite", force = TRUE)
unlink("C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_2/data/temp_storage/org.Aequina.eg.db", recursive = TRUE, force = TRUE)

# Optional: confirm they are gone
file.exists("C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_2/data/temp_storage/org.Aequina.eg.sqlite")
dir.exists("C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_2/data/temp_storage/org.Aequina.eg.db")

detach("package:org.Aequina.eg.db", unload = TRUE)

# Combine BP and MF gene-to-GO mappings
all_go_raw <- bind_rows(
  GENES_BP %>% dplyr::select(GO, GENE),
  GENES_MF %>% dplyr::select(GO, GENE)
) %>%
  filter(!is.na(GO), !is.na(GENE)) %>%
  distinct()

# Minimal gene table for OrgDb
gene_info <- all_go_raw %>%
  distinct(GENE) %>%
  transmute(
    GID = GENE,
    SYMBOL = GENE,
    GENENAME = GENE
  ) %>%
  as.data.frame(stringsAsFactors = FALSE)

# GO table in the exact format AnnotationForge expects
go_table <- all_go_raw %>%
  transmute(
    GID = GENE,
    GO = GO,
    EVIDENCE = "IEA"
  ) %>%
  distinct() %>%
  as.data.frame(stringsAsFactors = FALSE)

rownames(gene_info) <- NULL
rownames(go_table) <- NULL

# Sanity checks
stopifnot(identical(names(gene_info), c("GID", "SYMBOL", "GENENAME")))
stopifnot(identical(names(go_table), c("GID", "GO", "EVIDENCE")))
stopifnot(!anyDuplicated(gene_info$GID))
stopifnot(all(grepl("^GO:\\d{7}$", go_table$GO)))

# 5. Build package
# Make a brand new output folder
pkg_path <- makeOrgPackage(
  gene_info = gene_info,
  go = go_table,
  version = "0.2.0",
  maintainer = "r02hw22h Williams <r02hw22hgruff@outlook.com>",
  author = "r02hw22h Williams",
  outputDir = "C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_2/data",
  tax_id = "6106",
  genus = "Actinia",
  species = "equina",
  goTable = "go"
)

# Test if it works, please.............
install.packages(pkg_path, repos = NULL, type = "source")
library(org.Aequina.eg.db)
```

```{r enrichment_results_str}
# Gather and prepare nested dataframes to combine
# Acute experiment
## ORA
ORA_results_exp1_combined_clean
## GSEA
GSEA_exp1_data

# Primed Experiemnt
## ORA
ORA_results_exp2_combined_clean
## GSEA
GSEA_exp2_data

library(dplyr)
library(tibble)
library(stringr)
library(purrr)
library(cluster)

# --------------------------------------------------
# 1. Standardise ontology labels across all inputs
# --------------------------------------------------
normalize_ontology <- function(x) {
  x <- tolower(trimws(as.character(x)))

  case_when(
    str_detect(x, "biol|process|^bp$") ~ "BP",
    str_detect(x, "mole|function|^mf$") ~ "MF",
    TRUE ~ NA_character_
  )
}

# --------------------------------------------------
# 2. Prepare one results table in a common format
# --------------------------------------------------
prep_go_results <- function(df, experiment, method) {
  df %>%
    ungroup() %>%
    transmute(
      ID = str_extract(as.character(.data$ID), "GO:[0-9]{7}"),
      Description = as.character(.data$Description),
      ontology = normalize_ontology(.data$GO),
      p.adjust = suppressWarnings(as.numeric(.data$p.adjust)),
      experiment = experiment,
      method = method
    ) %>%
    filter(
      !is.na(ID), ID != "",
      !is.na(Description), Description != "",
      !is.na(ontology)
    )
}

# --------------------------------------------------
# 3. Combine all four enrichment result tables
# --------------------------------------------------
all_go_results <- bind_rows(
  prep_go_results(ORA_results_exp1_combined_clean, "Acute",  "ORA"),
  prep_go_results(GSEA_exp1_data,                  "Acute",  "GSEA"),
  prep_go_results(ORA_results_exp2_combined_clean, "Primed", "ORA"),
  prep_go_results(GSEA_exp2_data,                  "Primed", "GSEA")
)

# --------------------------------------------------
# 4. Collapse to one unique term universe
#    (one row per GO ID + ontology)
# --------------------------------------------------
most_common_value <- function(x) {
  x <- x[!is.na(x) & x != ""]
  if (length(x) == 0) return(NA_character_)
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

unique_terms <- all_go_results %>%
  group_by(ID, ontology) %>%
  summarise(
    Description = most_common_value(Description),
    n_hits = n(),
    best_p_adjust = if (all(is.na(p.adjust))) NA_real_ else min(p.adjust, na.rm = TRUE),
    sources = paste(sort(unique(paste(experiment, method, sep = "_"))), collapse = ";"),
    .groups = "drop"
  )

# --------------------------------------------------
# 5. Safe GO semantic similarity matrix builder
# --------------------------------------------------
make_sim_matrix <- function(term_ids, sem_data) {
  term_ids <- unique(na.omit(as.character(term_ids)))
  term_ids <- term_ids[grepl("^GO:[0-9]{7}$", term_ids)]

  if (length(term_ids) == 0) {
    return(matrix(numeric(0), nrow = 0, ncol = 0))
  }

  if (length(term_ids) == 1) {
    sim <- matrix(1, nrow = 1, ncol = 1,
                  dimnames = list(term_ids, term_ids))
    return(sim)
  }

  sim <- termSim(
    t1 = term_ids,
    t2 = term_ids,
    semdata = sem_data,
    method = "Wang"
  )

  sim <- as.matrix(sim)

  if (is.null(dim(sim)) || length(dim(sim)) != 2) {
    stop("termSim() did not return a 2D matrix.")
  }

  # force order to match term_ids if possible
  if (!is.null(rownames(sim)) && all(term_ids %in% rownames(sim))) {
    sim <- sim[term_ids, term_ids, drop = FALSE]
  }

  sim[is.na(sim)] <- 0
  diag(sim) <- 1

  sim
}

# --------------------------------------------------
# 6. Fit semantic clusters
# --------------------------------------------------
fit_semantic_clusters <- function(term_ids, sem_data, max_k = 100) {
  term_ids <- unique(na.omit(as.character(term_ids)))
  term_ids <- term_ids[grepl("^GO:[0-9]{7}$", term_ids)]
  n_terms <- length(term_ids)

  if (n_terms == 0) {
    return(list(
      sim = matrix(numeric(0), nrow = 0, ncol = 0),
      hc = NULL,
      k_eval = tibble(k = integer(), avg_silhouette = numeric()),
      best_k = NA_integer_,
      clusters = setNames(integer(), character())
    ))
  }

  if (n_terms == 1) {
    sim <- matrix(1, nrow = 1, ncol = 1,
                  dimnames = list(term_ids, term_ids))
    return(list(
      sim = sim,
      hc = NULL,
      k_eval = tibble(k = 1, avg_silhouette = NA_real_),
      best_k = 1L,
      clusters = setNames(1L, term_ids)
    ))
  }

  sim <- make_sim_matrix(term_ids, sem_data)

  if (!is.matrix(sim) || nrow(sim) != ncol(sim)) {
    stop("Semantic similarity result is not a square matrix.")
  }

  dist_mat <- as.dist(1 - sim)
  hc <- hclust(dist_mat, method = "average")

  if (n_terms == 2) {
    clusters <- setNames(c(1L, 1L), rownames(sim))
    return(list(
      sim = sim,
      hc = hc,
      k_eval = tibble(k = 1, avg_silhouette = NA_real_),
      best_k = 1L,
      clusters = clusters
    ))
  }

  k_grid <- 2:min(max_k, n_terms - 1)

  k_eval <- map_dfr(k_grid, function(k) {
    cl <- cutree(hc, k = k)
    sil <- silhouette(cl, dist_mat)
    sizes <- table(cl)

    tibble(
      k = k,
      avg_silhouette = mean(sil[, "sil_width"]),
      n_singletons = sum(sizes == 1),
      prop_singletons = mean(sizes == 1)
    )
  })

  best_k <- k_eval %>%
    filter(!is.na(avg_silhouette)) %>%
    slice_max(order_by = avg_silhouette, n = 1, with_ties = FALSE) %>%
    pull(k)

  if (length(best_k) == 0 || is.na(best_k)) {
    best_k <- 1L
    clusters <- setNames(rep(1L, n_terms), rownames(sim))
  } else {
    clusters <- cutree(hc, k = best_k)
  }

  list(
    sim = sim,
    hc = hc,
    k_eval = k_eval,
    best_k = best_k,
    clusters = clusters
  )
}


# --------------------------------------------------
# 7. Build lookup table:
#    each original GO term -> representative GO term
# --------------------------------------------------
build_semantic_lookup <- function(term_tbl, fit, ontology_label) {
  if (nrow(term_tbl) == 0 || length(fit$clusters) == 0) {
    return(tibble(
      ID = character(),
      Description = character(),
      ontology = character(),
      semantic_cluster_id = character(),
      semantic_cluster_GO = character(),
      semantic_cluster_Description = character(),
      cluster_size = integer(),
      mean_similarity_to_cluster = numeric(),
      n_hits = integer(),
      best_p_adjust = numeric(),
      sources = character()
    ))
  }

  cluster_members <- tibble(
    ID = names(fit$clusters),
    semantic_cluster_id = paste0(ontology_label, "_", unname(fit$clusters))
  ) %>%
    left_join(term_tbl, by = "ID")

  sim <- fit$sim

  member_scores <- cluster_members %>%
    group_by(semantic_cluster_id) %>%
    group_modify(function(.x, .y) {
      ids <- .x$ID
      sub_sim <- sim[ids, ids, drop = FALSE]

      .x %>%
        mutate(
          mean_similarity_to_cluster = rowMeans(sub_sim, na.rm = TRUE),
          cluster_size = n()
        )
    }) %>%
    ungroup()

  representatives <- member_scores %>%
    group_by(semantic_cluster_id) %>%
    arrange(
      desc(mean_similarity_to_cluster),
      best_p_adjust,
      desc(n_hits),
      Description,
      ID,
      .by_group = TRUE
    ) %>%
    slice(1) %>%
    ungroup() %>%
    transmute(
      semantic_cluster_id,
      semantic_cluster_GO = ID,
      semantic_cluster_Description = Description
    )

  member_scores %>%
    left_join(representatives, by = "semantic_cluster_id") %>%
    select(
      ID,
      Description,
      ontology,
      semantic_cluster_id,
      semantic_cluster_GO,
      semantic_cluster_Description,
      cluster_size,
      mean_similarity_to_cluster,
      n_hits,
      best_p_adjust,
      sources
    ) %>%
    arrange(ontology, semantic_cluster_id, desc(mean_similarity_to_cluster))
}

# --------------------------------------------------
# 8. Split by ontology and cluster separately
# --------------------------------------------------
bp_terms_tbl <- unique_terms %>% filter(ontology == "BP")
mf_terms_tbl <- unique_terms %>% filter(ontology == "MF")

bp_fit <- fit_semantic_clusters(bp_terms_tbl$ID, sem_bp, max_k = 100)
mf_fit <- fit_semantic_clusters(mf_terms_tbl$ID, sem_mf, max_k = 100)

best_k_bp <- bp_fit$best_k
best_k_bp
  # 31
best_k_mf <- mf_fit$best_k
best_k_mf
  # 39

# Optional: inspect k selection
print(bp_fit$k_eval, n = 50)
print(mf_fit$k_eval, n = 50)

par(mfrow = c(2,1))
plot(bp_fit$k_eval$k, bp_fit$k_eval$avg_silhouette, type = "b", main = "bp clusters",
     xlab = "k-value tested", ylab = "Average silhouette")
abline(v = best_k_bp, lty = "dashed", col = "grey60")
plot(mf_fit$k_eval$k, mf_fit$k_eval$avg_silhouette, type = "b", main = "mf clusters",
     xlab = "k-value tested", ylab = "Average silhouette")
abline(v =best_k_mf, lty = "dashed", col = "grey60")
par(mfrow = c(1,1))
  # WE want to maximise average silhouette since this means the clustering best fits the data
    # A rough guide:
    # 
    # > 0.5 =  strong clustering
    # 0.25 to 0.5 = moderate / usable structure # our clustering seems roughly at a usable level
    # < 0.25 = weak clustering structure
    # negative average = poor clustering overall

bp_lookup <- build_semantic_lookup(bp_terms_tbl, bp_fit, "BP")
mf_lookup <- build_semantic_lookup(mf_terms_tbl, mf_fit, "MF")

go_semantic_lookup <- bind_rows(bp_lookup, mf_lookup)

# --------------------------------------------------
# 9. Write the lookup file
# --------------------------------------------------
write.csv(go_semantic_lookup, "C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_2/data/temp_storage/GO_semantic_lookup_all_results.csv", row.names = FALSE)

# optional cluster-level summary table
go_cluster_summary <- go_semantic_lookup %>%
  distinct(
    ontology,
    semantic_cluster_id,
    semantic_cluster_GO,
    semantic_cluster_Description,
    cluster_size
  ) %>%
  arrange(ontology, semantic_cluster_id)

write.csv(go_cluster_summary, "C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_2/Results/GO_semantic_cluster_summary.csv", row.names = FALSE)


# Add semantic cluster labels to orginal dataframes:
add_semantic_labels <- function(df, lookup) {
  df %>%
    ungroup() %>%
    left_join(
      lookup %>%
        select(
          ID,
          semantic_cluster_id,
          semantic_cluster_GO,
          semantic_cluster_Description
        ),
      by = "ID"
    )
}

ORA_results_exp1_semantic  <- add_semantic_labels(ORA_results_exp1_combined_clean, go_semantic_lookup)
GSEA_exp1_semantic         <- add_semantic_labels(GSEA_exp1_data, go_semantic_lookup)
ORA_results_exp2_semantic  <- add_semantic_labels(ORA_results_exp2_combined_clean, go_semantic_lookup)
GSEA_exp2_semantic         <- add_semantic_labels(GSEA_exp2_data, go_semantic_lookup)

# --------------------------------------------------
# Add ontology labels consistently after semantic join
# --------------------------------------------------
ORA_results_exp1_semantic <- ORA_results_exp1_semantic %>%
  mutate(ontology = normalize_ontology(GO))

GSEA_exp1_semantic <- GSEA_exp1_semantic %>%
  mutate(ontology = normalize_ontology(GO))

ORA_results_exp2_semantic <- ORA_results_exp2_semantic %>%
  mutate(ontology = normalize_ontology(GO))

GSEA_exp2_semantic <- GSEA_exp2_semantic %>%
  mutate(ontology = normalize_ontology(GO))

# --------------------------------------------------
# Build semantic-cluster sets for Venn diagrams
# Use semantic_cluster_GO as the clustered term identity
# --------------------------------------------------
bp_sets <- list(
  "Acute ORA"   = ORA_results_exp1_semantic %>%
    filter(ontology == "BP", !is.na(semantic_cluster_GO), semantic_cluster_GO != "") %>%
    pull(semantic_cluster_GO) %>%
    unique(),
  
  "Acute GSEA"  = GSEA_exp1_semantic %>%
    filter(ontology == "BP", !is.na(semantic_cluster_GO), semantic_cluster_GO != "") %>%
    pull(semantic_cluster_GO) %>%
    unique(),
  
  "Primed ORA"  = ORA_results_exp2_semantic %>%
    filter(ontology == "BP", !is.na(semantic_cluster_GO), semantic_cluster_GO != "") %>%
    pull(semantic_cluster_GO) %>%
    unique(),
  
  "Primed GSEA" = GSEA_exp2_semantic %>%
    filter(ontology == "BP", !is.na(semantic_cluster_GO), semantic_cluster_GO != "") %>%
    pull(semantic_cluster_GO) %>%
    unique()
)

mf_sets <- list(
  "Acute ORA"   = ORA_results_exp1_semantic %>%
    filter(ontology == "MF", !is.na(semantic_cluster_GO), semantic_cluster_GO != "") %>%
    pull(semantic_cluster_GO) %>%
    unique(),
  
  "Acute GSEA"  = GSEA_exp1_semantic %>%
    filter(ontology == "MF", !is.na(semantic_cluster_GO), semantic_cluster_GO != "") %>%
    pull(semantic_cluster_GO) %>%
    unique(),
  
  "Primed ORA"  = ORA_results_exp2_semantic %>%
    filter(ontology == "MF", !is.na(semantic_cluster_GO), semantic_cluster_GO != "") %>%
    pull(semantic_cluster_GO) %>%
    unique(),
  
  "Primed GSEA" = GSEA_exp2_semantic %>%
    filter(ontology == "MF", !is.na(semantic_cluster_GO), semantic_cluster_GO != "") %>%
    pull(semantic_cluster_GO) %>%
    unique()
)

# Optional: remove empty sets so ggVennDiagram does not fail
bp_sets <- bp_sets[lengths(bp_sets) > 0]
mf_sets <- mf_sets[lengths(mf_sets) > 0]

# --------------------------------------------------
# Plot BP Venn diagram
# --------------------------------------------------
p_bp <- ggVennDiagram(
  bp_sets,
  category.names = names(bp_sets),
  set_color = "black",
  set_size = 3.5,
  label_alpha = 0,
  label_size = 6
) +
  scale_fill_gradient(low = "white", high = "#4c43e8") +
  theme_void(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.margin = margin(20, 30, 20, 30),
    legend.position = "none"
  ) +
  ggtitle("Biological Processes")

p_bp

# --------------------------------------------------
# Plot MF Venn diagram
# --------------------------------------------------
p_mf <- ggVennDiagram(
  mf_sets,
  category.names = names(mf_sets),
  set_color = "black",
  set_size = 3.5,
  label_alpha = 0,
  label_size = 6
) +
  scale_fill_gradient(low = "white", high = "#e8a943") +
  coord_equal(clip = "off") +
  theme_void(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.margin = margin(20, 25, 20, 25),
    legend.position = "none"
  ) +
  ggtitle("Molecular Functions")

p_mf

# Optional: save venn diagrams
ggsave("C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_2/Results/bp_venn_semantic_clusters_diagram.png", p_bp, width = 6, height = 6, dpi = 300)
ggsave("C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_2/Results/mf_venn_semantic_clusters_diagram.png", p_mf, width = 6, height = 6, dpi = 300)
```

## New plots with semantic clusters

```{r semantic_cluster_plots}
# Acute ORA
ORA_exp1_dotplot_semantic <- ggplot(data = ORA_results_exp1_semantic, 
                              aes(y = reorder(semantic_cluster_Description, 
                                              Count, 
                                              increasing = T), 
                                  x = GeneRatio, 
                                  color = region)) + 
                       geom_point(aes(size = Count), 
                                  stroke = 4) +
                       labs(title = "Biological Processes", y = "GO Description", x = "Gene Ratio") +
                       theme_light() + 
                       #scale_color_gradient(low = "red", high = "blue") + 
                       scale_size_continuous(breaks = seq(min(ORA_results_exp1_combined$Count), max(ORA_results_exp1_combined$Count), by = 1)) +
                       theme(
                         legend.spacing = unit(0.5, "cm"),         # Adjust overall legend spacing
                         legend.key.height = unit(3, "lines"), # Adjust the height between size examples
                         axis.text.y = element_text(size = 11)
                         ) + 
                       guides(size = guide_legend(title = "Count"))  + 
                       scale_y_discrete(labels = function(x) str_wrap(x, width = 28)) + # Adjust width for line breaks 
                       facet_wrap(~GO, scales = "free")

# Merge ORA_exp1_BP results and dotplot together, and rm dotplot
plot(ORA_exp1_dotplot_semantic)

# Primed ORA
ORA_exp2_dotplot_semantic <- ggplot(data = ORA_results_exp2_semantic, 
                              aes(y = reorder(semantic_cluster_Description, 
                                              Count, 
                                              increasing = T), 
                                  x = GeneRatio, 
                                  color = region)) + 
                       geom_point(aes(size = Count), 
                                  stroke = 4) +
                       labs(title = "Biological Processes", y = "GO Description", x = "Gene Ratio") +
                       theme_light() + 
                       #scale_color_gradient(low = "red", high = "blue") + 
                       scale_size_continuous(breaks = seq(min(ORA_results_exp2_combined$Count), max(ORA_results_exp2_combined$Count), by = 1)) +
                       theme(
                         legend.spacing = unit(0.5, "cm"),         # Adjust overall legend spacing
                         legend.key.height = unit(3, "lines"), # Adjust the height between size examples
                         axis.text.y = element_text(size = 11)
                         ) + 
                       guides(size = guide_legend(title = "Count"))  + 
                       scale_y_discrete(labels = function(x) str_wrap(x, width = 28)) + # Adjust width for line breaks 
                       facet_wrap(~GO, scales = "free")

# Merge ORA_exp2_BP results and dotplot together, and rm dotplot
plot(ORA_exp2_dotplot_semantic)


# ACute GSEA
# Continue to plotting
{
  expanded_semantic_exp1 <- GSEA_exp1_semantic %>%
    tidyr::unnest(c(Gene_IDs_list, MethDiff_list))
  expanded_semantic_exp1$MethDiff_list <- as.numeric(expanded_semantic_exp1$MethDiff_list)
  expanded_semantic_exp1 <- expanded_semantic_exp1 %>% distinct()
  expanded_semantic_exp1$WrappedDescription <- str_wrap(expanded_semantic_exp1$semantic_cluster_Description, width = 20)
}

# Plot
g1_semantic <- ggplot(expanded_semantic_exp1, aes(
  x = reorder(WrappedDescription, MethDiff_list),
  y = MethDiff_list
)) +
  geom_boxplot(
    aes(group = interaction(WrappedDescription, semantic_cluster_id)),
    fill = "lightgrey",
    #fill = "enrichmentScore",
   # side = "l",
    alpha = 1
  ) +
  geom_jitter(
    aes(colour = region),
   # side = "r",
    alpha = 0.7,
    size = 3,
  #  position = position_nudge(x = 0),
   width = 0.2
  ) +
  coord_flip() +
  theme_classic() +
  facet_wrap(~GO, scales = "free",
             labeller = as_labeller(
               c("biological_processes" = "Biological Processes",
                 "molecular_function" = "Molecular Functions")
             )
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.7) +
  labs(
    y = "Methylation Difference (%)",
    x = "Gene Ontology (GO)",
    #fill = "Enrichment Score",
    title ="GSEA for Acute Experiment - Semantically Clustered"
  ) #+
# scale_color_manual(
#   name = "Region",
#   values = c("red", "#a1d61a","#0072B2", "#CC79A7"),
#   labels = c("Repeat","Exon", "Gene", "Promoter")
# )

library(gridExtra)
grid.arrange(g1, g1_semantic, ncol = 2)

# Primed GSEA
# Continue to plotting
{
  expanded_semantic_exp2 <- GSEA_exp2_semantic %>%
    tidyr::unnest(c(Gene_IDs_list, MethDiff_list))
  expanded_semantic_exp2$MethDiff_list <- as.numeric(expanded_semantic_exp2$MethDiff_list)
  expanded_semantic_exp2 <- expanded_semantic_exp2 %>% distinct()
  expanded_semantic_exp2$WrappedDescription <- str_wrap(expanded_semantic_exp2$semantic_cluster_Description, width = 20)
}

# Plot
g2_semantic <- ggplot(expanded_semantic_exp2, aes(
  x = reorder(WrappedDescription, MethDiff_list),
  y = MethDiff_list
)) +
  geom_boxplot(
    aes(group = interaction(WrappedDescription, semantic_cluster_id)),
    fill = "lightgrey",
    #fill = "enrichmentScore",
   # side = "l",
    alpha = 1
  ) +
  geom_jitter(
    aes(colour = region),
   # side = "r",
    alpha = 0.7,
    size = 3,
  #  position = position_nudge(x = 0),
   width = 0.2
  ) +
  coord_flip() +
  theme_classic() +
  facet_wrap(~GO, scales = "free",
             labeller = as_labeller(
               c("biological_processes" = "Biological Processes",
                 "molecular_function" = "Molecular Functions")
             )
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 0.7) +
  labs(
    y = "Methylation Difference (%)",
    x = "Gene Ontology (GO)",
    #fill = "Enrichment Score",
    title ="GSEA for Acute Experiment - Semantically Clustered"
  ) +
scale_color_manual(
  name = "Region",
  values = c("red", "#a1d61a","#0072B2", "#CC79A7"),
  labels = c("Repeat","Exon", "Gene", "Promoter")
)

library(gridExtra)
grid.arrange(g2, g2_semantic, ncol = 2)
```

## Pie charts of region type distribution of enriched/overrepresented results

```{r pie_charts}
library(dplyr)
library(ggplot2)
library(scales)
library(patchwork)

# --------------------------------------------------
# Helper: summarise region distribution from a
# semantically clustered results dataframe
# --------------------------------------------------
summarise_region_pie <- function(df, dataset_name) {
  df %>%
    ungroup() %>%
    filter(!is.na(region), region != "") %>%
    mutate(
      cluster_term = case_when(
        !is.na(semantic_cluster_GO) & semantic_cluster_GO != "" ~ semantic_cluster_GO,
        TRUE ~ ID
      )
    ) %>%
    distinct(region, cluster_term) %>%
    count(region, name = "n") %>%
    mutate(
      percentage = 100 * n / sum(n),
      label = paste0(region, "\n", n, " (", round(percentage, 1), "%)"),
      dataset = dataset_name
    )
}

# --------------------------------------------------
# Helper: plot one pie chart
# --------------------------------------------------
plot_region_pie <- function(region_summary, title_text) {
  ggplot(region_summary, aes(x = "", y = n, fill = region)) +
    geom_col(width = 1, color = "white") +
    coord_polar(theta = "y") +
    geom_text(
      aes(label = label),
      position = position_stack(vjust = 0.5),
      size = 4
    ) +
    labs(
      title = title_text,
      fill = "Region type"
    ) +
    theme_void(base_size = 14) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      legend.position = "right"
    )
}

# --------------------------------------------------
# Build summaries for each semantically clustered dataframe
# ORA = overrepresented
# GSEA = enriched
# --------------------------------------------------
acute_ora_region_summary <- summarise_region_pie(
  ORA_results_exp1_semantic,
  "Acute ORA"
)

acute_gsea_region_summary <- summarise_region_pie(
  GSEA_exp1_semantic,
  "Acute GSEA"
)

primed_ora_region_summary <- summarise_region_pie(
  ORA_results_exp2_semantic,
  "Primed ORA"
)

primed_gsea_region_summary <- summarise_region_pie(
  GSEA_exp2_semantic,
  "Primed GSEA"
)

# --------------------------------------------------
# Plot individual pie charts
# --------------------------------------------------
p_acute_ora <- plot_region_pie(
  acute_ora_region_summary,
  "Acute ORA\nRegion distribution of semantic clusters"
)

p_acute_gsea <- plot_region_pie(
  acute_gsea_region_summary,
  "Acute GSEA\nRegion distribution of semantic clusters"
)

p_primed_ora <- plot_region_pie(
  primed_ora_region_summary,
  "Primed ORA\nRegion distribution of semantic clusters"
)

p_primed_gsea <- plot_region_pie(
  primed_gsea_region_summary,
  "Primed GSEA\nRegion distribution of semantic clusters"
)

p_acute_ora
p_acute_gsea
p_primed_ora
p_primed_gsea

# --------------------------------------------------
# Optional: combine all four into one figure
# --------------------------------------------------
combined_region_pies <- (p_acute_ora | p_acute_gsea) / (p_primed_ora | p_primed_gsea)
combined_region_pies

# --------------------------------------------------
# Optional: save the combined figure
# --------------------------------------------------
ggsave(
  "semantic_cluster_region_distribution_pies.png",
  combined_region_pies,
  width = 14,
  height = 12,
  dpi = 600
)

# --------------------------------------------------
# Optional: export the summary tables used for the plots
# --------------------------------------------------
region_distribution_summary <- bind_rows(
  acute_ora_region_summary,
  acute_gsea_region_summary,
  primed_ora_region_summary,
  primed_gsea_region_summary
)

write.csv(
  region_distribution_summary,
  "semantic_cluster_region_distribution_summary.csv",
  row.names = FALSE
)

```