---
title: "Enrichment_Analysis_C1"
author: "Hamish Williams"
date: "`r Sys.Date()`"
output: 
  html_document: 
    fig_height: 7
    toc: true
    toc_depth: 3
    toc_float: true
editor_options: 
  chunk_output_type: console
---

# Overview

This Markdown script will walk through all of the necessary steps undertaken to complete the **Gene Set Enrichment Analysis** and Over-representation **analysis** for the paired exposure experiment. This follows immediately from the **Differential Expression Analysis** results, and uses them to complete this analysis.

I was finding that even with the previous markdown sheet, that I generated a massive markdown sheet which was not easy to follow, which made it very difficult to understand the flow of the analysis, and at what stage of the process I was in. I have gone away and created a nice flow diagram which will neatly show the general flow of the analysis, and should make it much easier for myself and any other researchers who read this markdown to understand the process, and the steps undertaken to perform this analysis.

However, this is not a general guide to GSEA and Enrichment based analysis. Although I think that these types of content are very lacking in the scientific sphere, this script is not to be used as a single end all be all one step guide to completing this analysis. This markdown will wlakthrough the steps I have taken for my analysis, explaining and justifying parameter settings and additional actions required to complete the analysis. Experiments and analysis as a whole should be tailored to each experiment, and consideration of how best to tune the analyses to that experiment is important for statistical design, and for answering the specific questions of one's own research aims.

# Building Functions

## check_packages_ClusterProfilerR & Loading Packages

Function built to make loading and downloading of packages easier for reproducibility.

```{r check_packages_ClusterProfilerR, message=FALSE, warning=FALSE, results='hide'}
check_packages_ClusterProfileR <- function(pkg_list) {
  for (pkg in pkg_list) {
    if (!require(pkg, character.only = TRUE)) {
      print(paste(pkg, "is not installed, installing package..."))
      install.packages(pkg)
    } else {
      suppressMessages(library(pkg, character.only = TRUE))
      print(paste(pkg, "is installed and loaded"))
    }
  }
}

# Generate a dataframe with a list of packages required

pkg_list_ClusterProfileR<-c("tidyverse", # manipulations
                            "tibble", # tables
                            "stats", # stats
                            "ggplot2", # plottin
                            "dplyr", # manipulation
                            "tidyr", # manipulation
                            "ggfortify", # plotting
                            "knitr",# markdown
                            "clusterProfiler", # enrichment analysis
                            "enrichplot", # plotting
                            "data.table", # data formatting
                            "plotly", # plotting
                            "ggVennDiagram", # plotting
                            "VennDiagram", # plotting
                            "limma", # plotting
                            "ggridges", # plotting
                            "stringr") # manipulations

# Run Function::
check_packages_ClusterProfileR(pkg_list_ClusterProfileR)
```

## convertFractionToNumeric

Used to convert gene ratios in final analyses to numerical values for generating figures

```{r convertFractionToNumeric, message=FALSE, warning=FALSE}
convertFractionToNumeric <- function(fraction) {
  fraction_parts <- strsplit(fraction, "/")[[1]]
  numerator <- as.numeric(fraction_parts[1])
  denominator <- as.numeric(fraction_parts[2])
  numeric_value <- numerator / denominator
  return(numeric_value)
}
```

# Importing and Formatting Data

As outlined by the flow diagram, there are 3 main pieces of data needed to start the analysis. This include **Gene Ontology** data file (UNIPROT_results.txt), which contains the GO data for Biological Processes and Molecular Functions for the blastp matched genes of the UNIPROT database; a **Matching File**, which contains the names of the genes from *A. equina* vs. the Blastp result names - will be used to match the gene names to the blastp results; and finally the **Differential Expression Results**, which includes both the experiments completed (short and long)

```{r Importing_Data, results='hide'}
# Import BP and MF terms file ----
BP_MF_Terms <- read.table("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Data/UNIPROT_results.txt",
                          header = T, 
                          sep = '\t', 
                          stringsAsFactors = FALSE)

# Import the blastp (matching) file ----
matching_file <- read.table("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Data/genome_matching_file.txt", 
                            header = F, 
                            sep = '\t', 
                            stringsAsFactors = FALSE)

# Change the name of the columns to be identical ----
  names(matching_file)[names(matching_file) == "V13"] <- "ID"
  names(BP_MF_Terms)[names(BP_MF_Terms) == "From"] <- "ID"
  names(BP_MF_Terms)[names(BP_MF_Terms) == "Gene.Names"] <- "GENENAMES"
  names(BP_MF_Terms)[names(BP_MF_Terms) == "Gene.Ontology..biological.process."] <- "GO:BP"
  names(BP_MF_Terms)[names(BP_MF_Terms) == "Gene.Ontology..molecular.function."] <- "GO:MF"
  
# Merge blastp & GO terms ----
  # Merge DEG Files:
merged_df <- merge(matching_file, BP_MF_Terms, by = "ID")
# Subset Dataframe
merged_df <- subset(merged_df[, c("V1","ID","GO:BP","GO:MF")])
# Change Variable names
names(merged_df)[names(merged_df) == "V1"] <- "GENE"

# Separate BP and MF GOs ----
# Select BP column:
BP <- merged_df[,c("GENE","ID","GO:BP")]
BP <- BP[!apply(BP == "", 1, any), ]
BP <- BP[!duplicated(BP), ]
BP <- na.omit(BP)
# Extract GO terms:
GENES_BP <- BP
GENES_BP$GO <- ""
bracketed_rows <- grepl("\\[.*\\]", GENES_BP$`GO:BP`)
GENES_BP$GO[bracketed_rows] <- gsub(".*\\[(.*)\\].*", "\\1", GENES_BP$`GO:BP`[bracketed_rows])
GENES_BP <- GENES_BP[, c("GO","GENE")]
GENES_BP <- GENES_BP[!duplicated(GENES_BP), ]
GENES_BP <- GENES_BP[complete.cases(GENES_BP),]

## Extract Terms:
TERMs_BP <- BP
TERMs_BP$GO <- ""
bracketed_rows <- grepl("\\[.*\\]", TERMs_BP$`GO:BP`)
TERMs_BP$GO[bracketed_rows] <- gsub(".*\\[(.*)\\].*", "\\1", TERMs_BP$`GO:BP`[bracketed_rows])
TERMs_BP$`GO:BP`[bracketed_rows] <- gsub("\\[.*\\]", "", TERMs_BP$`GO:BP`[bracketed_rows])
TERMs_BP <- TERMs_BP[, c("GO","GO:BP")]
TERMs_BP <- TERMs_BP[!duplicated(TERMs_BP), ]
names(TERMs_BP)[names(TERMs_BP) == "GO:BP"] <- "GENENAMES_BP"
TERMs_BP <- TERMs_BP[complete.cases(TERMs_BP),]

# Select MF column:
MF <- merged_df[,c("GENE","ID","GO:MF")]
MF <- MF[!apply(MF == "", 1, any), ]
MF <- MF[!duplicated(MF), ]
MF <- na.omit(MF)
# Extract GO terms:
GENES_MF <- MF
GENES_MF$GO <- ""
bracketed_rows <- grepl("\\[.*\\]", GENES_MF$`GO:MF`)
GENES_MF$GO[bracketed_rows] <- gsub(".*\\[(.*)\\].*", "\\1", GENES_MF$`GO:MF`[bracketed_rows])
GENES_MF <- GENES_MF[, c("GO","GENE")]
GENES_MF <- GENES_MF[!duplicated(GENES_MF), ]
GENES_MF <- GENES_MF[complete.cases(GENES_MF),]

## Extract Terms:
TERMs_MF <- MF
TERMs_MF$GO <- ""
bracketed_rows <- grepl("\\[.*\\]", TERMs_MF$`GO:MF`)
TERMs_MF$GO[bracketed_rows] <- gsub(".*\\[(.*)\\].*", "\\1", TERMs_MF$`GO:MF`[bracketed_rows])
TERMs_MF$`GO:MF`[bracketed_rows] <- gsub("\\[.*\\]", "", TERMs_MF$`GO:MF`[bracketed_rows])
TERMs_MF <- TERMs_MF[, c("GO","GO:MF")]
TERMs_MF <- TERMs_MF[!duplicated(TERMs_MF), ]
names(TERMs_MF)[names(TERMs_MF) == "GO:MF"] <- "GENENAMES_MF"
TERMs_MF <- TERMs_MF[complete.cases(TERMs_MF),]

# Import DEA results ----
DE_results_1 <- read.table("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/DEA_results_exp1_no_outiler_and_shrunk_values.csv",
                         header = T, 
                         sep = ',', 
                         stringsAsFactors = TRUE)

DE_results_2 <- read.table("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/DEA_results_exp2_shrunk_values.csv",
                         header = T, 
                         sep = ',', 
                         stringsAsFactors = TRUE)

## Perform naming transformation to make gene IDs matching ----
DE_results_1$X <- gsub("\\.TU", "", DE_results_1$X) # Removing "TU" part of names
DE_results_1$X <- gsub("path", "mrna", DE_results_1$X) # Replacing "path" -> "mrna"
DE_results_2$X <- gsub("\\.TU", "", DE_results_2$X) # Removing "TU" part of names
DE_results_2$X <- gsub("path", "mrna", DE_results_2$X) # Replacing "path" -> "mrna"

# DEGs lists for ORA ----
DEGs <- DE_results_1[(DE_results_1$padj < 0.1),]
DEGs <- DEGs[complete.cases(DEGs),]
DEGs_1 <- as.vector(DEGs[,1])
length(DEGs_1)
DEGs <- DE_results_2[(DE_results_2$padj < 0.1),]
DEGs <- DEGs[complete.cases(DEGs),]
DEGs_2 <- as.vector(DEGs[,1])
length(DEGs_2)

# Create geneList objects for GSEA analyses ----
# Feature 1: numeric vector (LFC)
geneList_1 = DE_results_1[,3]
geneList_2 = DE_results_2[,3]
# Feature 2: Named Vector (rownames)
names(geneList_1) = as.character(DE_results_1[,1])
names(geneList_2) = as.character(DE_results_2[,1])
# feature 3: decreasing order;
geneList_1 = sort(geneList_1, decreasing = TRUE)
geneList_2 = sort(geneList_2, decreasing = TRUE)
```

## Tidy up Environment

```{r Tidy_up, results='hide'}
# Tidy up environment..
    rm(MF,
       BP,
       BP_MF_Terms,
       matching_file,
       DEGs,
       merged_df,
       bracketed_rows
       )

# Save Environment
save.image(file = "ORA_exp1_prep.RData")
```

# Overrepresentation Analysis

## Experiment 1

```{r ORA_exp1, results='hide'}
# Load environment for analysis
load("ORA_exp1_prep.RData")

# Run ORA
set.seed(100)
ORA_exp1_BP <- enricher(DEGs_1, 
                TERM2GENE = GENES_BP,
                TERM2NAME = TERMs_BP,
                minGSSize=1)

set.seed(100)
ORA_exp1_MF <- enricher(DEGs_1, 
                TERM2GENE = GENES_MF,
                TERM2NAME = TERMs_MF,
                minGSSize=1)

# Extract results and add in GO identifiers
ORA_exp1_BP_results <- ORA_exp1_BP@result
ORA_exp1_BP_results$GO <- "biological_processes"

ORA_exp1_MF_results <- ORA_exp1_MF@result
ORA_exp1_MF_results$GO <- "molecular_functions"

# combine results and subset for significant values
ORA_exp1_results <- rbind(ORA_exp1_BP_results, ORA_exp1_MF_results)
rm(ORA_exp1_BP_results, ORA_exp1_MF_results, ORA_exp1_BP, ORA_exp1_MF)
ORA_exp1_results <- subset(ORA_exp1_results[ORA_exp1_results$p.adjust < 0.1,])

# Change fractions to numeric values
ORA_exp1_results$GeneRatio <- sapply(ORA_exp1_results$GeneRatio, convertFractionToNumeric)
ORA_exp1_results$BgRatio <- sapply(ORA_exp1_results$BgRatio, convertFractionToNumeric)

## Export Dataframe results
write.csv(ORA_exp1_results, "C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/ORA_exp1_results.csv", row.names = FALSE)

## ... Replace GO descriptions with matched descriptions of GO:ID from QuickGO
  # import results with QuickGO descriptions

ORA_exp1_results <- read.table("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/ORA_exp1_results_QuickGO_names.csv",
                         header = T, 
                         sep = ',', 
                         stringsAsFactors = TRUE)

# Map the LFC values to ORA results using the gene names ----
# Make a lookup table: gene -> log2FoldChange
lfc_map <- setNames(DE_results_1$log2FoldChange, DE_results_1$X)

# Version 1: keep ORA rows intact, add a nested LFC column matching geneID order
ORA_exp1_combined_nested <- ORA_exp1_results %>%
  mutate(
    geneID = as.character(geneID),
    log2FoldChange = sapply(strsplit(geneID, "/"), function(genes) {
      paste(lfc_map[genes], collapse = "/")
    })
  )
# Version 2: long format, one gene per row with its matching LFC
ORA_exp1_combined_long <- ORA_exp1_results %>%
  mutate(geneID = as.character(geneID)) %>%
  separate_rows(geneID, sep = "/") %>%
  mutate(log2FoldChange = lfc_map[geneID])
```


## Experiment 2
```{r ORA_exp2, results='hide'}
# Load environment for analysis
#load("ORA_exp1_prep.RData")

# Run ORA
set.seed(100)
ORA_exp2_BP <- enricher(DEGs_2, 
                TERM2GENE = GENES_BP,
                TERM2NAME = TERMs_BP,
                minGSSize=1)

set.seed(100)
ORA_exp2_MF <- enricher(DEGs_2, 
                TERM2GENE = GENES_MF,
                TERM2NAME = TERMs_MF,
                minGSSize=1)

# Extract results and add in GO identifiers
ORA_exp2_BP_results <- ORA_exp2_BP@result
ORA_exp2_BP_results$GO <- "biological_processes"

ORA_exp2_MF_results <- ORA_exp2_MF@result
ORA_exp2_MF_results$GO <- "molecular_functions"

# combine results and subset for significant values
ORA_exp2_results <- rbind(ORA_exp2_BP_results, ORA_exp2_MF_results)
rm(ORA_exp2_BP_results, ORA_exp2_MF_results, ORA_exp2_BP, ORA_exp2_MF)
ORA_exp2_results <- subset(ORA_exp2_results[ORA_exp2_results$p.adjust < 0.1,])

# Change fractions to numeric values
ORA_exp2_results$GeneRatio <- sapply(ORA_exp2_results$GeneRatio, convertFractionToNumeric)
ORA_exp2_results$BgRatio <- sapply(ORA_exp2_results$BgRatio, convertFractionToNumeric)

## Export Dataframe results
write.csv(ORA_exp2_results, "C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/ORA_exp2_results.csv", row.names = FALSE)

## ... Replace GO descriptions with matched descriptions of GO:ID from QuickGO
  # import results with QuickGO descriptions

ORA_exp2_results <- read.table("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/ORA_exp2_results_QuickGO_names.csv",
                         header = T, 
                         sep = ',', 
                         stringsAsFactors = TRUE)

# Map the LFC values to ORA results using the gene names ----
# Make a lookup table: gene -> log2FoldChange
rm(lfc_map)
lfc_map <- setNames(DE_results_2$log2FoldChange, DE_results_2$X)

# Version 1: keep ORA rows intact, add a nested LFC column matching geneID order
ORA_exp2_combined_nested <- ORA_exp2_results %>%
  mutate(
    geneID = as.character(geneID),
    log2FoldChange = sapply(strsplit(geneID, "/"), function(genes) {
      paste(lfc_map[genes], collapse = "/")
    })
  )
# Version 2: long format, one gene per row with its matching LFC
ORA_exp2_combined_long <- ORA_exp2_results %>%
  mutate(geneID = as.character(geneID)) %>%
  separate_rows(geneID, sep = "/") %>%
  mutate(log2FoldChange = lfc_map[geneID])
```

# Gene Set Enrichment Analysis

## Experiment 1
```{r GSEA_exp1}
# Load Environment
# load("GSEA_exp1.RData")

# Run GSEA
set.seed(100)
GSEA_exp1_BP<- GSEA(geneList_1,
           TERM2GENE = GENES_BP,
           TERM2NAME = TERMs_BP,
           pvalueCutoff = 0.1)
set.seed(100)
GSEA_exp1_MF<- GSEA(geneList_1,
           TERM2GENE = GENES_MF,
           TERM2NAME = TERMs_MF,
           pvalueCutoff = 0.1)

# Get Results
GSEA_exp1_BP_results <- GSEA_exp1_BP@result
GSEA_exp1_BP_results$GO <- "biological_processes"

GSEA_exp1_MF_results <- GSEA_exp1_MF@result
GSEA_exp1_MF_results$GO <- "molecular_function"

# Merge results
GSEA_exp1_results <- rbind(GSEA_exp1_BP_results, GSEA_exp1_MF_results)

# Subset for only significant results
GSEA_exp1_results <- subset(GSEA_exp1_results[GSEA_exp1_results$p.adjust < 0.1,])

# Combine all results into single listed df
GSEA_exp1 <- list("BP" = GSEA_exp1_BP,
                  "MF" = GSEA_exp1_MF,
                  "Results" = GSEA_exp1_results)

# Clean-up 
rm(GSEA_exp1_MF_results, 
   GSEA_exp1_BP_results,
   GSEA_exp1_results,
   GSEA_exp1_BP,
   GSEA_exp1_MF)

## Export Dataframe results
write.csv(GSEA_exp1[["Results"]], "C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/GSEA_exp1_results.csv", row.names = FALSE)

# Capture QuickGO names... re-import
  # no results...
```

## Experiment 2
```{r GSEA_exp2}
# Load Environment
# load("GSEA_exp2.RData")

# Run GSEA
set.seed(100)
GSEA_exp2_BP<- GSEA(geneList_2,
           TERM2GENE = GENES_BP,
           TERM2NAME = TERMs_BP,
           pvalueCutoff = 0.1)
set.seed(100)
GSEA_exp2_MF<- GSEA(geneList_2,
           TERM2GENE = GENES_MF,
           TERM2NAME = TERMs_MF,
           pvalueCutoff = 0.1)

# Get Results
GSEA_exp2_BP_results <- GSEA_exp2_BP@result
GSEA_exp2_BP_results$GO <- "biological_processes"

GSEA_exp2_MF_results <- GSEA_exp2_MF@result
GSEA_exp2_MF_results$GO <- "molecular_function"

# Merge results
GSEA_exp2_results <- rbind(GSEA_exp2_BP_results, GSEA_exp2_MF_results)

# Subset for only significant results
GSEA_exp2_results <- subset(GSEA_exp2_results[GSEA_exp2_results$p.adjust < 0.1,])

# Combine all results into single listed df
GSEA_exp2 <- list("BP" = GSEA_exp2_BP,
                  "MF" = GSEA_exp2_MF,
                  "Results" = GSEA_exp2_results)

# Clean-up 
rm(GSEA_exp2_MF_results, 
   GSEA_exp2_BP_results,
   GSEA_exp2_results,
   GSEA_exp2_BP,
   GSEA_exp2_MF)

## Export Dataframe results
write.csv(GSEA_exp2[["Results"]], "C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/GSEA_exp2_results.csv", row.names = FALSE)

# Capture QuickGO names... re-import
GSEA_exp2_results <- read.table("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/GSEA_exp2_results_QuickGO_names.csv",
                         header = T, 
                         sep = ',', 
                         stringsAsFactors = TRUE)

# nested and long versions
# Make a lookup table: gene -> log2FoldChange
lfc_map <- setNames(DE_results_2$log2FoldChange, DE_results_2$X)

# Version 1: keep ORA rows intact, add a nested LFC column matching geneID order
GSEA_exp2_results_nested <- GSEA_exp2_results %>%
  mutate(
    core_enrichment = as.character(core_enrichment),
    log2FoldChange = sapply(strsplit(core_enrichment, "/"), function(genes) {
      paste(lfc_map[genes], collapse = "/")
    })
  )
# Version 2: long format, one gene per row with its matching LFC
GSEA_exp2_results_long <- GSEA_exp2_results %>%
  mutate(core_enrichment = as.character(core_enrichment)) %>%
  separate_rows(core_enrichment, sep = "/") %>%
  mutate(log2FoldChange = lfc_map[core_enrichment])
```

# Plotting Figures

```{r enrichment_plots}
p_boxes <- ORA_exp1_combined_long %>%
  mutate(
    Description_QuickGO = fct_reorder(
      Description_QuickGO,
      log2FoldChange,
      .fun = median,
      .na_rm = TRUE
    )
  ) %>%
  ggplot(aes(
    x = log2FoldChange,
    y = Description_QuickGO
  )) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    linewidth = 0.5,
    colour = "grey60"
  ) +
  geom_boxplot(
    width = 0.7,
    outlier.size = 1.5,
    outlier.alpha = 0.8,
    linewidth = 0.40,
    colour = "grey20",
    fill = "grey90"
  ) +
  labs(
    x = "log2 fold-change (LFC) - Ashr Shrunk",
    y = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.y = element_text(size = 9, colour = "black"),
    axis.text.x = element_text(colour = "black"),
    axis.line = element_line(linewidth = 0.5),
    plot.margin = margin(5.5, 8, 5.5, 5.5)
  ) + 
  facet_wrap(~GO, scales = "free",
             labeller = labeller(
               GO = c(
                 biological_processes = "Biolocial Processes",
                 molecular_functions = "Molecular Functions"
               )
             )
            )
p_boxes


## GSEA plot ----
p_boxes <- GSEA_exp2_results_long %>%
  mutate(
    Description = fct_reorder(
      Description,
      log2FoldChange,
      .fun = median,
      .na_rm = TRUE
    )
  ) %>%
  ggplot(aes(
    x = log2FoldChange,
    y = Description
  )) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    linewidth = 0.5,
    colour = "grey60"
  ) +
  geom_boxplot(
    width = 0.7,
    outlier.size = 1.5,
    outlier.alpha = 0.8,
    linewidth = 0.40,
    colour = "grey20",
    fill = "grey90"
  ) +
  labs(
    x = "log2 fold-change (LFC) - Ashr Shrunk",
    y = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.y = element_text(size = 9, colour = "black"),
    axis.text.x = element_text(colour = "black"),
    axis.line = element_line(linewidth = 0.5),
    plot.margin = margin(5.5, 8, 5.5, 5.5)
  ) + 
  facet_wrap(~GO, scales = "free",
             labeller = labeller(
               GO = c(
                 biological_processes = "Biolocial Processes",
                 molecular_functions = "Molecular Functions"
               )
             )
            )
p_boxes

```

# Semantic Clustering

Only really applicable with the ORA results since there are 82 in total, which is far too many to represent in a single plot, so semantic clustering allows us to group related terms together to better represent them in figures. But for supplementary materials it is still best to include all of the individual results since they should still be investigated and interpreted individually for best possible qualitative assessment of functional terms in relation to the experiments completed

## Make OrgDB

```{r make_a_equina_orgdb}
library(AnnotationForge)
library(AnnotationDbi)
library(GO.db)
library(GOSemSim)

# If you use GOSemSim in published research, please cite:
# - Guangchuang Yu. Gene Ontology Semantic Similarity Analysis Using GOSemSim. In: Kidder B. (eds) Stem Cell Transcriptional Networks. Methods in Molecular Biology, 2020, 2117:207-215. Humana, New York, NY.  doi:10.1007/978-1-0716-0301-7_11
# - Guangchuang Yu, Fei Li, Yide Qin, Xiaochen Bo, Yibo Wu, Shengqi Wang. GOSemSim: an R package formeasuring semantic similarity among GO terms and gene products Bioinformatics 2010, 26(7):976-978.  doi:10.1093/bioinformatics/btq064

# Need to make a dataframe with the following columns:
  # GID = the gene ID, e.g., ev,.utg4.1
  # Symbol = optional, dont need
  # GENENAME = Description of GO, e.g., DNA transcription factor ...

# GENES_BP and TERMS_BP contain the data I need

# Close any open connections that might keep the SQLite file locked
try(closeAllConnections(), silent = TRUE)

# Remove old SQLite file and old package build directory
unlink("./org.Aequina.eg.sqlite", force = TRUE)
unlink("./org.Aequina.eg.db", recursive = TRUE, force = TRUE)

# Optional: confirm they are gone
file.exists("./org.Aequina.eg.sqlite")
dir.exists("./org.Aequina.eg.db")

# Combine BP and MF gene-to-GO mappings
all_go_raw <- bind_rows(
  GENES_BP %>% dplyr::select(GO, GENE),
  GENES_MF %>% dplyr::select(GO, GENE)
) %>%
  filter(!is.na(GO), !is.na(GENE)) %>%
  distinct()

# Minimal gene table for OrgDb
gene_info <- all_go_raw %>%
  distinct(GENE) %>%
  transmute(
    GID = GENE,
    SYMBOL = GENE,
    GENENAME = GENE
  ) %>%
  as.data.frame(stringsAsFactors = FALSE)

# GO table in the exact format AnnotationForge expects
go_table <- all_go_raw %>%
  transmute(
    GID = GENE,
    GO = GO,
    EVIDENCE = "IEA"
  ) %>%
  distinct() %>%
  as.data.frame(stringsAsFactors = FALSE)

rownames(gene_info) <- NULL
rownames(go_table) <- NULL

# Sanity checks
stopifnot(identical(names(gene_info), c("GID", "SYMBOL", "GENENAME")))
stopifnot(identical(names(go_table), c("GID", "GO", "EVIDENCE")))
stopifnot(!anyDuplicated(gene_info$GID))
stopifnot(all(grepl("^GO:\\d{7}$", go_table$GO)))

# 5. Build package
# Make a brand new output folder
pkg_path <- makeOrgPackage(
  gene_info = gene_info,
  go = go_table,
  version = "0.1.0",
  maintainer = "Hamish Williams <hamishgruff@outlook.com>",
  author = "Hamish Williams",
  outputDir = ".",
  tax_id = "6106",
  genus = "Actinia",
  species = "equina",
  goTable = "go"
)

# Test if it works, please.............
install.packages(pkg_path, repos = NULL, type = "source")
library(org.Aequina.eg.db)
```

## Cluster

```{r semantic_clustering}
library(dplyr)
library(purrr)
library(cluster)
library(GOSemSim)
library(org.Aequina.eg.db) # orgdb w just made

# Build semantic data for each ontology
sem_bp <- godata(org.Aequina.eg.db, keytype = "GID", ont = "BP", computeIC = FALSE)
sem_mf <- godata(org.Aequina.eg.db, keytype = "GID", ont = "MF", computeIC = FALSE)

# Get unique GO IDs for one ontology from ORA results
get_terms <- function(df, go_class) {
  df %>%
    mutate(
      ID = as.character(ID),
      GO = as.character(GO)
    ) %>%
    filter(GO == go_class) %>%
    pull(ID) %>%
    unique()
}

# Build a clean similarity matrix
make_sim_matrix <- function(terms, sem_data) {
  sim <- termSim(terms, terms, semData = sem_data, method = "Wang")
  sim[is.na(sim)] <- 0
  diag(sim) <- 1
  sim
}

# Test candidate k values and score them with silhouette width
evaluate_k <- function(sim_matrix, k_range = 2:40) {
  dist_mat <- as.dist(1 - sim_matrix)
  hc <- hclust(dist_mat, method = "average")

  k_eval <- map_dfr(k_range, function(k) {
    cl <- cutree(hc, k = k)
    sil <- silhouette(cl, dist_mat)
    sizes <- table(cl)

    tibble(
      k = k,
      avg_silhouette = mean(sil[, "sil_width"]),
      n_singletons = sum(sizes == 1),
      prop_singletons = mean(sizes == 1)
    )
  })

  best_k <- k_eval %>%
    filter(!is.na(avg_silhouette)) %>%
    slice_max(avg_silhouette, n = 1) %>%
    pull(k)

  list(hc = hc, k_eval = k_eval, best_k = best_k)
}

# Extract GO terms by ontology
bp_terms <- get_terms(ORA_exp1_results, "biological_processes")
mf_terms <- get_terms(ORA_exp1_results, "molecular_functions")

# Build similarity matrices
sim_bp <- make_sim_matrix(bp_terms, sem_bp)
sim_mf <- make_sim_matrix(mf_terms, sem_mf)

# Choose best k separately for BP and MF
bp_fit <- evaluate_k(sim_bp, k_range = 2:40) # max index
mf_fit <- evaluate_k(sim_mf, k_range = 2:33) # max index

best_k_bp <- bp_fit$best_k
  # 29
best_k_mf <- mf_fit$best_k
  # 14

# Optional: inspect k selection
print(bp_fit$k_eval, n = 50)
print(mf_fit$k_eval, n = 50)

par(mfrow = c(2,1))
plot(bp_fit$k_eval$k, bp_fit$k_eval$avg_silhouette, type = "b", main = "bp clusters",
     xlab = "k-value tested", ylab = "Average silhouette")
abline(v = 29, lty = "dashed", col = "grey60")
plot(mf_fit$k_eval$k, mf_fit$k_eval$avg_silhouette, type = "b", main = "mf clusters",
     xlab = "k-value tested", ylab = "Average silhouette")
abline(v =14, lty = "dashed", col = "grey60")
par(mfrow = c(1,1))
  # WE want to maximise average silhouette since this means the clustering best fits the data
    # A rough guide:
    # 
    # > 0.5 = pretty strong clustering
    # 0.25 to 0.5 = moderate / usable structure # our clustering seems roughly at a usable level
    # < 0.25 = weak clustering structure
    # negative average = poor clustering overall

# save plot for reference:
{
png("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/Figures/k_selection_plots.png", width = 2600, height = 2000, res = 200)
par(mfrow = c(2,1))
plot(bp_fit$k_eval$k, bp_fit$k_eval$avg_silhouette, type = "b", main = "bp clusters",
     xlab = "k-value tested", ylab = "Average silhouette")
abline(v = 29, lty = "dashed", col = "grey60")
plot(mf_fit$k_eval$k, mf_fit$k_eval$avg_silhouette, type = "b", main = "mf clusters",
     xlab = "k-value tested", ylab = "Average silhouette")
abline(v =14, lty = "dashed", col = "grey60")
par(mfrow = c(1,1))
dev.off()
}

# Cut dendrograms using best k
bp_clusters <- cutree(bp_fit$hc, k = best_k_bp)
mf_clusters <- cutree(mf_fit$hc, k = best_k_mf)

# Combine cluster assignments
term_clusters <- bind_rows(
  tibble(ID = names(bp_clusters), semantic_cluster = bp_clusters),
  tibble(ID = names(mf_clusters), semantic_cluster = mf_clusters)
)

# Join clusters back onto ORA results
ORA_exp1_results_clustered <- ORA_exp1_results %>%
  mutate(
    ID = as.character(ID),
    GO = as.character(GO),
    Description_QuickGO = as.character(Description_QuickGO)
  ) %>%
  left_join(term_clusters, by = "ID") %>%
  arrange(GO, semantic_cluster, p.adjust)

# One representative term per cluster = lowest adjusted p-value
cluster_representatives <- ORA_exp1_results_clustered %>%
  group_by(GO, semantic_cluster) %>%
  slice_min(p.adjust, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  dplyr::select(GO, semantic_cluster, ID, Description_QuickGO, p.adjust, Count)

print(cluster_representatives, n = 50)

# Final summary table:
# - singleton terms are kept as-is
# - clustered terms keep only the best representative
ORA_exp1_summary_df <- ORA_exp1_results_clustered %>%
  group_by(GO, semantic_cluster) %>%
  mutate(cluster_size = n()) %>%
  arrange(p.adjust, .by_group = TRUE) %>%
  dplyr::slice(1) %>%
  ungroup()

print(ORA_exp1_summary_df, n = 50)
```

## Add LFC data

```{r SC_lfc_attach}
lfc_map <- setNames(DE_results_1$log2FoldChange, DE_results_1$X)

# Version 1: keep ORA rows intact, add a nested LFC column matching geneID order
ORA_exp1_combined_nested <- ORA_exp1_summary_df %>%
  mutate(
    geneID = as.character(geneID),
    log2FoldChange = sapply(strsplit(geneID, "/"), function(genes) {
      paste(lfc_map[genes], collapse = "/")
    })
  )
# Version 2: long format, one gene per row with its matching LFC
ORA_exp1_combined_long <- ORA_exp1_summary_df %>%
  mutate(geneID = as.character(geneID)) %>%
  separate_rows(geneID, sep = "/") %>%
  mutate(log2FoldChange = lfc_map[geneID])
```

## plot clustered results

```{r plot_clustered_results}
p_boxes <- ORA_exp1_combined_long %>%
  mutate(
    Description_QuickGO = fct_reorder(
      Description_QuickGO,
      log2FoldChange,
      .fun = median,
      .na_rm = TRUE
    )
  ) %>%
  ggplot(aes(
    x = log2FoldChange,
    y = Description_QuickGO
  )) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    linewidth = 0.5,
    colour = "grey60"
  ) +
  geom_boxplot(
    width = 0.7,
    outlier.size = 1.5,
    outlier.alpha = 0.8,
    linewidth = 0.40,
    colour = "grey20",
    fill = "grey90"
  ) +
  labs(
    x = "log2 fold-change (LFC) - Ashr Shrunk",
    y = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.y = element_text(size = 9, colour = "black"),
    axis.text.x = element_text(colour = "black"),
    axis.line = element_line(linewidth = 0.5),
    plot.margin = margin(5.5, 8, 5.5, 5.5)
  ) + 
  facet_wrap(~GO, scales = "free",
             labeller = labeller(
               GO = c(
                 biological_processes = "Biolocial Processes",
                 molecular_functions = "Molecular Functions"
               )
             )
            )
p_boxes
```

# Extra details

## Extra - Adding protein names to DEA results

```{r attach_protin_names}
# Load in DE results files
DE_results_1 <- read.table("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/DEA_results_exp1_no_outiler_and_shrunk_values.csv",
                         header = T, 
                         sep = ',', 
                         stringsAsFactors = TRUE)

DE_results_2 <- read.table("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/DEA_results_exp2_shrunk_values.csv",
                         header = T, 
                         sep = ',', 
                         stringsAsFactors = TRUE)

# only significant results
DE_results_1 <- as.data.frame(na.omit(DE_results_1[DE_results_1$padj <= 0.1,]))
DE_results_2 <- as.data.frame(na.omit(DE_results_2[DE_results_2$padj <= 0.1,]))

## Perform naming transformation to make gene IDs matching ----
DE_results_1$X <- gsub("\\.TU", "", DE_results_1$X) # Removing "TU" part of names
DE_results_1$X <- gsub("path", "mrna", DE_results_1$X) # Replacing "path" -> "mrna"
DE_results_2$X <- gsub("\\.TU", "", DE_results_2$X) # Removing "TU" part of names
DE_results_2$X <- gsub("path", "mrna", DE_results_2$X) # Replacing "path" -> "mrna"
names(DE_results_1)[names(DE_results_1) == "X"] <- "geneID"
names(DE_results_2)[names(DE_results_2) == "X"] <- "geneID"

# Keep only the best protein hit per gene (lowest e-value in V11)
matching_file_best <- matching_file %>%
  transmute(
    geneID = as.character(V1),
    proteinID = as.character(V13),
    evalue = as.numeric(V11)
  ) %>%
  group_by(geneID) %>%
  slice_min(order_by = evalue, n = 1, with_ties = FALSE) %>%
  ungroup()

# Named lookup: geneID -> proteinID
protein_map <- setNames(matching_file_best$proteinID, matching_file_best$geneID)

# Version 1: keep ORA rows intact, add protein IDs in geneID order
DE_results_1_with_protein_names <- DE_results_1 %>%
  mutate(
    geneID = as.character(geneID),
    proteinID = sapply(strsplit(geneID, "/"), function(genes) {
      paste(protein_map[genes], collapse = "/")
    })
  )

DE_results_2_with_protein_names <- DE_results_2 %>%
  mutate(
    geneID = as.character(geneID),
    proteinID = sapply(strsplit(geneID, "/"), function(genes) {
      paste(protein_map[genes], collapse = "/")
    })
  )

# re-order columns
DE_results_1_with_protein_names <- DE_results_1_with_protein_names %>%
  dplyr::select(geneID, proteinID, baseMean, log2FoldChange, lfcSE, pvalue, padj) %>%
  arrange(padj, desc(abs(log2FoldChange)))

DE_results_2_with_protein_names <- DE_results_2_with_protein_names %>%
  dplyr::select(geneID, proteinID, baseMean, log2FoldChange, lfcSE, pvalue, padj) %>%
  arrange(padj, desc(abs(log2FoldChange)))

# combined into 1 df
DE_results_with_protein_names <- rbind(DE_results_1_with_protein_names, 
                                       DE_results_2_with_protein_names)

# save files for supplementary materials:
write.csv(DE_results_with_protein_names, "C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/DE_results_with_protein_namess.csv", row.names = FALSE)

```


# Semantically clustering all enrichment analysis results together 

```{r enrichment_results_str}
# look at structures of all results files
str(ORA_exp1_combined_nested)
str(ORA_exp2_combined_nested)
str(GSEA_exp2_results_nested)

library(dplyr)
library(purrr)
library(cluster)
library(GOSemSim)
library(org.Aequina.eg.db)
library(ggVennDiagram)

# -----------------------------
# 1. Prepare and combine results
# -----------------------------

prep_results <- function(df, experiment, method) {
  df %>%
    transmute(
      ID = as.character(ID),
      Description = as.character(Description),
      Description_QuickGO = as.character(Description_QuickGO),
      ontology = as.character(GO),
      p.adjust = as.numeric(p.adjust),
      experiment = experiment,
      method = method
    ) %>%
    distinct()
}

result_list <- list(
  prep_results(ORA_exp1_combined_nested,  experiment = "exp1", method = "ORA"),
  prep_results(ORA_exp2_combined_nested,  experiment = "exp2", method = "ORA"),
  prep_results(GSEA_exp2_results_nested,  experiment = "exp2", method = "GSEA")
)

all_enrichment_results <- bind_rows(result_list)

# This is the minimal combined enrichment dataframe
all_enrichment_results_minimal <- all_enrichment_results %>%
  dplyr::select(ID, Description, Description_QuickGO, ontology, experiment, method, p.adjust)

# -----------------------------
# 2. Create unique term table for clustering
#    Keep one row per GO term per ontology
#    Use the lowest p.adjust across all sources
# -----------------------------

best_padj <- function(x) {
  if (all(is.na(x))) NA_real_ else min(x, na.rm = TRUE)
}

unique_terms <- all_enrichment_results_minimal %>%
  group_by(ontology, ID) %>%
  summarise(
    Description = dplyr::first(Description),
    Description_QuickGO = dplyr::first(Description_QuickGO),
    best_p.adjust = best_padj(p.adjust),
    .groups = "drop"
  )

# -----------------------------
# 3. Build semantic similarity data
# -----------------------------

sem_bp <- godata(org.Aequina.eg.db, keytype = "GID", ont = "BP", computeIC = FALSE)
sem_mf <- godata(org.Aequina.eg.db, keytype = "GID", ont = "MF", computeIC = FALSE)

make_sim_matrix <- function(term_ids, sem_data) {
  sim <- termSim(
    t1 = term_ids,
    t2 = term_ids,
    semData = sem_data,
    method = "Wang"
  )
  sim[is.na(sim)] <- 0
  diag(sim) <- 1
  sim
}

fit_semantic_clusters <- function(term_ids, sem_data, max_k = 40) {
  sim <- make_sim_matrix(term_ids, sem_data)
  n_terms <- length(term_ids)

  # handle very small term sets
  if (n_terms == 1) {
    return(list(
      sim = sim,
      hc = NULL,
      k_eval = tibble(k = 1, avg_silhouette = NA_real_),
      best_k = 1,
      clusters = setNames(1, term_ids)
    ))
  }

  dist_mat <- as.dist(1 - sim)
  hc <- hclust(dist_mat, method = "average")

  if (n_terms == 2) {
    clusters <- cutree(hc, k = 2)
    return(list(
      sim = sim,
      hc = hc,
      k_eval = tibble(k = 2, avg_silhouette = NA_real_),
      best_k = 2,
      clusters = clusters
    ))
  }

  k_grid <- 2:min(max_k, n_terms - 1)

  k_eval <- map_dfr(k_grid, function(k) {
    cl <- cutree(hc, k = k)
    sil <- silhouette(cl, dist_mat)
    sizes <- table(cl)

    tibble(
      k = k,
      avg_silhouette = mean(sil[, "sil_width"]),
      n_singletons = sum(sizes == 1),
      prop_singletons = mean(sizes == 1)
    )
  })

  best_k <- k_eval %>%
    filter(!is.na(avg_silhouette)) %>%
    slice_max(order_by = avg_silhouette, n = 1, with_ties = FALSE) %>%
    pull(k)

  clusters <- cutree(hc, k = best_k)

  list(
    sim = sim,
    hc = hc,
    k_eval = k_eval,
    best_k = best_k,
    clusters = clusters
  )
}

# -----------------------------
# 4. Fit BP and MF clustering separately
# -----------------------------

bp_terms <- unique_terms %>%
  filter(ontology == "biological_processes") %>%
  pull(ID)

mf_terms <- unique_terms %>%
  filter(ontology == "molecular_functions") %>%
  pull(ID)

bp_fit <- fit_semantic_clusters(bp_terms, sem_bp, max_k = 40)
mf_fit <- fit_semantic_clusters(mf_terms, sem_mf, max_k = 40)

bp_fit$best_k
  # 31
mf_fit$best_k
  # 15

# Optional: inspect the k selection tables
print(bp_fit$k_eval, n = 50)
print(mf_fit$k_eval, n = 50)

# Optional: plot k selection
par(mfrow = c(2, 1))

plot(
  bp_fit$k_eval$k, bp_fit$k_eval$avg_silhouette,
  type = "b", main = "BP clusters",
  xlab = "k tested", ylab = "Average silhouette"
)
abline(v = bp_fit$best_k, lty = "dashed", col = "grey60")

plot(
  mf_fit$k_eval$k, mf_fit$k_eval$avg_silhouette,
  type = "b", main = "MF clusters",
  xlab = "k tested", ylab = "Average silhouette"
)
abline(v = mf_fit$best_k, lty = "dashed", col = "grey60")

par(mfrow = c(1, 1))

# -----------------------------
# 5. Assign clusters to each unique GO term
# -----------------------------

term_clusters <- bind_rows(
  tibble(
    ontology = "biological_processes",
    ID = names(bp_fit$clusters),
    semantic_cluster = as.integer(bp_fit$clusters)
  ),
  tibble(
    ontology = "molecular_functions",
    ID = names(mf_fit$clusters),
    semantic_cluster = as.integer(mf_fit$clusters)
  )
)

unique_terms_clustered <- unique_terms %>%
  left_join(term_clusters, by = c("ontology", "ID"))

# -----------------------------
# 6. Pick one representative term per cluster
#    Representative = lowest adjusted p-value
# -----------------------------

cluster_representatives <- unique_terms_clustered %>%
  group_by(ontology, semantic_cluster) %>%
  slice_min(order_by = best_p.adjust, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(
    ontology,
    semantic_cluster,
    representative_ID = ID,
    representative_Description = Description,
    representative_Description_QuickGO = Description_QuickGO
  )

# Map every term to its representative
term_to_representative <- unique_terms_clustered %>%
  left_join(cluster_representatives, by = c("ontology", "semantic_cluster"))

# -----------------------------
# 7. Replace clustered terms in the combined enrichment results
# -----------------------------

all_enrichment_results_replaced <- all_enrichment_results_minimal %>%
  left_join(
    term_to_representative %>%
      dplyr::select(
        ontology, ID,
        semantic_cluster,
        representative_ID,
        representative_Description,
        representative_Description_QuickGO
      ),
    by = c("ontology", "ID")
  ) %>%
  mutate(
    original_ID = ID,
    original_Description = Description,
    original_Description_QuickGO = Description_QuickGO,
    ID = representative_ID,
    Description = representative_Description,
    Description_QuickGO = representative_Description_QuickGO
  ) %>%
  dplyr::select(
    experiment, method, ontology,
    original_ID, original_Description, original_Description_QuickGO,
    semantic_cluster,
    ID, Description, Description_QuickGO,
    p.adjust
  )

# Remove duplicate representative terms within each experiment/method/ontology
all_enrichment_results_collapsed <- all_enrichment_results_replaced %>%
  distinct(experiment, method, ontology, ID, .keep_all = TRUE)

# -----------------------------
# 8. Make GO sets for overlap analysis
# -----------------------------

make_go_sets <- function(df, ontology_name) {
  df %>%
    filter(ontology == ontology_name) %>%
    mutate(source = paste(experiment, method, sep = "_")) %>%
    distinct(source, ID) %>%
    split(.$source) %>%
    map(~ .x$ID)
}

bp_sets <- make_go_sets(all_enrichment_results_collapsed, "biological_processes")
mf_sets <- make_go_sets(all_enrichment_results_collapsed, "molecular_functions")

# -----------------------------
# 9. Plot venn diagrams
# -----------------------------

# options
names(bp_sets) <- c("Acute ORA", "Primed GSEA", "Primed ORA")
bp_sets <- bp_sets[c("Acute ORA", "Primed ORA", "Primed GSEA")]
names(mf_sets) <- c("Acute ORA", "Primed ORA")
mf_sets <- mf_sets[c("Acute ORA", "Primed ORA")]

p_bp <- ggVennDiagram(
  bp_sets,
  category.names = names(bp_sets),
  set_color = "black",
      set_size = 3.5,
  label_alpha = 0
) +
  scale_fill_gradient(low = "white", high = "#4c43e8") +
  #coord_equal(clip = "off") +
  #ggtitle("Biological processes") +
  theme_void(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.margin = margin(20, 30, 20, 30),
    legend.position = "none"
  )

p_bp

p_mf <- ggVennDiagram(
  mf_sets,
  category.names = names(mf_sets),
  set_color = "black",
    set_size = 3.5,
  label_alpha = 0
) +
  scale_fill_gradient(low = "white", high = "#e8a943") +
  coord_equal(clip = "off") +
  #ggtitle("Biological processes") +
  theme_void(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.margin = margin(20, 25, 20, 25),
    legend.position = "none"
  )

p_mf

# Optional: save venn diagrams
ggsave("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/Figures/Venn Diagrams/bp_venn_semantic_clusters_diagram.png", p_bp, width = 6, height = 6, dpi = 300)
ggsave("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/Figures/Venn Diagrams/mf_venn_semantic_clusters_diagram.png", p_mf, width = 6, height = 6, dpi = 300)

# Capture intersctions ----
# optional: make nicer source labels
intersection_df <- all_enrichment_results_collapsed %>%
  mutate(
    source = case_when(
      experiment == "exp1" & method == "ORA"  ~ "Experiment 1 ORA",
      experiment == "exp2" & method == "ORA"  ~ "Experiment 2 ORA",
      experiment == "exp2" & method == "GSEA" ~ "Experiment 2 GSEA",
      TRUE ~ paste(experiment, method)
    )
  ) %>%
  group_by(ontology, ID) %>%
  summarise(
    Description = dplyr::first(Description),
    Description_QuickGO = dplyr::first(Description_QuickGO),
    groups_intersecting = paste(sort(unique(source)), collapse = " | "),
    n_groups = n_distinct(source),
    .groups = "drop"
  ) %>%
  filter(n_groups > 1) %>%   # remove non-intersecting terms
  arrange(ontology, desc(n_groups), ID)

print(intersection_df)

library(dplyr)
library(gridExtra)
library(grid)

intersection_table <- intersection_df %>%
  dplyr::select(ontology, ID, Description_QuickGO, groups_intersecting, n_groups)

tbl <- tableGrob(intersection_table, rows = NULL)

grid.newpage()
grid.draw(tbl)
```


```{r}
library(dplyr) 
library(stringr) 
library(gridExtra) 
library(grid) 

# Clean and relabel table content
intersection_table_pub <- intersection_df %>% 
  mutate(ontology = recode(ontology, 
                           biological_processes = "Biological process", 
                           molecular_functions = "Molecular function" ), 
         groups_intersecting = groups_intersecting 
         %>% str_replace_all("Experiment 1 ORA", 
                             "Experiment 1 (ORA)") 
         %>% str_replace_all("Experiment 2 ORA", 
                             "Experiment 2 (ORA)") %>% 
           str_replace_all("Experiment 2 GSEA", 
                           "Experiment 2 (GSEA)") %>% 
           str_replace_all(" \\| ", ", "), 
         Description_QuickGO = str_replace_all(Description_QuickGO, "_", " ") ) %>% 
  transmute( Ontology = ontology, 
             `GO ID` = ID, 
             `GO term` = str_wrap(Description_QuickGO, 
                                  width = 32), 
             `Shared between` = str_wrap(groups_intersecting, 
                                         width = 38), 
             `No. of groups` = n_groups ) %>% 
  arrange(Ontology, desc(`No. of groups`), 
          `GO ID`) 
# Table theme 
table_theme <- ttheme_minimal( base_size = 10, padding = unit(c(6, 6), "mm"), core = list( fg_params = list(hjust = 0, x = 0.02, fontsize = 10), bg_params = list(fill = c("white", "#F7F7F7"), col = NA) ), colhead = list( fg_params = list(fontface = "bold", fontsize = 10.5, col = "black"), bg_params = list(fill = "#D9D9D9", col = NA) ) ) 
# Build table 

tbl <- tableGrob( intersection_table_pub, rows = NULL, theme = table_theme ) 
# Manually widen columns for readability 
tbl$widths <- unit(c(1.8, 1.1, 3.0, 3.4, 1.0), "in") 

# Draw table 
grid.newpage() 
grid.draw(tbl)

### Save as PDF 
pdf("intersection_table_publication.pdf", width = 10.5, height = 3.2) 
grid.draw(tbl) 
dev.off() 

### Save as PNG `

png("intersection_table_publication.png", width = 3200, 
    height = 1000, 
    res = 300) 
grid.draw(tbl) 
dev.off() 


# Plot venns and table together ----
# Convert ggplot objects to grobs
p_bp <- p_bp +
  theme(
    plot.margin = margin(20, 5, 20, 10)   # top, right, bottom, left
  )

p_mf <- p_mf +
  theme(
    plot.margin = margin(20, 10, 20, 5)
  )

bp_grob <- ggplotGrob(p_bp)
mf_grob <- ggplotGrob(p_mf)

# Layout:
# row 1 = bp_grob | mf_grob
# row 2 = tbl spans both columns
combined_plot <- arrangeGrob(
  bp_grob,
  mf_grob,
  tbl,
  layout_matrix = rbind(
    c(1, 2),
    c(3, 3)
  ),
  heights = c(2.2, 1.4)   # adjust this to give more/less room to the table
)

# Draw combined figure
grid.newpage()
grid.draw(combined_plot)

png("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/Figures/Venn Diagrams/venn_and_intersection_table.png", width = 4200, height = 3000, res = 300)
grid.draw(combined_plot)
dev.off()
```