---
title: "Enrichment_Analysis_C1"
author: "Hamish Williams"
date: "`r Sys.Date()`"
output: 
  html_document: 
    fig_height: 7
    toc: true
    toc_depth: 3
    toc_float: true
editor_options: 
  chunk_output_type: console
---

# Overview

This Markdown script will walk through all of the necessary steps undertaken to complete the **Gene Set Enrichment Analysis** and Over-representation **analysis** for the paired exposure experiment. This follows immediately from the **Differential Expression Analysis** results, and uses them to complete this analysis.

I was finding that even with the previous markdown sheet, that I generated a massive markdown sheet which was not easy to follow, which made it very difficult to understand the flow of the analysis, and at what stage of the process I was in. I have gone away and created a nice flow diagram which will neatly show the general flow of the analysis, and should make it much easier for myself and any other researchers who read this markdown to understand the process, and the steps undertaken to perform this analysis.

However, this is not a general guide to GSEA and Enrichment based analysis. Although I think that these types of content are very lacking in the scientific sphere, this script is not to be used as a single end all be all one step guide to completing this analysis. This markdown will wlakthrough the steps I have taken for my analysis, explaining and justifying parameter settings and additional actions required to complete the analysis. Experiments and analysis as a whole should be tailored to each experiment, and consideration of how best to tune the analyses to that experiment is important for statistical design, and for answering the specific questions of one's own research aims.

# Building Functions

## check_packages_ClusterProfilerR & Loading Packages

Function built to make loading and downloading of packages easier for reproducibility.

```{r check_packages_ClusterProfilerR, message=FALSE, warning=FALSE, results='hide'}
check_packages_ClusterProfileR <- function(pkg_list) {
  for (pkg in pkg_list) {
    if (!require(pkg, character.only = TRUE)) {
      print(paste(pkg, "is not installed, installing package..."))
      install.packages(pkg)
    } else {
      suppressMessages(library(pkg, character.only = TRUE))
      print(paste(pkg, "is installed and loaded"))
    }
  }
}

# Generate a dataframe with a list of packages required

pkg_list_ClusterProfileR<-c("tidyverse", # manipulations
                            "tibble", # tables
                            "stats", # stats
                            "ggplot2", # plottin
                            "dplyr", # manipulation
                            "tidyr", # manipulation
                            "ggfortify", # plotting
                            "knitr",# markdown
                            "clusterProfiler", # enrichment analysis
                            "enrichplot", # plotting
                            "data.table", # data formatting
                            "plotly", # plotting
                            "ggVennDiagram", # plotting
                            "VennDiagram", # plotting
                            "limma", # plotting
                            "ggridges", # plotting
                            "stringr") # manipulations

# Run Function::
check_packages_ClusterProfileR(pkg_list_ClusterProfileR)
```

## convertFractionToNumeric

Used to convert gene ratios in final analyses to numerical values for generating figures

```{r convertFractionToNumeric, message=FALSE, warning=FALSE}
convertFractionToNumeric <- function(fraction) {
  fraction_parts <- strsplit(fraction, "/")[[1]]
  numerator <- as.numeric(fraction_parts[1])
  denominator <- as.numeric(fraction_parts[2])
  numeric_value <- numerator / denominator
  return(numeric_value)
}
```

# Importing and Formatting Data

As outlined by the flow diagram, there are 3 main pieces of data needed to start the analysis. This include **Gene Ontology** data file (UNIPROT_results.txt), which contains the GO data for Biological Processes and Molecular Functions for the blastp matched genes of the UNIPROT database; a **Matching File**, which contains the names of the genes from *A. equina* vs. the Blastp result names - will be used to match the gene names to the blastp results; and finally the **Differential Expression Results**, which includes both the experiments completed (short and long)

```{r Importing_Data, results='hide'}
# Import BP and MF terms file ----
BP_MF_Terms <- read.table("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Data/UNIPROT_results.txt",
                          header = T, 
                          sep = '\t', 
                          stringsAsFactors = FALSE)

# Import the blastp (matching) file ----
matching_file <- read.table("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Data/genome_matching_file.txt", 
                            header = F, 
                            sep = '\t', 
                            stringsAsFactors = FALSE)

# Change the name of the columns to be identical ----
  names(matching_file)[names(matching_file) == "V13"] <- "ID"
  names(BP_MF_Terms)[names(BP_MF_Terms) == "From"] <- "ID"
  names(BP_MF_Terms)[names(BP_MF_Terms) == "Gene.Names"] <- "GENENAMES"
  names(BP_MF_Terms)[names(BP_MF_Terms) == "Gene.Ontology..biological.process."] <- "GO:BP"
  names(BP_MF_Terms)[names(BP_MF_Terms) == "Gene.Ontology..molecular.function."] <- "GO:MF"
  
# Merge blastp & GO terms ----
  # Merge DEG Files:
merged_df <- merge(matching_file, BP_MF_Terms, by = "ID")
# Subset Dataframe
merged_df <- subset(merged_df[, c("V1","ID","GO:BP","GO:MF")])
# Change Variable names
names(merged_df)[names(merged_df) == "V1"] <- "GENE"

# Separate BP and MF GOs ----
# Select BP column:
BP <- merged_df[,c("GENE","ID","GO:BP")]
BP <- BP[!apply(BP == "", 1, any), ]
BP <- BP[!duplicated(BP), ]
BP <- na.omit(BP)
# Extract GO terms:
GENES_BP <- BP
GENES_BP$GO <- ""
bracketed_rows <- grepl("\\[.*\\]", GENES_BP$`GO:BP`)
GENES_BP$GO[bracketed_rows] <- gsub(".*\\[(.*)\\].*", "\\1", GENES_BP$`GO:BP`[bracketed_rows])
GENES_BP <- GENES_BP[, c("GO","GENE")]
GENES_BP <- GENES_BP[!duplicated(GENES_BP), ]
GENES_BP <- GENES_BP[complete.cases(GENES_BP),]

## Extract Terms:
TERMs_BP <- BP
TERMs_BP$GO <- ""
bracketed_rows <- grepl("\\[.*\\]", TERMs_BP$`GO:BP`)
TERMs_BP$GO[bracketed_rows] <- gsub(".*\\[(.*)\\].*", "\\1", TERMs_BP$`GO:BP`[bracketed_rows])
TERMs_BP$`GO:BP`[bracketed_rows] <- gsub("\\[.*\\]", "", TERMs_BP$`GO:BP`[bracketed_rows])
TERMs_BP <- TERMs_BP[, c("GO","GO:BP")]
TERMs_BP <- TERMs_BP[!duplicated(TERMs_BP), ]
names(TERMs_BP)[names(TERMs_BP) == "GO:BP"] <- "GENENAMES_BP"
TERMs_BP <- TERMs_BP[complete.cases(TERMs_BP),]

# Select MF column:
MF <- merged_df[,c("GENE","ID","GO:MF")]
MF <- MF[!apply(MF == "", 1, any), ]
MF <- MF[!duplicated(MF), ]
MF <- na.omit(MF)
# Extract GO terms:
GENES_MF <- MF
GENES_MF$GO <- ""
bracketed_rows <- grepl("\\[.*\\]", GENES_MF$`GO:MF`)
GENES_MF$GO[bracketed_rows] <- gsub(".*\\[(.*)\\].*", "\\1", GENES_MF$`GO:MF`[bracketed_rows])
GENES_MF <- GENES_MF[, c("GO","GENE")]
GENES_MF <- GENES_MF[!duplicated(GENES_MF), ]
GENES_MF <- GENES_MF[complete.cases(GENES_MF),]

## Extract Terms:
TERMs_MF <- MF
TERMs_MF$GO <- ""
bracketed_rows <- grepl("\\[.*\\]", TERMs_MF$`GO:MF`)
TERMs_MF$GO[bracketed_rows] <- gsub(".*\\[(.*)\\].*", "\\1", TERMs_MF$`GO:MF`[bracketed_rows])
TERMs_MF$`GO:MF`[bracketed_rows] <- gsub("\\[.*\\]", "", TERMs_MF$`GO:MF`[bracketed_rows])
TERMs_MF <- TERMs_MF[, c("GO","GO:MF")]
TERMs_MF <- TERMs_MF[!duplicated(TERMs_MF), ]
names(TERMs_MF)[names(TERMs_MF) == "GO:MF"] <- "GENENAMES_MF"
TERMs_MF <- TERMs_MF[complete.cases(TERMs_MF),]

# Import DEA results ----
DE_results_1 <- read.table("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/DEA_results_exp1_no_outiler_and_shrunk_values.csv",
                         header = T, 
                         sep = ',', 
                         stringsAsFactors = TRUE)

DE_results_2 <- read.table("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/DEA_results_exp2_shrunk_values.csv",
                         header = T, 
                         sep = ',', 
                         stringsAsFactors = TRUE)

## Perform naming transformation to make gene IDs matching ----
DE_results_1$X <- gsub("\\.TU", "", DE_results_1$X) # Removing "TU" part of names
DE_results_1$X <- gsub("path", "mrna", DE_results_1$X) # Replacing "path" -> "mrna"
DE_results_2$X <- gsub("\\.TU", "", DE_results_2$X) # Removing "TU" part of names
DE_results_2$X <- gsub("path", "mrna", DE_results_2$X) # Replacing "path" -> "mrna"

# DEGs lists for ORA ----
DEGs <- DE_results_1[(DE_results_1$padj < 0.1),]
DEGs <- DEGs[complete.cases(DEGs),]
DEGs_1 <- as.vector(DEGs[,1])
length(DEGs_1)
DEGs <- DE_results_2[(DE_results_2$padj < 0.1),]
DEGs <- DEGs[complete.cases(DEGs),]
DEGs_2 <- as.vector(DEGs[,1])
length(DEGs_2)

# Create geneList objects for GSEA analyses ----
# Feature 1: numeric vector (LFC)
geneList_1 = DE_results_1[,3]
geneList_2 = DE_results_2[,3]
# Feature 2: Named Vector (rownames)
names(geneList_1) = as.character(DE_results_1[,1])
names(geneList_2) = as.character(DE_results_2[,1])
# feature 3: decreasing order;
geneList_1 = sort(geneList_1, decreasing = TRUE)
geneList_2 = sort(geneList_2, decreasing = TRUE)
```

## Tidy up Environment

```{r Tidy_up, results='hide'}
# Tidy up environment..
    rm(MF,
       BP,
       BP_MF_Terms,
       matching_file,
       DEGs,
       merged_df,
       bracketed_rows
       )

# Save Environment
save.image(file = "ORA_exp1_prep.RData")
```

# Overrepresentation Analysis

## Experiment 1

```{r ORA_exp1, results='hide'}
# Load environment for analysis
load("ORA_exp1_prep.RData")

# Run ORA
set.seed(100)
ORA_exp1_BP <- enricher(DEGs_1, 
                TERM2GENE = GENES_BP,
                TERM2NAME = TERMs_BP,
                minGSSize=1)

set.seed(100)
ORA_exp1_MF <- enricher(DEGs_1, 
                TERM2GENE = GENES_MF,
                TERM2NAME = TERMs_MF,
                minGSSize=1)

# Extract results and add in GO identifiers
ORA_exp1_BP_results <- ORA_exp1_BP@result
ORA_exp1_BP_results$GO <- "biological_processes"

ORA_exp1_MF_results <- ORA_exp1_MF@result
ORA_exp1_MF_results$GO <- "molecular_functions"

# combine results and subset for significant values
ORA_exp1_results <- rbind(ORA_exp1_BP_results, ORA_exp1_MF_results)
rm(ORA_exp1_BP_results, ORA_exp1_MF_results, ORA_exp1_BP, ORA_exp1_MF)
ORA_exp1_results <- subset(ORA_exp1_results[ORA_exp1_results$p.adjust < 0.1,])

# Change fractions to numeric values
ORA_exp1_results$GeneRatio <- sapply(ORA_exp1_results$GeneRatio, convertFractionToNumeric)
ORA_exp1_results$BgRatio <- sapply(ORA_exp1_results$BgRatio, convertFractionToNumeric)

## Export Dataframe results
write.csv(ORA_exp1_results, "C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/ORA_exp1_results.csv", row.names = FALSE)

## ... Replace GO descriptions with matched descriptions of GO:ID from QuickGO
  # import results with QuickGO descriptions

ORA_exp1_results <- read.table("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/ORA_exp1_results_QuickGO_names.csv",
                         header = T, 
                         sep = ',', 
                         stringsAsFactors = TRUE)

# Map the LFC values to ORA results using the gene names ----
# Make a lookup table: gene -> log2FoldChange
lfc_map <- setNames(DE_results_1$log2FoldChange, DE_results_1$X)

# Version 1: keep ORA rows intact, add a nested LFC column matching geneID order
ORA_exp1_combined_nested <- ORA_exp1_results %>%
  mutate(
    geneID = as.character(geneID),
    log2FoldChange = sapply(strsplit(geneID, "/"), function(genes) {
      paste(lfc_map[genes], collapse = "/")
    })
  )
# Version 2: long format, one gene per row with its matching LFC
ORA_exp1_combined_long <- ORA_exp1_results %>%
  mutate(geneID = as.character(geneID)) %>%
  separate_rows(geneID, sep = "/") %>%
  mutate(log2FoldChange = lfc_map[geneID])
```


## Experiment 2
```{r ORA_exp2, results='hide'}
# Load environment for analysis
#load("ORA_exp1_prep.RData")

# Run ORA
set.seed(100)
ORA_exp2_BP <- enricher(DEGs_2, 
                TERM2GENE = GENES_BP,
                TERM2NAME = TERMs_BP,
                minGSSize=1)

set.seed(100)
ORA_exp2_MF <- enricher(DEGs_2, 
                TERM2GENE = GENES_MF,
                TERM2NAME = TERMs_MF,
                minGSSize=1)

# Extract results and add in GO identifiers
ORA_exp2_BP_results <- ORA_exp2_BP@result
ORA_exp2_BP_results$GO <- "biological_processes"

ORA_exp2_MF_results <- ORA_exp2_MF@result
ORA_exp2_MF_results$GO <- "molecular_functions"

# combine results and subset for significant values
ORA_exp2_results <- rbind(ORA_exp2_BP_results, ORA_exp2_MF_results)
rm(ORA_exp2_BP_results, ORA_exp2_MF_results, ORA_exp2_BP, ORA_exp2_MF)
ORA_exp2_results <- subset(ORA_exp2_results[ORA_exp2_results$p.adjust < 0.1,])

# Change fractions to numeric values
ORA_exp2_results$GeneRatio <- sapply(ORA_exp2_results$GeneRatio, convertFractionToNumeric)
ORA_exp2_results$BgRatio <- sapply(ORA_exp2_results$BgRatio, convertFractionToNumeric)

## Export Dataframe results
write.csv(ORA_exp2_results, "C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/ORA_exp2_results.csv", row.names = FALSE)

## ... Replace GO descriptions with matched descriptions of GO:ID from QuickGO
  # import results with QuickGO descriptions

ORA_exp2_results <- read.table("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/ORA_exp2_results_QuickGO_names.csv",
                         header = T, 
                         sep = ',', 
                         stringsAsFactors = TRUE)

# Map the LFC values to ORA results using the gene names ----
# Make a lookup table: gene -> log2FoldChange
lfc_map <- setNames(DE_results_2$log2FoldChange, DE_results_2$X)

# Version 1: keep ORA rows intact, add a nested LFC column matching geneID order
ORA_exp2_combined_nested <- ORA_exp2_results %>%
  mutate(
    geneID = as.character(geneID),
    log2FoldChange = sapply(strsplit(geneID, "/"), function(genes) {
      paste(lfc_map[genes], collapse = "/")
    })
  )
# Version 2: long format, one gene per row with its matching LFC
ORA_exp2_combined_long <- ORA_exp2_results %>%
  mutate(geneID = as.character(geneID)) %>%
  separate_rows(geneID, sep = "/") %>%
  mutate(log2FoldChange = lfc_map[geneID])
```

# Gene Set Enrichment Analysis

## Experiment 1
```{r GSEA_exp1}
# Load Environment
# load("GSEA_exp1.RData")

# Run GSEA
set.seed(100)
GSEA_exp1_BP<- GSEA(geneList_1,
           TERM2GENE = GENES_BP,
           TERM2NAME = TERMs_BP,
           pvalueCutoff = 0.1)
set.seed(100)
GSEA_exp1_MF<- GSEA(geneList_1,
           TERM2GENE = GENES_MF,
           TERM2NAME = TERMs_MF,
           pvalueCutoff = 0.1)

# Get Results
GSEA_exp1_BP_results <- GSEA_exp1_BP@result
GSEA_exp1_BP_results$GO <- "biological_processes"

GSEA_exp1_MF_results <- GSEA_exp1_MF@result
GSEA_exp1_MF_results$GO <- "molecular_function"

# Merge results
GSEA_exp1_results <- rbind(GSEA_exp1_BP_results, GSEA_exp1_MF_results)

# Subset for only significant results
GSEA_exp1_results <- subset(GSEA_exp1_results[GSEA_exp1_results$p.adjust < 0.1,])

# Combine all results into single listed df
GSEA_exp1 <- list("BP" = GSEA_exp1_BP,
                  "MF" = GSEA_exp1_MF,
                  "Results" = GSEA_exp1_results)

# Clean-up 
rm(GSEA_exp1_MF_results, 
   GSEA_exp1_BP_results,
   GSEA_exp1_results,
   GSEA_exp1_BP,
   GSEA_exp1_MF)

## Export Dataframe results
write.csv(GSEA_exp1[["Results"]], "C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/GSEA_exp1_results.csv", row.names = FALSE)

# Capture QuickGO names... re-import
  # no results...
```

## Experiment 2
```{r GSEA_exp2}
# Load Environment
# load("GSEA_exp2.RData")

# Run GSEA
set.seed(100)
GSEA_exp2_BP<- GSEA(geneList_2,
           TERM2GENE = GENES_BP,
           TERM2NAME = TERMs_BP,
           pvalueCutoff = 0.1)
set.seed(100)
GSEA_exp2_MF<- GSEA(geneList_2,
           TERM2GENE = GENES_MF,
           TERM2NAME = TERMs_MF,
           pvalueCutoff = 0.1)

# Get Results
GSEA_exp2_BP_results <- GSEA_exp2_BP@result
GSEA_exp2_BP_results$GO <- "biological_processes"

GSEA_exp2_MF_results <- GSEA_exp2_MF@result
GSEA_exp2_MF_results$GO <- "molecular_function"

# Merge results
GSEA_exp2_results <- rbind(GSEA_exp2_BP_results, GSEA_exp2_MF_results)

# Subset for only significant results
GSEA_exp2_results <- subset(GSEA_exp2_results[GSEA_exp2_results$p.adjust < 0.1,])

# Combine all results into single listed df
GSEA_exp2 <- list("BP" = GSEA_exp2_BP,
                  "MF" = GSEA_exp2_MF,
                  "Results" = GSEA_exp2_results)

# Clean-up 
rm(GSEA_exp2_MF_results, 
   GSEA_exp2_BP_results,
   GSEA_exp2_results,
   GSEA_exp2_BP,
   GSEA_exp2_MF)

## Export Dataframe results
write.csv(GSEA_exp2[["Results"]], "C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/GSEA_exp2_results.csv", row.names = FALSE)

# Capture QuickGO names... re-import
GSEA_exp2_results <- read.table("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 1/Results/GSEA_exp2_results_QuickGO_names.csv",
                         header = T, 
                         sep = ',', 
                         stringsAsFactors = TRUE)

# nested and long versions
# Make a lookup table: gene -> log2FoldChange
lfc_map <- setNames(DE_results_2$log2FoldChange, DE_results_2$X)

# Version 1: keep ORA rows intact, add a nested LFC column matching geneID order
GSEA_exp2_results_nested <- GSEA_exp2_results %>%
  mutate(
    core_enrichment = as.character(core_enrichment),
    log2FoldChange = sapply(strsplit(core_enrichment, "/"), function(genes) {
      paste(lfc_map[genes], collapse = "/")
    })
  )
# Version 2: long format, one gene per row with its matching LFC
GSEA_exp2_results_long <- GSEA_exp2_results %>%
  mutate(core_enrichment = as.character(core_enrichment)) %>%
  separate_rows(core_enrichment, sep = "/") %>%
  mutate(log2FoldChange = lfc_map[core_enrichment])
```

# Plotting Figures

```{r enrichment_plots}
p_boxes <- ORA_exp1_combined_long %>%
  mutate(
    Description_QuickGO = fct_reorder(
      Description_QuickGO,
      log2FoldChange,
      .fun = median,
      .na_rm = TRUE
    )
  ) %>%
  ggplot(aes(
    x = log2FoldChange,
    y = Description_QuickGO
  )) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    linewidth = 0.5,
    colour = "grey60"
  ) +
  geom_boxplot(
    width = 0.7,
    outlier.size = 1.5,
    outlier.alpha = 0.8,
    linewidth = 0.40,
    colour = "grey20",
    fill = "grey90"
  ) +
  labs(
    x = "log2 fold-change (LFC) - Ashr Shrunk",
    y = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.y = element_text(size = 9, colour = "black"),
    axis.text.x = element_text(colour = "black"),
    axis.line = element_line(linewidth = 0.5),
    plot.margin = margin(5.5, 8, 5.5, 5.5)
  ) + 
  facet_wrap(~GO, scales = "free",
             labeller = labeller(
               GO = c(
                 biological_processes = "Biolocial Processes",
                 molecular_functions = "Molecular Functions"
               )
             )
            )
p_boxes


## GSEA plot ----
p_boxes <- GSEA_exp2_results_long %>%
  mutate(
    Description_QuickGO = fct_reorder(
      Description_QuickGO,
      log2FoldChange,
      .fun = median,
      .na_rm = TRUE
    )
  ) %>%
  ggplot(aes(
    x = log2FoldChange,
    y = Description_QuickGO
  )) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    linewidth = 0.5,
    colour = "grey60"
  ) +
  geom_boxplot(
    width = 0.7,
    outlier.size = 1.5,
    outlier.alpha = 0.8,
    linewidth = 0.40,
    colour = "grey20",
    fill = "grey90"
  ) +
  labs(
    x = "log2 fold-change (LFC) - Ashr Shrunk",
    y = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.y = element_text(size = 9, colour = "black"),
    axis.text.x = element_text(colour = "black"),
    axis.line = element_line(linewidth = 0.5),
    plot.margin = margin(5.5, 8, 5.5, 5.5)
  ) + 
  facet_wrap(~GO, scales = "free",
             labeller = labeller(
               GO = c(
                 biological_processes = "Biolocial Processes",
                 molecular_functions = "Molecular Functions"
               )
             )
            )
p_boxes

```