---
# ============================================================
# Calculate whether each GSEA ridge peaks below or above zero
# ============================================================

get_density_peak <- function(x) {
  x <- x[!is.na(x)]

  if (length(unique(x)) < 2) {
    return(NA_real_)
  }

  d <- density(x)
  d$x[which.max(d$y)]
}

ridge_peak_df <- GSEA_plot_df %>%
  dplyr::group_by(contrast, term) %>%
  dplyr::summarise(
    peak_lfc = get_density_peak(lfc),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    peak_direction = dplyr::case_when(
      is.na(peak_lfc) ~ "Peak unavailable",
      peak_lfc < 0   ~ "Peak < 0",
      peak_lfc > 0   ~ "Peak > 0",
      TRUE           ~ "Peak = 0"
    )
  )

GSEA_plot_df_coloured <- GSEA_plot_df %>%
  dplyr::left_join(
    ridge_peak_df,
    by = c("contrast", "term")
  )---
title: "Chapter 3 - Enrichment Analyses"
author: "Hamish Williams"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

RMarkdown Script for the enrichment analyses portion of this chapter, allows me to make easier comments on whats going on

# Load Packages

```{r packages}
# Load packages ----
check_packages <- function(pkg_list) {
  for (pkg in pkg_list) {
    if (!require(pkg, character.only = TRUE)) {
      print(paste(pkg, "is not installed, installing package..."))
      install.packages(pkg)
    } else {
      suppressMessages(library(pkg, character.only = TRUE))
      print(paste(pkg, "is installed and loaded"))
    }
  }
}

# Generate a dataframe with a list of packages required
pkg_list<-c("tidyverse", 
                            "tibble", 
                            "stats",
                            "ggplot2", 
                            "dplyr", 
                            "tidyr", 
                            "ggfortify", 
                            "knitr",
                            "clusterProfiler",
                            #"AnnotationForge",
                            "enrichplot",
                            "data.table",
                            "stats",
                            "plotly",
                            "ggVennDiagram",
                            "VennDiagram",
                            "limma",
                            "ggridges",
                            # "pathview",
                            "KEGGREST",
                            "stringr",
            "forcats",
            "ComplexHeatmap",
            "grid",
            "viridisLite",
            "RColorBrewer",
            "circlize")

# Run Function::
check_packages(pkg_list)

# Generate Functions ----
convertFractionToNumeric <- function(fraction) {
  fraction_parts <- strsplit(fraction, "/")[[1]]
  numerator <- as.numeric(fraction_parts[1])
  denominator <- as.numeric(fraction_parts[2])
  numeric_value <- numerator / denominator
  return(numeric_value)
}
```

## Extract GO terms

```{r get_GO_data}
## Merge Blastp and GO terms files together
BP_MF_Terms <- read.table(
  "C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_2/data/temp_storage/UNIPROT_results.txt",
  header = TRUE,
  sep = "\t",
  stringsAsFactors = FALSE,
  quote = ""
)

matching_file <- read.table(
  "C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_2/data/temp_storage/genome_matching_file.txt",
  header = FALSE,
  sep = "\t",
  stringsAsFactors = FALSE
)

# Change the name of the columns to be identical ----
{
  names(matching_file)[names(matching_file) == "V13"] <- "ID"
  names(matching_file)[names(matching_file) == "V1"] <- "GENE"
  names(BP_MF_Terms)[names(BP_MF_Terms) == "From"] <- "ID"
  names(BP_MF_Terms)[names(BP_MF_Terms) == "Gene.Names"] <- "GENENAMES"
  names(BP_MF_Terms)[names(BP_MF_Terms) == "Gene.Ontology..biological.process."] <- "GO:BP"
  names(BP_MF_Terms)[names(BP_MF_Terms) == "Gene.Ontology..molecular.function."] <- "GO:MF"
}

GO_lookup <- BP_MF_Terms %>%
  group_by(ID) %>%
  summarise(
    has_GO = any(
      (!is.na(`GO:BP`) & trimws(`GO:BP`) != "") |
      (!is.na(`GO:MF`) & trimws(`GO:MF`) != "")
    ),
    .groups = "drop"
  )

matching_file <- matching_file %>%
  left_join(GO_lookup, by = "ID") %>%
  filter(has_GO) %>%
  group_by(GENE) %>%
  slice_min(order_by = as.numeric(V11), n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(-has_GO)

matching_file <- unique(matching_file)
  
# Merge all matched BLAST rows with UniProt GO annotations
merged_df <- merge(matching_file, BP_MF_Terms, by = "ID") # genrates duplicates when matching
merged_df <- unique(merged_df)

# Helper function:  -------------------------------
# expands one GO column so that each GO term becomes its own row

expand_go_column_mem <- function(
  df,
  go_col,
  gene_col = "GENE",
  id_col = "ID",
  deduplicate = FALSE
) {
  empty_out <- data.frame(
    GENE = character(),
    ID   = character(),
    GO   = character(),
    TERM = character(),
    stringsAsFactors = FALSE
  )

  # Check required columns exist
  if (!all(c(go_col, gene_col, id_col) %in% names(df))) {
    stop("One or more specified columns do not exist in the data frame.")
  }

  go_text  <- as.character(df[[go_col]])
  gene_vec <- as.character(df[[gene_col]])
  id_vec   <- as.character(df[[id_col]])

  # Keep only usable rows up front
  keep <- !is.na(go_text) &
          trimws(go_text) != "" &
          !is.na(gene_vec) &
          !is.na(id_vec)

  if (!any(keep)) {
    return(empty_out)
  }

  go_text  <- go_text[keep]
  gene_vec <- gene_vec[keep]
  id_vec   <- id_vec[keep]

  # Clean up extracted GO term descriptions
  clean_term <- function(x) {
    x <- trimws(x)

    # Normalise curly quotes to straight quotes
    x <- gsub("[\u2018\u2019]", "'", x, perl = TRUE)
    x <- gsub("[\u201C\u201D]", "\"", x, perl = TRUE)

    # Remove surrounding quotes/backticks
    x <- gsub("^[[:space:]]*[\"'`]+", "", x, perl = TRUE)
    x <- gsub("[\"'`]+[[:space:]]*$", "", x, perl = TRUE)

    # Remove leading/trailing separators left over from splitting
    x <- gsub("^[[:space:]]*[,;:]+[[:space:]]*", "", x, perl = TRUE)
    x <- gsub("[[:space:]]*[,;:]+[[:space:]]*$", "", x, perl = TRUE)

    # Final trim
    x <- trimws(x)

    x
  }

  # Extract GO-term pairs from one cell
  extract_go_pairs <- function(x) {
    x <- as.character(x)
    if (is.na(x) || trimws(x) == "") return(NULL)

    # Normalise whitespace and quotes
    x <- gsub("[\r\n\t]+", " ", x, perl = TRUE)
    x <- gsub("\\s+", " ", x, perl = TRUE)
    x <- gsub("[\u2018\u2019]", "'", x, perl = TRUE)
    x <- gsub("[\u201C\u201D]", "\"", x, perl = TRUE)
    x <- trimws(x)

    # Split first on semicolons, then extract every GO match within each chunk
    chunks <- trimws(unlist(strsplit(x, ";", fixed = TRUE)))
    chunks <- chunks[chunks != ""]

    if (length(chunks) == 0L) return(NULL)

    go_ids   <- character()
    go_terms <- character()

    for (chunk in chunks) {
      # Match each "description [GO:0000000]" unit
      matches <- regmatches(
        chunk,
        gregexpr("[^\\[]*\\[GO:[0-9]{7}\\]", chunk, perl = TRUE)
      )[[1]]

      if (length(matches) == 0L) next

      for (m in matches) {
        this_go <- sub(".*\\[(GO:[0-9]{7})\\].*", "\\1", m, perl = TRUE)
        this_term <- sub("\\s*\\[GO:[0-9]{7}\\].*", "", m, perl = TRUE)
        this_term <- clean_term(this_term)

        go_ids   <- c(go_ids, this_go)
        go_terms <- c(go_terms, if (nzchar(this_term)) this_term else NA_character_)
      }
    }

    if (length(go_ids) == 0L) return(NULL)

    data.frame(
      GO   = go_ids,
      TERM = go_terms,
      stringsAsFactors = FALSE
    )
  }

  # First pass: extract all GO-term pairs and count output rows
  parsed_list <- lapply(go_text, extract_go_pairs)
  counts <- vapply(parsed_list, function(x) {
    if (is.null(x)) 0L else nrow(x)
  }, integer(1))

  total_rows <- sum(counts)

  if (total_rows == 0L) {
    return(empty_out)
  }

  # Preallocate final vectors once
  GENE <- character(total_rows)
  ID   <- character(total_rows)
  GO   <- character(total_rows)
  TERM <- character(total_rows)

  pos <- 1L

  # Second pass: fill preallocated vectors
  for (i in seq_along(parsed_list)) {
    parsed <- parsed_list[[i]]
    if (is.null(parsed) || nrow(parsed) == 0L) next

    n_terms <- nrow(parsed)
    idx <- pos:(pos + n_terms - 1L)

    GENE[idx] <- gene_vec[i]
    ID[idx]   <- id_vec[i]
    GO[idx]   <- parsed$GO
    TERM[idx] <- parsed$TERM

    pos <- pos + n_terms
  }

  out <- data.frame(
    GENE = GENE,
    ID   = ID,
    GO   = GO,
    TERM = TERM,
    stringsAsFactors = FALSE
  )

  rownames(out) <- NULL

  if (deduplicate) {
    out <- unique(out)
    rownames(out) <- NULL
  }

  out
}


# Expand BP and MF properly -------------------------------

# remove files for space
rm(BP_MF_Terms,
   matching_file)

# Split merged_df into thirds  -------------------------------

n <- nrow(merged_df)

idx1 <- 1:floor(n / 3)
idx2 <- (floor(n / 3) + 1):floor(2 * n / 3)
idx3 <- (floor(2 * n / 3) + 1):n

part1 <- merged_df[idx1, , drop = FALSE]
part2 <- merged_df[idx2, , drop = FALSE]
part3 <- merged_df[idx3, , drop = FALSE]

rm(merged_df,
   idx1,
   idx2,
   idx3,
   n)

# Expand GO:BP in chunks  -------------------------------

BP_long_1 <- expand_go_column_mem(part1, "GO:BP", deduplicate = FALSE)
BP_long_2 <- expand_go_column_mem(part2, "GO:BP", deduplicate = FALSE)
BP_long_3 <- expand_go_column_mem(part3, "GO:BP", deduplicate = FALSE)

BP_long <- unique(rbind(BP_long_1, BP_long_2, BP_long_3))
rm(BP_long_1, BP_long_2, BP_long_3)
BP_long <- unique(BP_long) # no duplicated detected

# Expand GO:MF in chunks -------------------------------

MF_long_1 <- expand_go_column_mem(part1, "GO:MF", deduplicate = FALSE)
MF_long_2 <- expand_go_column_mem(part2, "GO:MF", deduplicate = FALSE)
MF_long_3 <- expand_go_column_mem(part3, "GO:MF", deduplicate = FALSE)

MF_long <- unique(rbind(MF_long_1, MF_long_2, MF_long_3))
rm(MF_long_1, MF_long_2, MF_long_3)
MF_long <- unique(MF_long) # no duplicates detected

# Only make downstream objects after BP_long / MF_long are finished
GENES_BP <- unique(BP_long[, c("GO", "GENE")])
GENES_MF <- unique(MF_long[, c("GO", "GENE")])

TERMs_BP <- unique(BP_long[, c("GO", "TERM")])
TERMs_MF <- unique(MF_long[, c("GO", "TERM")])

# remove redundant dfs now
rm(part1,
   part2,
   part3,
   BP_long,
   MF_long)

save.image(file = "C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_3/data/BP_MF_long_extractions.Rdata")

# load when needed
  # load("C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_3/data/BP_MF_long_extractions.Rdata")

```

## Load DE data

```{r data}
# Import DEA results
getwd()
DE_results <- read.table("chapter_3/results/Multi_Stressor_all_Differential_Expression_Analysis_results.csv",
                           header = T, 
                           sep = ',', 
                           stringsAsFactors = TRUE)

## Perform naming transformation to make gene IDs matching
DE_results$gene <- gsub("\\.TU", "", DE_results$gene) # Removing "TU" part of names
DE_results$gene <- gsub("path", "mrna", DE_results$gene) # Replacing "path" -> "mrna"
```

## Generate DEGs lists and ranked lists

```{r DEGs_and_Ranked_lists}
# Diesel_only
diesel_only_geneList = DE_results[,2]
names(diesel_only_geneList) = as.character(DE_results[,1])
diesel_only_geneList = sort(diesel_only_geneList, decreasing = TRUE)

# Salinity only
salinity_only_geneList = DE_results[,4]
names(salinity_only_geneList) = as.character(DE_results[,1])
salinity_only_geneList = sort(salinity_only_geneList, decreasing = TRUE)

# Interactive
interactive_only_geneList = DE_results[,10]
names(interactive_only_geneList) = as.character(DE_results[,1])
interactive_only_geneList = sort(interactive_only_geneList, decreasing = TRUE)

# Diesel Added
diesel_added_geneList = DE_results[,6]
names(diesel_added_geneList) = as.character(DE_results[,1])
diesel_added_geneList = sort(diesel_added_geneList, decreasing = TRUE)

# Salinity added
salinity_added_geneList = DE_results[,8]
names(salinity_added_geneList) = as.character(DE_results[,1])
salinity_added_geneList = sort(salinity_added_geneList, decreasing = TRUE)

# Additive
combined_geneList = DE_results[,12]
names(combined_geneList) = as.character(DE_results[,1])
combined_geneList = sort(combined_geneList, decreasing = TRUE)

# Generate DEGs lists for each contrast ----
# diesel only
DEGs <- subset(DE_results[, c("gene", "padj_diesel_only_wald")])
DEGs <- DEGs[(DEGs$padj_diesel_only_wald < 0.05),]
DEGs <- DEGs[complete.cases(DEGs),]
diesel_only_Wald_DEGs <- as.vector(DEGs[,1])

# salinity only
DEGs <- subset(DE_results[, c("gene", "padj_salinity_only_wald")])
DEGs <- DEGs[(DEGs$padj_salinity_only_wald < 0.05),]
DEGs <- DEGs[complete.cases(DEGs),]
salinity_only_Wald_DEGs <- as.vector(DEGs[,1])

# Interactive only
DEGs <- subset(DE_results[, c("gene", "padj_interactive_only_wald")])
DEGs <- DEGs[(DEGs$padj_interactive_only_wald < 0.05),]
DEGs <- DEGs[complete.cases(DEGs),]
interactive_only_Wald_DEGs <- as.vector(DEGs[,1])

# Diesel Added
DEGs <- subset(DE_results[, c("gene", "padj_diesel_added_wald")])
DEGs <- DEGs[(DEGs$padj_diesel_added_wald < 0.05),]
DEGs <- DEGs[complete.cases(DEGs),]
diesel_added_Wald_DEGs <- as.vector(DEGs[,1])

# salinity Added
DEGs <- subset(DE_results[, c("gene", "padj_salinity_added_wald")])
DEGs <- DEGs[(DEGs$padj_salinity_added_wald < 0.05),]
DEGs <- DEGs[complete.cases(DEGs),]
salinity_added_Wald_DEGs <- as.vector(DEGs[,1])

# Additive
DEGs <- subset(DE_results[, c("gene", "padj_combined_wald")])
DEGs <- DEGs[(DEGs$padj_combined_wald < 0.05),]
DEGs <- DEGs[complete.cases(DEGs),]
combined_Wald_DEGs <- as.vector(DEGs[,1])
```

# Overrepresentation Analysis

```{r ORA}
## Looping script to perform all ORAs ----
run_ORA_loop <- function(groups, out_dir = "chapter_3/results/model_outputs", assign_to_env = TRUE, env = .GlobalEnv) {
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
  res_list <- list()
  
  safe_enricher <- function(degs, TERM2GENE, TERM2NAME, label) {
    tryCatch({
      set.seed(100)
      obj <- enricher(degs, TERM2GENE = TERM2GENE, TERM2NAME = TERM2NAME, minGSSize = 1)
      res <- obj@result
      if (!is.null(res) && nrow(res) > 0) res$GO <- label
      res
    }, error = function(e) {
      warning(sprintf("ORA failed for '%s': %s", label, conditionMessage(e)))
      NULL
    })
  }
  safe_convert <- function(x) tryCatch(convertFractionToNumeric(x), error = function(e) NA_real_)
  
  for (g in groups) {
    tryCatch({
      deg_var <- paste0(g, "_DEGs")
      if (!exists(deg_var)) {
        warning(sprintf("Skipping '%s' — object '%s' not found.", g, deg_var))
        res_list[[g]] <- NULL
        next
      }
      degs <- get(deg_var, inherits = TRUE)
      
      bp_res <- safe_enricher(degs, GENES_BP, TERMs_BP, "biological_processes")
      mf_res <- safe_enricher(degs, GENES_MF, TERMs_MF, "molecular_functions")
      pieces <- Filter(Negate(is.null), list(bp_res, mf_res))
      if (length(pieces) == 0) {
        warning(sprintf("No ORA results produced for '%s'.", g))
        res_list[[g]] <- NULL
        next
      }
      
      res <- tryCatch(do.call(rbind, pieces), error = function(e) { warning(sprintf("Combine fail '%s': %s", g, e$message)); NULL })
      if (is.null(res) || nrow(res) == 0) { res_list[[g]] <- NULL; next }
      
      if ("p.adjust" %in% names(res)) res <- res[!is.na(res$p.adjust) & res$p.adjust < 0.05, , drop = FALSE]
      if (nrow(res) == 0) { warning(sprintf("No significant terms for '%s'.", g)); res_list[[g]] <- res; next }
      
      if ("GeneRatio" %in% names(res)) res$GeneRatio <- sapply(res$GeneRatio, safe_convert)
      if ("BgRatio"   %in% names(res)) res$BgRatio   <- sapply(res$BgRatio,   safe_convert)
      
      outfile <- file.path(out_dir, sprintf("%s_ORA_results.csv", g))
      tryCatch(write.csv(res, outfile, row.names = FALSE),
               error = function(e) warning(sprintf("Write fail '%s': %s", g, e$message)))
      
      res_list[[g]] <- res
      
      if (assign_to_env) {
        obj_name <- sprintf("%s_ORA_results", g)
        assign(obj_name, res, envir = env)
      }
    }, error = function(e) {
      warning(sprintf("Unexpected error in group '%s': %s", g, conditionMessage(e)))
      res_list[[g]] <- NULL
    })
  }
  
  return(res_list)
}
groups <- c("diesel_only_Wald",
            "salinity_only_Wald",
            "interactive_only_Wald",
            "diesel_added_Wald",
            "salinity_added_Wald",
            "combined_Wald") 

res <- run_ORA_loop(groups, assign_to_env = TRUE)
```

# Gene Set Enrichment Analysis

```{r GSEA}
# Make loop
run_GSEA_loop <- function(groups,
                          out_dir = "chapter_3/results/model_outputs",
                          p_cutoff = 0.05,
                          assign_to_env = TRUE,
                          env = .GlobalEnv) {
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
  res_list <- list()
  
  # Safe GSEA runner -> returns a list(GSEA_obj = <enrichResult>, df = <data.frame>|NULL)
  safe_gsea <- function(geneList, TERM2GENE, TERM2NAME) {
    tryCatch({
      set.seed(100)
      g <- GSEA(geneList, TERM2GENE = TERM2GENE, TERM2NAME = TERM2NAME, pvalueCutoff = p_cutoff)
      df <- g@result
      list(GSEA_obj = g, df = df)
    }, error = function(e) {
      warning(sprintf("GSEA failed: %s", conditionMessage(e)))
      list(GSEA_obj = NULL, df = NULL)
    })
  }
  
  for (g in groups) {
    tryCatch({
      gl_var <- paste0(g, "_geneList")
      if (!exists(gl_var)) {
        warning(sprintf("Skipping '%s' — object '%s' not found.", g, gl_var))
        res_list[[g]] <- NULL
        next
      }
      geneList <- get(gl_var, inherits = TRUE)
      
      # Run BP and MF
      bp <- safe_gsea(geneList, GENES_BP, TERMs_BP)
      mf <- safe_gsea(geneList, GENES_MF, TERMs_MF)
      
      bp_df <- bp$df
      if (!is.null(bp_df) && nrow(bp_df) > 0) bp_df$GO <- "biological_processes"
      mf_df <- mf$df
      if (!is.null(mf_df) && nrow(mf_df) > 0) mf_df$GO <- "molecular_functions"
      
      # Merge results (skip NULLs)
      pieces <- Filter(function(x) !is.null(x) && nrow(x) > 0, list(bp_df, mf_df))
      if (length(pieces) == 0) {
        warning(sprintf("No GSEA results produced for '%s'.", g))
        res_list[[g]] <- NULL
        next
      }
      merged <- tryCatch({
        do.call(rbind, pieces)
      }, error = function(e) {
        warning(sprintf("Failed to combine GSEA results for '%s': %s", g, conditionMessage(e)))
        NULL
      })
      
      if (is.null(merged) || nrow(merged) == 0) {
        warning(sprintf("Empty combined GSEA results for '%s'.", g))
        res_list[[g]] <- NULL
        next
      }
      
      # Filter significant results
      if ("p.adjust" %in% names(merged)) {
        merged <- merged[!is.na(merged$p.adjust) & merged$p.adjust < p_cutoff, , drop = FALSE]
      } else {
        warning(sprintf("'p.adjust' column missing for '%s'; skipping significance filter.", g))
      }
      
      # Package like your example: list(BP, MF, Results)
      group_list <- list(
        BP = bp$GSEA_obj,
        MF = mf$GSEA_obj,
        Results = merged
      )
      
      # Write CSV
      outfile <- file.path(out_dir, sprintf("%s_GSEA_results.csv", g))
      tryCatch({
        write.csv(group_list$Results, outfile, row.names = FALSE)
      }, error = function(e) {
        warning(sprintf("Failed to write CSV for '%s' to '%s': %s", g, outfile, conditionMessage(e)))
      })
      
      res_list[[g]] <- group_list
      
      # Optional: assign to environment for easy inspection
      if (assign_to_env) {
        assign(sprintf("%s_GSEA", g), group_list, envir = env)
        assign(sprintf("%s_GSEA_results", g), group_list$Results, envir = env)
      }
    }, error = function(e) {
      warning(sprintf("Unexpected error in group '%s': %s", g, conditionMessage(e)))
      res_list[[g]] <- NULL
    })
  }
  
  return(res_list)  # visible return
}

# Run loop
groups <- c("diesel_only","salinity_only","interactive_only","diesel_added","salinity_added","combined")
set.seed(100)
gsea_out <- run_GSEA_loop(groups, p_cutoff = 0.05, assign_to_env = TRUE)

# Save image
save.image(file = "C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 3/RData/ORA_and_GSEA_Wald_completed.RData")
```

# Combine and Save results files ----

## ORA

```{r combined_save_ORA}
# results files names:
  # diesel_only_Wald_ORA_results
  # diesel_added_Wald_ORA_results
  # salinity_only_Wald_ORA_results
  # salinity_added_Wald_ORA_results
  # combined_Wald_ORA_results
  # interactive_only_Wald_ORA_results

# Add contrast column so differentiate between contrasts in the same file:
diesel_only_Wald_ORA_results$contrast <- "diesel_only"
diesel_added_Wald_ORA_results$contrast <- "diesel_added"
salinity_only_Wald_ORA_results$contrast <- "salinity_only"
salinity_added_Wald_ORA_results$contrast <- "salinity_added"
combined_Wald_ORA_results$contrast <- "combined"
interactive_only_Wald_ORA_results$contrast <- "interactive"

# merge
all_ORA_results <- bind_rows(
  diesel_only_Wald_ORA_results,
  diesel_added_Wald_ORA_results,
  salinity_only_Wald_ORA_results,
  salinity_added_Wald_ORA_results,
  combined_Wald_ORA_results,
  interactive_only_Wald_ORA_results
)

# Write.csv
write.csv(all_ORA_results, "chapter_3/results/model_outputs/MS_all_ORA_results.csv", row.names = F)
```

## GSEA

```{r combined_save_GSEA}
# Adding contrast column
# diesel_only_GSEA_results$contrast <- "diesel_only" # No Results
salinity_only_GSEA_results$contrast <- "salinity_only"
interactive_only_GSEA_results$contrast <- "interactive"
diesel_added_GSEA_results$contrast <- "diesel_added"
salinity_added_GSEA_results$contrast <- "salinity_added"
combined_GSEA_results$contrast <- "additive"

# Combine
all_GSEA_results <- bind_rows(#diesel_only_GSEA_results,
                         salinity_only_GSEA_results,
                         #interactive_only_GSEA_results,
                         diesel_added_GSEA_results,
                         salinity_added_GSEA_results,
                         combined_GSEA_results
)

# Save
write.csv(all_GSEA_results, "chapter_3/results/model_outputs/MS_all_GSEA_results.csv", row.names = F)
```

# Plots

## ORA plots

### Attach LFC data to GO data

```{r ORA_plots}
# only significant results
ORA_results <- subset(all_ORA_results[all_ORA_results$p.adjust < 0.05,])
ORA_results <- ORA_results[!apply(is.na(ORA_results), 1, all), ]

# empty df to add data to:
ORA_plotting_data <- tibble(
  term = character(),
  gene = character(),
  lfc = numeric(),
  contrast = factor()
)

# Diesel only
{
data_temp <- subset(DE_results[, c("gene", "padj_diesel_only_wald","LFC_diesel_only_wald")])

  # change these if your column names differ
  lfc_col        <- "LFC_diesel_only_wald"  # shrunken LFC column in DEA file
  gene_col       <- "gene"            
  
  # ORA columns 
  ora_df         <- subset(ORA_results[ORA_results$contrast == "diesel_only",]) #                     # Change this to new contrast
  ora_term_col   <- if ("Description" %in% names(ora_df)) "Description" else "ID"
  ora_padj_col   <- if ("p.adjust"   %in% names(ora_df)) "p.adjust"   else "pvalue"  # fallback
  ora_genes_col  <- "geneID"         # nested gene list in each row
  

  lfc_map <- data_temp %>%
    transmute(gene = .data[[gene_col]],
              lfc  = .data[[lfc_col]]) %>%
    distinct()
  

  attach_lfc <- function(df, term_col, genes_col, lfc_tbl,
                         split_pattern = "[,;/\\|\\s]+") {
    out <- df %>%
      transmute(term  = .data[[term_col]],
                genes = .data[[genes_col]]) %>%
      mutate(gene = str_split(genes, split_pattern)) %>%
      select(-genes) %>%
      unnest(gene) %>%
      mutate(gene = str_trim(gene)) %>%
      filter(!is.na(gene), gene != "") %>%
      distinct(term, gene) %>%  
      left_join(lfc_tbl, by = "gene")
    out
  }
  

  ora_long  <- attach_lfc(ora_df,  ora_term_col,  ora_genes_col,  lfc_map)

  ora_terms  <- ora_df  %>% arrange(.data[[ora_padj_col]]) %>% pull(!!ora_term_col)
  
  ora_plotdf  <- ora_long  %>% filter(term %in% ora_terms) %>% mutate(contrast = "diesel_only")   # Change contrast name here

  
  # order terms inside each panel by median LFC (you can change to FDR ordering)
  ora_levels  <- ora_plotdf  %>% group_by(term)  %>% summarise(med = median(lfc, na.rm = TRUE))  %>%
    arrange(med) %>% pull(term)

  
  ora_plotdf$term  <- factor(ora_plotdf$term,  levels = ora_levels)

  ORA_plotting_data <- bind_rows(ora_plotdf, ORA_plotting_data)
}

# salinity only
{
data_temp <- subset(DE_results[, c("gene", "padj_salinity_only_wald","LFC_salinity_only_wald")])

  # change these if your column names differ
  lfc_col        <- "LFC_salinity_only_wald"  # shrunken LFC column in DEA file
  gene_col       <- "gene"            
  
  # ORA columns 
  ora_df         <- subset(ORA_results[ORA_results$contrast == "salinity_only",]) #                     # Change this to new contrast
  ora_term_col   <- if ("Description" %in% names(ora_df)) "Description" else "ID"
  ora_padj_col   <- if ("p.adjust"   %in% names(ora_df)) "p.adjust"   else "pvalue"  # fallback
  ora_genes_col  <- "geneID"         # nested gene list in each row
  

  lfc_map <- data_temp %>%
    transmute(gene = .data[[gene_col]],
              lfc  = .data[[lfc_col]]) %>%
    distinct()
  

  attach_lfc <- function(df, term_col, genes_col, lfc_tbl,
                         split_pattern = "[,;/\\|\\s]+") {
    out <- df %>%
      transmute(term  = .data[[term_col]],
                genes = .data[[genes_col]]) %>%
      mutate(gene = str_split(genes, split_pattern)) %>%
      select(-genes) %>%
      unnest(gene) %>%
      mutate(gene = str_trim(gene)) %>%
      filter(!is.na(gene), gene != "") %>%
      distinct(term, gene) %>%  
      left_join(lfc_tbl, by = "gene")
    out
  }
  

  ora_long  <- attach_lfc(ora_df,  ora_term_col,  ora_genes_col,  lfc_map)

  ora_terms  <- ora_df  %>% arrange(.data[[ora_padj_col]]) %>% pull(!!ora_term_col)
  
  ora_plotdf  <- ora_long  %>% filter(term %in% ora_terms) %>% mutate(contrast = "salinity_only")   # Change contrast name here

  
  # order terms inside each panel by median LFC (you can change to FDR ordering)
  ora_levels  <- ora_plotdf  %>% group_by(term)  %>% summarise(med = median(lfc, na.rm = TRUE))  %>%
    arrange(med) %>% pull(term)

  
  ora_plotdf$term  <- factor(ora_plotdf$term,  levels = ora_levels)

  ORA_plotting_data <- bind_rows(ORA_plotting_data, ora_plotdf)
}

# Diesel added
{
data_temp <- subset(DE_results[, c("gene", "padj_diesel_added_wald","LFC_diesel_added_wald")])

  # change these if your column names differ
  lfc_col        <- "LFC_diesel_added_wald"  # shrunken LFC column in DEA file
  gene_col       <- "gene"            
  
  # ORA columns 
  ora_df         <- subset(ORA_results[ORA_results$contrast == "diesel_added",]) #                     # Change this to new contrast
  ora_term_col   <- if ("Description" %in% names(ora_df)) "Description" else "ID"
  ora_padj_col   <- if ("p.adjust"   %in% names(ora_df)) "p.adjust"   else "pvalue"  # fallback
  ora_genes_col  <- "geneID"         # nested gene list in each row
  

  lfc_map <- data_temp %>%
    transmute(gene = .data[[gene_col]],
              lfc  = .data[[lfc_col]]) %>%
    distinct()
  

  attach_lfc <- function(df, term_col, genes_col, lfc_tbl,
                         split_pattern = "[,;/\\|\\s]+") {
    out <- df %>%
      transmute(term  = .data[[term_col]],
                genes = .data[[genes_col]]) %>%
      mutate(gene = str_split(genes, split_pattern)) %>%
      select(-genes) %>%
      unnest(gene) %>%
      mutate(gene = str_trim(gene)) %>%
      filter(!is.na(gene), gene != "") %>%
      distinct(term, gene) %>%  
      left_join(lfc_tbl, by = "gene")
    out
  }
  

  ora_long  <- attach_lfc(ora_df,  ora_term_col,  ora_genes_col,  lfc_map)

  ora_terms  <- ora_df  %>% arrange(.data[[ora_padj_col]]) %>% pull(!!ora_term_col)
  
  ora_plotdf  <- ora_long  %>% filter(term %in% ora_terms) %>% mutate(contrast = "diesel_added")   # Change contrast name here

  
  # order terms inside each panel by median LFC (you can change to FDR ordering)
  ora_levels  <- ora_plotdf  %>% group_by(term)  %>% summarise(med = median(lfc, na.rm = TRUE))  %>%
    arrange(med) %>% pull(term)

  
  ora_plotdf$term  <- factor(ora_plotdf$term,  levels = ora_levels)

  ORA_plotting_data <- bind_rows(ora_plotdf, ORA_plotting_data)
}

# salinity added
{
data_temp <- subset(DE_results[, c("gene", "padj_salinity_added_wald","LFC_salinity_added_wald")])

  # change these if your column names differ
  lfc_col        <- "LFC_salinity_added_wald"  # shrunken LFC column in DEA file
  gene_col       <- "gene"            
  
 # ORA columns
ora_df         <- subset(ORA_results[ORA_results$contrast == "salinity_added",])
ora_term_col   <- if ("Description" %in% names(ora_df)) "Description" else "ID"
ora_padj_col   <- if ("p.adjust"   %in% names(ora_df)) "p.adjust"   else "pvalue"  # fallback
ora_genes_col  <- "geneID"         # nested gene list in each row

  
lfc_map <- data_temp %>%
  transmute(gene = .data[[gene_col]],
            lfc  = .data[[lfc_col]]) %>%
  distinct()


attach_lfc <- function(df, term_col, genes_col, lfc_tbl,
                       split_pattern = "[,;/\\|\\s]+") {
  out <- df %>%
    transmute(term  = .data[[term_col]],
              genes = .data[[genes_col]]) %>%
    mutate(gene = str_split(genes, split_pattern)) %>%
    select(-genes) %>%
    unnest(gene) %>%
    mutate(gene = str_trim(gene)) %>%
    filter(!is.na(gene), gene != "") %>%
    distinct(term, gene) %>%                      # drop duplicates within a term
    left_join(lfc_tbl, by = "gene")
  out
}


ora_long  <- attach_lfc(ora_df,  ora_term_col,  ora_genes_col,  lfc_map)
  

  ora_long  <- attach_lfc(ora_df,  ora_term_col,  ora_genes_col,  lfc_map)

  ora_terms  <- ora_df  %>% arrange(.data[[ora_padj_col]]) %>% pull(!!ora_term_col)
  
  ora_plotdf  <- ora_long  %>% filter(term %in% ora_terms) %>% mutate(contrast = "salinity_added")   # Change contrast name here

  
  # order terms inside each panel by median LFC (you can change to FDR ordering)
  ora_levels  <- ora_plotdf  %>% group_by(term)  %>% summarise(med = median(lfc, na.rm = TRUE))  %>%
    arrange(med) %>% pull(term)

  
  ora_plotdf$term  <- factor(ora_plotdf$term,  levels = ora_levels)

  ORA_plotting_data <- bind_rows(ORA_plotting_data, ora_plotdf)
}

# combined
{
data_temp <- subset(DE_results[, c("gene", "padj_combined_wald","LFC_combined_wald")])

  # change these if your column names differ
  lfc_col        <- "LFC_combined_wald"  # shrunken LFC column in DEA file
  gene_col       <- "gene"            
  
 # ORA columns
ora_df         <- subset(ORA_results[ORA_results$contrast == "combined",])
ora_term_col   <- if ("Description" %in% names(ora_df)) "Description" else "ID"
ora_padj_col   <- if ("p.adjust"   %in% names(ora_df)) "p.adjust"   else "pvalue"  # fallback
ora_genes_col  <- "geneID"         # nested gene list in each row

  
lfc_map <- data_temp %>%
  transmute(gene = .data[[gene_col]],
            lfc  = .data[[lfc_col]]) %>%
  distinct()


attach_lfc <- function(df, term_col, genes_col, lfc_tbl,
                       split_pattern = "[,;/\\|\\s]+") {
  out <- df %>%
    transmute(term  = .data[[term_col]],
              genes = .data[[genes_col]]) %>%
    mutate(gene = str_split(genes, split_pattern)) %>%
    select(-genes) %>%
    unnest(gene) %>%
    mutate(gene = str_trim(gene)) %>%
    filter(!is.na(gene), gene != "") %>%
    distinct(term, gene) %>%                      # drop duplicates within a term
    left_join(lfc_tbl, by = "gene")
  out
}


ora_long  <- attach_lfc(ora_df,  ora_term_col,  ora_genes_col,  lfc_map)
  

  ora_long  <- attach_lfc(ora_df,  ora_term_col,  ora_genes_col,  lfc_map)

  ora_terms  <- ora_df  %>% arrange(.data[[ora_padj_col]]) %>% pull(!!ora_term_col)
  
  ora_plotdf  <- ora_long  %>% filter(term %in% ora_terms) %>% mutate(contrast = "combined")   # Change contrast name here

  
  # order terms inside each panel by median LFC (you can change to FDR ordering)
  ora_levels  <- ora_plotdf  %>% group_by(term)  %>% summarise(med = median(lfc, na.rm = TRUE))  %>%
    arrange(med) %>% pull(term)

  
  ora_plotdf$term  <- factor(ora_plotdf$term,  levels = ora_levels)

  ORA_plotting_data <- bind_rows(ORA_plotting_data, ora_plotdf)
}

# interactive
{
data_temp <- subset(DE_results[, c("gene", "padj_interactive_only_wald","LFC_interactive_only_wald")])

  # change these if your column names differ
  lfc_col        <- "LFC_interactive_only_wald"  # shrunken LFC column in DEA file
  gene_col       <- "gene"            
  
 # ORA columns
ora_df         <- subset(ORA_results[ORA_results$contrast == "interactive",])
ora_term_col   <- if ("Description" %in% names(ora_df)) "Description" else "ID"
ora_padj_col   <- if ("p.adjust"   %in% names(ora_df)) "p.adjust"   else "pvalue"  # fallback
ora_genes_col  <- "geneID"         # nested gene list in each row

  
lfc_map <- data_temp %>%
  transmute(gene = .data[[gene_col]],
            lfc  = .data[[lfc_col]]) %>%
  distinct()


attach_lfc <- function(df, term_col, genes_col, lfc_tbl,
                       split_pattern = "[,;/\\|\\s]+") {
  out <- df %>%
    transmute(term  = .data[[term_col]],
              genes = .data[[genes_col]]) %>%
    mutate(gene = str_split(genes, split_pattern)) %>%
    select(-genes) %>%
    unnest(gene) %>%
    mutate(gene = str_trim(gene)) %>%
    filter(!is.na(gene), gene != "") %>%
    distinct(term, gene) %>%                      # drop duplicates within a term
    left_join(lfc_tbl, by = "gene")
  out
}


ora_long  <- attach_lfc(ora_df,  ora_term_col,  ora_genes_col,  lfc_map)
  

  ora_long  <- attach_lfc(ora_df,  ora_term_col,  ora_genes_col,  lfc_map)

  ora_terms  <- ora_df  %>% arrange(.data[[ora_padj_col]]) %>% pull(!!ora_term_col)
  
  ora_plotdf  <- ora_long  %>% filter(term %in% ora_terms) %>% mutate(contrast = "interactive")   # Change contrast name here

  
  # order terms inside each panel by median LFC (you can change to FDR ordering)
  ora_levels  <- ora_plotdf  %>% group_by(term)  %>% summarise(med = median(lfc, na.rm = TRUE))  %>%
    arrange(med) %>% pull(term)

  
  ora_plotdf$term  <- factor(ora_plotdf$term,  levels = ora_levels)

  ORA_plotting_data <- bind_rows(ORA_plotting_data, ora_plotdf)
}
```

### Plot ORA

```{r image_load_ORA}
load("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 3/RData/MS_Enrichment_Analyses.RData")
```

```{r plot_ORA_results}
# 
library(dplyr)
library(ggplot2)
library(ggridges)
library(stringr)
library(forcats)
library(scales)

# Options
plot_df <- ORA_plotting_data %>%
  group_by(term) %>%
  summarise(
    n_contrasts = n_distinct(contrast),
    mean_abs_lfc = mean(abs(lfc), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(n_contrasts), desc(mean_abs_lfc), term) %>%
  pull(term) -> term_levels

plot_df <- ORA_plotting_data %>%
  mutate(
    term = factor(term, levels = rev(term_levels)),
    contrast = factor(
      contrast,
      levels = c(
        "diesel_only",
        "diesel_added",
        "salinity_only",
        "salinity_added",
        "combined",
        "interactive"
      )
    )
  )

# code to modify:
  p_boxes <- ggplot(plot_df %>% filter(!is.na(lfc)),
                    aes(x = lfc, 
                        y = term, 
                        #fill = contrast, 
                        #color = contrast
                        )) +
    geom_boxplot(outlier.size = 0.6, width = 0.6, 
                 color = "grey30", 
                 fill = "grey88"
                 ) +
    geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.6, col = "grey") +
    labs(x = "Shrunken log2 fold-change (LFC)", y = NULL) +
    facet_grid(
      cols = vars(contrast),
      scales = "free",
      labeller = labeller(
        contrast = c(
          diesel_only = "Diesel",
          diesel_added = "+ Diesel",
          salinity_only = "Salinity",
          salinity_added = "+ Salinity",
          additive = "Additive",
          interactive = "Interactive"
        )
      )
    ) +
    #xlim(-25,25) +
    theme_bw(base_size = 8) +
    theme(strip.background = element_rect(fill = "white"),
          axis.text.y = element_text(size = 6))

p_boxes

# Save plot
ggsave(
  "C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_3/figures/main figures/ORA_boxplot.png",
  p_boxes,
  width = 246,
  height = 159,
  units = "mm",
  dpi = 300
)
```

## GSEA plots

## attach lfc data

```{r GSEA_lfc_data}
# only significant results
GSEA_results <- subset(all_GSEA_results[all_GSEA_results$p.adjust < 0.05,])
GSEA_results <- GSEA_results[!apply(is.na(GSEA_results), 1, all), ]

# empty df to add data to:
GSEA_plotting_data <- tibble(
  term = character(),
  gene = character(),
  lfc = numeric(),
  contrast = factor()
)

# No diesel only results, skip...

# salinity only
{
data_temp <- subset(DE_results[, c("gene", "padj_salinity_only_wald","LFC_salinity_only_wald")])

  # change these if your column names differ
  lfc_col        <- "LFC_salinity_only_wald"  # shrunken LFC column in DEA file
  gene_col       <- "gene"            
  
  # GSEA columns 
  GSEA_df         <- subset(GSEA_results[GSEA_results$contrast == "salinity_only",]) #              
  GSEA_term_col   <- if ("Description" %in% names(GSEA_df)) "Description" else "ID"
  GSEA_padj_col   <- if ("p.adjust"   %in% names(GSEA_df)) "p.adjust"   else "pvalue"  # fallback
  GSEA_genes_col <- "core_enrichment"  # nested gene list (leading edge)
  

  lfc_map <- data_temp %>%
    transmute(gene = .data[[gene_col]],
              lfc  = .data[[lfc_col]]) %>%
    distinct()
  

  attach_lfc <- function(df, term_col, genes_col, lfc_tbl,
                         split_pattern = "[,;/\\|\\s]+") {
    out <- df %>%
      transmute(term  = .data[[term_col]],
                genes = .data[[genes_col]]) %>%
      mutate(gene = str_split(genes, split_pattern)) %>%
      select(-genes) %>%
      unnest(gene) %>%
      mutate(gene = str_trim(gene)) %>%
      filter(!is.na(gene), gene != "") %>%
      distinct(term, gene) %>%  
      left_join(lfc_tbl, by = "gene")
    out
  }
  

  GSEA_long  <- attach_lfc(GSEA_df,  GSEA_term_col,  GSEA_genes_col,  lfc_map)

  GSEA_terms  <- GSEA_df  %>% arrange(.data[[GSEA_padj_col]]) %>% pull(!!GSEA_term_col)
  
  GSEA_plotdf  <- GSEA_long  %>% filter(term %in% GSEA_terms) %>% mutate(contrast = "salinity_only")   # Change contrast name here

  
  # order terms inside each panel by median LFC (you can change to FDR ordering)
  GSEA_levels  <- GSEA_plotdf  %>% group_by(term)  %>% summarise(med = median(lfc, na.rm = TRUE))  %>%
    arrange(med) %>% pull(term)

  
  GSEA_plotdf$term  <- factor(GSEA_plotdf$term,  levels = GSEA_levels)

  GSEA_plotting_data <- bind_rows(GSEA_plotting_data, GSEA_plotdf)
}

# Diesel added
{
data_temp <- subset(DE_results[, c("gene", "padj_diesel_added_wald","LFC_diesel_added_wald")])

  # change these if your column names differ
  lfc_col        <- "LFC_diesel_added_wald"  # shrunken LFC column in DEA file
  gene_col       <- "gene"            
  
  # GSEA columns 
  GSEA_df         <- subset(GSEA_results[GSEA_results$contrast == "diesel_added",]) #                     # Change this to new contrast
  GSEA_term_col   <- if ("Description" %in% names(GSEA_df)) "Description" else "ID"
  GSEA_padj_col   <- if ("p.adjust"   %in% names(GSEA_df)) "p.adjust"   else "pvalue"  # fallback
  GSEA_genes_col <- "core_enrichment"  # nested gene list (leading edge)
  

  lfc_map <- data_temp %>%
    transmute(gene = .data[[gene_col]],
              lfc  = .data[[lfc_col]]) %>%
    distinct()
  

  attach_lfc <- function(df, term_col, genes_col, lfc_tbl,
                         split_pattern = "[,;/\\|\\s]+") {
    out <- df %>%
      transmute(term  = .data[[term_col]],
                genes = .data[[genes_col]]) %>%
      mutate(gene = str_split(genes, split_pattern)) %>%
      select(-genes) %>%
      unnest(gene) %>%
      mutate(gene = str_trim(gene)) %>%
      filter(!is.na(gene), gene != "") %>%
      distinct(term, gene) %>%  
      left_join(lfc_tbl, by = "gene")
    out
  }
  

  GSEA_long  <- attach_lfc(GSEA_df,  GSEA_term_col,  GSEA_genes_col,  lfc_map)

  GSEA_terms  <- GSEA_df  %>% arrange(.data[[GSEA_padj_col]]) %>% pull(!!GSEA_term_col)
  
  GSEA_plotdf  <- GSEA_long  %>% filter(term %in% GSEA_terms) %>% mutate(contrast = "diesel_added")   # Change contrast name here

  
  # order terms inside each panel by median LFC (you can change to FDR ordering)
  GSEA_levels  <- GSEA_plotdf  %>% group_by(term)  %>% summarise(med = median(lfc, na.rm = TRUE))  %>%
    arrange(med) %>% pull(term)

  
  GSEA_plotdf$term  <- factor(GSEA_plotdf$term,  levels = GSEA_levels)

  GSEA_plotting_data <- bind_rows(GSEA_plotdf, GSEA_plotting_data)
}

# salinity added
{
data_temp <- subset(DE_results[, c("gene", "padj_salinity_added_wald","LFC_salinity_added_wald")])

  # change these if your column names differ
  lfc_col        <- "LFC_salinity_added_wald"  # shrunken LFC column in DEA file
  gene_col       <- "gene"            
  
 # GSEA columns
GSEA_df         <- subset(GSEA_results[GSEA_results$contrast == "salinity_added",])
GSEA_term_col   <- if ("Description" %in% names(GSEA_df)) "Description" else "ID"
GSEA_padj_col   <- if ("p.adjust"   %in% names(GSEA_df)) "p.adjust"   else "pvalue"  # fallback
GSEA_genes_col <- "core_enrichment"  # nested gene list (leading edge)

  
lfc_map <- data_temp %>%
  transmute(gene = .data[[gene_col]],
            lfc  = .data[[lfc_col]]) %>%
  distinct()


attach_lfc <- function(df, term_col, genes_col, lfc_tbl,
                       split_pattern = "[,;/\\|\\s]+") {
  out <- df %>%
    transmute(term  = .data[[term_col]],
              genes = .data[[genes_col]]) %>%
    mutate(gene = str_split(genes, split_pattern)) %>%
    select(-genes) %>%
    unnest(gene) %>%
    mutate(gene = str_trim(gene)) %>%
    filter(!is.na(gene), gene != "") %>%
    distinct(term, gene) %>%                      # drop duplicates within a term
    left_join(lfc_tbl, by = "gene")
  out
}


GSEA_long  <- attach_lfc(GSEA_df,  GSEA_term_col,  GSEA_genes_col,  lfc_map)
  

  GSEA_long  <- attach_lfc(GSEA_df,  GSEA_term_col,  GSEA_genes_col,  lfc_map)

  GSEA_terms  <- GSEA_df  %>% arrange(.data[[GSEA_padj_col]]) %>% pull(!!GSEA_term_col)
  
  GSEA_plotdf  <- GSEA_long  %>% filter(term %in% GSEA_terms) %>% mutate(contrast = "salinity_added")   # Change contrast name here

  
  # order terms inside each panel by median LFC (you can change to FDR ordering)
  GSEA_levels  <- GSEA_plotdf  %>% group_by(term)  %>% summarise(med = median(lfc, na.rm = TRUE))  %>%
    arrange(med) %>% pull(term)

  
  GSEA_plotdf$term  <- factor(GSEA_plotdf$term,  levels = GSEA_levels)

  GSEA_plotting_data <- bind_rows(GSEA_plotting_data, GSEA_plotdf)
}

# additive
{
data_temp <- subset(DE_results[, c("gene", "padj_combined_wald","LFC_combined_wald")]) # slight mis-match in names: additive = combined

  # change these if your column names differ
  lfc_col        <- "LFC_combined_wald"  # shrunken LFC column in DEA file
  gene_col       <- "gene"            
  
 # GSEA columns
GSEA_df         <- subset(GSEA_results[GSEA_results$contrast == "additive",])
GSEA_term_col   <- if ("Description" %in% names(GSEA_df)) "Description" else "ID"
GSEA_padj_col   <- if ("p.adjust"   %in% names(GSEA_df)) "p.adjust"   else "pvalue"  # fallback
GSEA_genes_col <- "core_enrichment"  # nested gene list (leading edge)

  
lfc_map <- data_temp %>%
  transmute(gene = .data[[gene_col]],
            lfc  = .data[[lfc_col]]) %>%
  distinct()


attach_lfc <- function(df, term_col, genes_col, lfc_tbl,
                       split_pattern = "[,;/\\|\\s]+") {
  out <- df %>%
    transmute(term  = .data[[term_col]],
              genes = .data[[genes_col]]) %>%
    mutate(gene = str_split(genes, split_pattern)) %>%
    select(-genes) %>%
    unnest(gene) %>%
    mutate(gene = str_trim(gene)) %>%
    filter(!is.na(gene), gene != "") %>%
    distinct(term, gene) %>%                      # drop duplicates within a term
    left_join(lfc_tbl, by = "gene")
  out
}


GSEA_long  <- attach_lfc(GSEA_df,  GSEA_term_col,  GSEA_genes_col,  lfc_map)
  

  GSEA_long  <- attach_lfc(GSEA_df,  GSEA_term_col,  GSEA_genes_col,  lfc_map)

  GSEA_terms  <- GSEA_df  %>% arrange(.data[[GSEA_padj_col]]) %>% pull(!!GSEA_term_col)
  
  GSEA_plotdf  <- GSEA_long  %>% filter(term %in% GSEA_terms) %>% mutate(contrast = "additive")   # Change contrast name here

  
  # order terms inside each panel by median LFC (you can change to FDR ordering)
  GSEA_levels  <- GSEA_plotdf  %>% group_by(term)  %>% summarise(med = median(lfc, na.rm = TRUE))  %>%
    arrange(med) %>% pull(term)

  
  GSEA_plotdf$term  <- factor(GSEA_plotdf$term,  levels = GSEA_levels)

  GSEA_plotting_data <- bind_rows(GSEA_plotting_data, GSEA_plotdf)
}

# interactive
{
data_temp <- subset(DE_results[, c("gene", "padj_interactive_only_wald","LFC_interactive_only_wald")])

  # change these if your column names differ
  lfc_col        <- "LFC_interactive_only_wald"  # shrunken LFC column in DEA file
  gene_col       <- "gene"            
  
 # GSEA columns
GSEA_df         <- subset(GSEA_results[GSEA_results$contrast == "interactive",])
GSEA_term_col   <- if ("Description" %in% names(GSEA_df)) "Description" else "ID"
GSEA_padj_col   <- if ("p.adjust"   %in% names(GSEA_df)) "p.adjust"   else "pvalue"  # fallback
GSEA_genes_col <- "core_enrichment"  # nested gene list (leading edge)

  
lfc_map <- data_temp %>%
  transmute(gene = .data[[gene_col]],
            lfc  = .data[[lfc_col]]) %>%
  distinct()


attach_lfc <- function(df, term_col, genes_col, lfc_tbl,
                       split_pattern = "[,;/\\|\\s]+") {
  out <- df %>%
    transmute(term  = .data[[term_col]],
              genes = .data[[genes_col]]) %>%
    mutate(gene = str_split(genes, split_pattern)) %>%
    select(-genes) %>%
    unnest(gene) %>%
    mutate(gene = str_trim(gene)) %>%
    filter(!is.na(gene), gene != "") %>%
    distinct(term, gene) %>%                      # drop duplicates within a term
    left_join(lfc_tbl, by = "gene")
  out
}


GSEA_long  <- attach_lfc(GSEA_df,  GSEA_term_col,  GSEA_genes_col,  lfc_map)
  

  GSEA_long  <- attach_lfc(GSEA_df,  GSEA_term_col,  GSEA_genes_col,  lfc_map)

  GSEA_terms  <- GSEA_df  %>% arrange(.data[[GSEA_padj_col]]) %>% pull(!!GSEA_term_col)
  
  GSEA_plotdf  <- GSEA_long  %>% filter(term %in% GSEA_terms) %>% mutate(contrast = "interactive")   # Change contrast name here

  
  # order terms inside each panel by median LFC (you can change to FDR ordering)
  GSEA_levels  <- GSEA_plotdf  %>% group_by(term)  %>% summarise(med = median(lfc, na.rm = TRUE))  %>%
    arrange(med) %>% pull(term)

  
  GSEA_plotdf$term  <- factor(GSEA_plotdf$term,  levels = GSEA_levels)

  GSEA_plotting_data <- bind_rows(GSEA_plotting_data, GSEA_plotdf)
}
```

### Plot GSEA

```{r image_load_ORA}
load("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 3/RData/MS_Enrichment_Analyses.RData")
```

```{r plot_ORA_results}
library(dplyr)
library(ggplot2)
library(ggridges)
library(stringr)
library(forcats)
library(scales)

plot_df <- GSEA_plotting_data %>%
  group_by(term) %>%
  summarise(
    n_contrasts = n_distinct(contrast),
    mean_abs_lfc = mean(abs(lfc), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(n_contrasts), desc(mean_abs_lfc), term) %>%
  pull(term) -> term_levels

plot_df <- GSEA_plotting_data %>%
  mutate(
    term = factor(term, levels = rev(term_levels)),
    contrast = factor(
      contrast,
      levels = c(
        "diesel_only",
        "diesel_added",
        "salinity_only",
        "salinity_added",
        "additive",
        "interactive"
      )
    )
  )

x_lim <- max(abs(plot_df$lfc), na.rm = TRUE)
hline_df <- data.frame(y = seq_along(levels(plot_df$term)))

final_GSEA_ridges <- ggplot(plot_df, aes(x = lfc, y = term)) +
  
  geom_density_ridges(
    scale = 0.9,
    rel_min_height = 0.01,
    fill = "grey85",
    colour = "grey20",
    linewidth = 0.3,
    position = position_nudge(y = -0.25)
  ) +
  
  geom_vline(
    xintercept = 0,
    linetype = 2,
    linewidth = 0.35,
    colour = "grey40"
  ) +
  
  # geom_hline(
  #   data = hline_df,
  #   aes(yintercept = y),
  #   inherit.aes = FALSE,
  #   colour = "grey85",
  #   linewidth = 0.25
  # ) +
  
  facet_grid(
    cols = vars(contrast),
    scales = "free",
    labeller = labeller(
      contrast = c(
        diesel_only = "Diesel",
        diesel_added = "+ Diesel",
        salinity_only = "Salinity",
        salinity_added = "+ Salinity",
        additive = "Additive",
        interactive = "Interactive"
      )
    )
  ) +
  
  scale_y_discrete(
    labels = function(x) stringr::str_wrap(x, width = 120)
  ) +
  
  scale_x_continuous(
    #limits = c(-x_lim, x_lim),
    breaks = pretty_breaks(n = 5),
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  
  labs(
    x = expression(log[2] * " fold-change (LFC, ashr)"),
    y = NULL
  ) +
  
  theme_bw(base_size = 10) +
  theme(
    strip.background = element_rect(fill = "grey95", colour = "black", linewidth = 0.4),
    strip.text.x = element_text(face = "bold", size = 8),
    axis.text.y = element_text(size = 5, colour = "black", lineheight = 0.9),
    axis.text.x = element_text(size = 6.5, colour = "black"),
    axis.title.x = element_text(size = 8, face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(colour = "grey85", linewidth = 0.25),
    panel.grid.major.x = element_line(colour = "grey88", linewidth = 0.25),
    panel.spacing.x = unit(0.5, "lines"),
    plot.margin = margin(5.5, 10, 5.5, 30)
  )

final_GSEA_ridges

ggsave(
  "C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_3/figures/main figures/final_GSEA_ridges.png",
  final_GSEA_ridges,
  width = 265,
  height = 159,
  units = "mm",
  dpi = 1200
)
```

# Clean-up for saving environment

```{r clean_up}
rm(BP_MF_Terms,
   data_temp,
   DE_results,
   diesel_added_Wald_ORA_results,
   diesel_only_Wald_ORA_results,
   GENES_BP,
   GENES_MF,
   GSEA_df,
   GSEA_plotdf,
   matching_file,
   ora_df,
   ora_long,
   ora_plotdf,
   TERMs_BP,
   TERMs_MF,
   tpm
)

save.image("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 3/RData/MS_Enrichment_Analyses.RData")
```

# Get protein names for lists

```{r get_protein_names}
lowest_v11_by_v1 <- function(df) {
  df |>
    dplyr::group_by(V1) |>
    dplyr::slice_min(order_by = V11, n = 1, with_ties = FALSE) |>
    dplyr::ungroup()
}

result <- lowest_v13_by_v1(matching_file)

# function to get accession ids for an input list
get_uniprot_ids <- function(df, input_values) {
  input_values <- unlist(input_values)

  output <- df$V13[df$V1 %in% input_values]

  cat(output, sep = "\n")
}

get_uniprot_ids(result, c("evm.utg729.38"))
```

# Semantic Clustering

# Need to apply semantic clustering

```{r semantic_clustering}
suppressPackageStartupMessages({
  library(dplyr)
  library(purrr)
  library(tibble)
  library(cluster)
  library(GOSemSim)
  library(org.Aequina.eg.db)
  library(ggplot2)
  library(ggVennDiagram)
  library(stringr)
})

# 0. User options # -----------------------------
OrgDb <- org.Aequina.eg.db
keytype_used <- "GID"
max_k <- 40

# 1. Helper functions-----------------------------

standardise_ontology <- function(x) {
  case_when(
    x %in% c("BP", "biological_process", "biological_processes") ~ "biological_processes",
    x %in% c("MF", "molecular_function", "molecular_functions") ~ "molecular_functions",
    x %in% c("CC", "cellular_component", "cellular_components") ~ "cellular_components",
    TRUE ~ as.character(x)
  )
}

first_non_missing <- function(x) {
  x <- x[!is.na(x) & x != ""]
  if (length(x) == 0) NA_character_ else x[1]
}

best_padj <- function(x) {
  if (all(is.na(x))) NA_real_ else min(x, na.rm = TRUE)
}

prep_enrichment <- function(df, method_name) {
  df <- as_tibble(df)

  if (!"Description_QuickGO" %in% names(df)) {
    df$Description_QuickGO <- df$Description
  }

  if (!"contrast" %in% names(df)) {
    df$contrast <- NA_character_
  }

  df %>%
    mutate(
      ID = as.character(ID),
      Description = as.character(Description),
      Description_QuickGO = as.character(Description_QuickGO),
      ontology = standardise_ontology(GO),
      p.adjust = as.numeric(p.adjust),
      contrast = as.character(contrast),
      method = method_name
    )
}

make_sim_matrix <- function(term_ids, sem_data) {
  sim <- termSim(
    t1 = term_ids,
    t2 = term_ids,
    semData = sem_data,
    method = "Wang"
  )

  sim[is.na(sim)] <- 0
  sim <- pmax(pmin(sim, 1), 0)
  diag(sim) <- 1

  sim
}

fit_semantic_clusters <- function(term_ids, sem_data, max_k = 40) {
  term_ids <- unique(term_ids)
  n_terms <- length(term_ids)

  if (n_terms == 0) {
    return(list(
      sim = NULL,
      hc = NULL,
      k_eval = tibble::tibble(),
      best_k = NA_integer_,
      clusters = integer()
    ))
  }

  sim <- make_sim_matrix(term_ids, sem_data)

  if (n_terms == 1) {
    return(list(
      sim = sim,
      hc = NULL,
      k_eval = tibble::tibble(k = 1, avg_silhouette = NA_real_),
      best_k = 1,
      clusters = setNames(1L, term_ids)
    ))
  }

  dist_mat <- stats::as.dist(1 - sim)
  hc <- stats::hclust(dist_mat, method = "average")

  if (n_terms == 2) {
    clusters <- stats::cutree(hc, k = 2)

    return(list(
      sim = sim,
      hc = hc,
      k_eval = tibble::tibble(k = 2, avg_silhouette = NA_real_),
      best_k = 2,
      clusters = clusters
    ))
  }

  k_grid <- 2:min(max_k, n_terms - 1)

  k_eval <- purrr::map_dfr(k_grid, function(k) {
    cl <- stats::cutree(hc, k = k)
    sil <- cluster::silhouette(cl, dist_mat)
    sizes <- table(cl)

    tibble::tibble(
      k = k,
      avg_silhouette = mean(sil[, "sil_width"]),
      n_singletons = sum(sizes == 1),
      prop_singletons = mean(sizes == 1)
    )
  })

  best_k <- k_eval %>%
    dplyr::filter(!is.na(avg_silhouette)) %>%
    dplyr::arrange(
      dplyr::desc(avg_silhouette),
      n_singletons,
      k
    ) %>%
    dplyr::slice_head(n = 1) %>%
    dplyr::pull(k)

  clusters <- stats::cutree(hc, k = best_k)

  list(
    sim = sim,
    hc = hc,
    k_eval = k_eval,
    best_k = best_k,
    clusters = clusters
  )
}

add_representative_go <- function(df, representative_terms_df) {
  df <- as_tibble(df)

  df %>%
    mutate(
      ID = as.character(ID),
      ontology = standardise_ontology(GO)
    ) %>%
    left_join(
      representative_terms_df,
      by = c("ontology", "ID")
    )
}

make_go_sets <- function(df, ontology_name, source_cols = c("contrast", "method")) {
  df %>%
    dplyr::filter(ontology == ontology_name) %>%
    dplyr::mutate(
      source = do.call(
        paste,
        c(dplyr::across(dplyr::all_of(source_cols)), sep = " | ")
      ),
      source = stringr::str_replace_all(source, "_", " ")
    ) %>%
    dplyr::distinct(source, ID) %>%
    split(.$source) %>%
    purrr::map(~ .x$ID)
}


plot_venn_if_possible <- function(
    go_sets,
    fill_high = "#4c43e8",
    selected_sets = NULL,
    max_sets = 7,
    label_size = 4,
    category_size = 3
) {
  go_sets <- go_sets[lengths(go_sets) > 0]

  if (!is.null(selected_sets)) {
    go_sets <- go_sets[names(go_sets) %in% selected_sets]
  }

  n_sets <- length(go_sets)

  if (n_sets < 2) {
    stop("Need at least 2 non-empty sets for a Venn diagram.")
  }

  if (n_sets > max_sets) {
    stop(
      paste0(
        "There are ", n_sets, " sets. ggVennDiagram is only practical up to ",
        max_sets, " sets.\n\nAvailable sets are:\n",
        paste(names(go_sets), collapse = "\n")
      )
    )
  }

  ggVennDiagram::ggVennDiagram(
    go_sets,
    category.names = names(go_sets),
    set_color = "black",
    set_size = category_size,
    label_alpha = 0,
    label_size = label_size
  ) +
    ggplot2::scale_fill_gradient(low = "white", high = fill_high) +
    ggplot2::theme_void(base_size = 14) +
    ggplot2::theme(
      legend.position = "none",
      plot.margin = ggplot2::margin(20, 20, 20, 20)
    )
}

plot_venn_if_possible <- function(
    go_sets,
    fill_high = "#4c43e8",
    selected_sets = NULL,
    max_sets = 7,
    label_size = 4,
    category_size = 3
) {
  go_sets <- go_sets[lengths(go_sets) > 0]

  if (!is.null(selected_sets)) {
    go_sets <- go_sets[names(go_sets) %in% selected_sets]
  }

  n_sets <- length(go_sets)

  if (n_sets < 2) {
    stop("Need at least 2 non-empty sets for a Venn diagram.")
  }

  if (n_sets > max_sets) {
    stop(
      paste0(
        "There are ", n_sets, " sets. ggVennDiagram is only practical up to ",
        max_sets, " sets.\n\nAvailable sets are:\n",
        paste(names(go_sets), collapse = "\n")
      )
    )
  }

  ggVennDiagram::ggVennDiagram(
    go_sets,
    category.names = names(go_sets),
    set_color = "black",
    set_size = category_size,
    label_alpha = 0,
    label_size = label_size
  ) +
    ggplot2::scale_fill_gradient(low = "white", high = fill_high) +
    ggplot2::theme_void(base_size = 14) +
    ggplot2::theme(
      legend.position = "none",
      plot.margin = ggplot2::margin(20, 20, 20, 20)
    )
}

# 2. Prepare and combine ORA + GSEA results -----------------------------

all_enrichment_results <- bind_rows(
  prep_enrichment(all_ORA_results,  method_name = "ORA"),
  prep_enrichment(all_GSEA_results, method_name = "GSEA")
) %>%
  mutate(enrichment_row_id = row_number())

all_enrichment_results_minimal <- all_enrichment_results %>%
  dplyr::select(
    enrichment_row_id,
    method,
    contrast,
    ontology,
    ID,
    Description,
    Description_QuickGO,
    p.adjust
  )

# 3. Create unique GO term table -----------------------------

unique_terms <- all_enrichment_results_minimal %>%
  group_by(ontology, ID) %>%
  summarise(
    Description = first_non_missing(Description),
    Description_QuickGO = first_non_missing(Description_QuickGO),
    best_p.adjust = best_padj(p.adjust),
    .groups = "drop"
  )


# 4. Build ontology-specific GOSemSim objects -----------------------------

ontology_code <- c(
  biological_processes = "BP",
  molecular_functions = "MF",
  cellular_components = "CC"
)

present_ontologies <- intersect(
  unique(unique_terms$ontology),
  names(ontology_code)
)

sem_data <- map(
  setNames(ontology_code[present_ontologies], present_ontologies),
  ~ godata(
    OrgDb,
    keytype = keytype_used,
    ont = .x,
    computeIC = FALSE
  )
)

# 5. Fit semantic clusters by ontology -----------------------------

semantic_fits <- purrr::imap(sem_data, function(sem_obj, ontology_name) {
  term_ids <- unique_terms %>%
    dplyr::filter(ontology == ontology_name) %>%
    dplyr::pull(ID)

  fit_semantic_clusters(
    term_ids = term_ids,
    sem_data = sem_obj,
    max_k = max_k
  )
})

best_k_table <- imap_dfr(semantic_fits, function(fit, ontology_name) {
  tibble(
    ontology = ontology_name,
    n_terms = length(fit$clusters),
    best_k = fit$best_k
  )
})

print(best_k_table)

k_eval_table <- imap_dfr(semantic_fits, function(fit, ontology_name) {
  fit$k_eval %>%
    mutate(
      ontology = ontology_name,
      best_k = fit$best_k,
      .before = 1
    )
})

print(k_eval_table, n = Inf)

# Optional k-selection plot
p_k_selection <- k_eval_table %>%
  filter(!is.na(avg_silhouette)) %>%
  ggplot(aes(x = k, y = avg_silhouette)) +
  geom_line() +
  geom_point() +
  geom_vline(
    aes(xintercept = best_k),
    linetype = "dashed",
    colour = "grey50"
  ) +
  facet_wrap(~ ontology, scales = "free_x") +
  labs(
    x = "k tested",
    y = "Average silhouette",
    title = "Semantic cluster k selection"
  ) +
  theme_classic(base_size = 12)

p_k_selection

# 6. Assign clusters to GO terms -----------------------------

term_clusters <- imap_dfr(semantic_fits, function(fit, ontology_name) {
  tibble(
    ontology = ontology_name,
    ID = names(fit$clusters),
    semantic_cluster = as.integer(fit$clusters),
    semantic_cluster_id = paste0(ontology_code[[ontology_name]], "_", semantic_cluster)
  )
})

unique_terms_clustered <- unique_terms %>%
  left_join(
    term_clusters,
    by = c("ontology", "ID")
  )

# 7. Pick representative GO term per semantic cluster  -----------------------------
#    Representative = lowest adjusted p-value

cluster_representatives <- unique_terms_clustered %>%
  filter(!is.na(semantic_cluster)) %>%
  group_by(ontology, semantic_cluster, semantic_cluster_id) %>%
  slice_min(
    order_by = best_p.adjust,
    n = 1,
    with_ties = FALSE
  ) %>%
  ungroup() %>%
  transmute(
    ontology,
    semantic_cluster,
    semantic_cluster_id,
    representative_ID = ID,
    representative_Description = Description,
    representative_Description_QuickGO = Description_QuickGO
  )

term_to_representative <- unique_terms_clustered %>%
  left_join(
    cluster_representatives,
    by = c("ontology", "semantic_cluster", "semantic_cluster_id")
  ) %>%
  mutate(
    representative_ID = coalesce(representative_ID, ID),
    representative_Description = coalesce(representative_Description, Description),
    representative_Description_QuickGO = coalesce(
      representative_Description_QuickGO,
      Description_QuickGO
    )
  )

# Useful mapping table for joining to any later dataframe
representative_terms_df <- term_to_representative %>%
  transmute(
    ontology,
    ID,
    original_GO = Description_QuickGO,
    semantic_cluster,
    semantic_cluster_id,
    representative_ID,
    Representative_GO = representative_Description_QuickGO
  ) %>%
  distinct()

# 8. Replace original GO terms with representative clustered terms  -----------------------------

all_enrichment_results_replaced <- all_enrichment_results_minimal %>%
  left_join(
    term_to_representative %>%
      dplyr::select(
        ontology,
        ID,
        semantic_cluster,
        semantic_cluster_id,
        representative_ID,
        representative_Description,
        representative_Description_QuickGO
      ),
    by = c("ontology", "ID")
  ) %>%
  transmute(
    enrichment_row_id,
    method,
    contrast,
    ontology,

    original_ID = ID,
    original_Description = Description,
    original_Description_QuickGO = Description_QuickGO,

    semantic_cluster,
    semantic_cluster_id,

    ID = representative_ID,
    Description = representative_Description,
    Description_QuickGO = representative_Description_QuickGO,

    p.adjust
  )

# Collapse duplicate representative terms within each contrast/method/ontology
all_enrichment_results_collapsed <- all_enrichment_results_replaced %>%
  distinct(
    contrast,
    method,
    ontology,
    ID,
    .keep_all = TRUE
  )

# 9. Cluster summary table  -----------------------------

cluster_summary <- term_to_representative %>%
  group_by(
    ontology,
    semantic_cluster,
    semantic_cluster_id,
    representative_ID,
    representative_Description_QuickGO
  ) %>%
  summarise(
    n_terms_in_cluster = n(),
    best_p.adjust = best_padj(best_p.adjust),
    clustered_GO_IDs = paste(sort(unique(ID)), collapse = "; "),
    clustered_GO_terms = paste(sort(unique(Description_QuickGO)), collapse = " | "),
    .groups = "drop"
  ) %>%
  arrange(ontology, semantic_cluster)

print(cluster_summary, n = Inf)

# 10. Intersection table after semantic collapsing  -----------------------------

intersection_df <- all_enrichment_results_collapsed %>%
  mutate(
    source = paste(
      str_replace_all(contrast, "_", " "),
      method,
      sep = " | "
    )
  ) %>%
  group_by(ontology, ID) %>%
  summarise(
    Description_QuickGO = first_non_missing(Description_QuickGO),
    groups_intersecting = paste(sort(unique(source)), collapse = " || "),
    n_groups = n_distinct(source),
    .groups = "drop"
  ) %>%
  filter(n_groups > 1) %>%
  arrange(ontology, desc(n_groups), ID)

print(intersection_df, n = Inf)

# 11. Add representative GO terms back to original ORA/GSEA results  -----------------------------

all_ORA_results_semantic <- add_representative_go(
  all_ORA_results,
  representative_terms_df
)

all_GSEA_results_semantic <- add_representative_go(
  all_GSEA_results,
  representative_terms_df
)


# 12. Make new enrichment plots with semantic names -----------------------------------

# Attach LFC data to semantically clustered ORA/GSEA results # ======================
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(stringr)
  library(ggplot2)
  library(ggridges)
  library(scales)
})

# 1. Contrast-to-LFC lookup table  -----------------------------
contrast_lfc_lookup <- tibble::tribble(
  ~contrast,          ~lfc_col,
  "diesel_only",      "LFC_diesel_only_wald",
  "diesel_added",     "LFC_diesel_added_wald",
  "salinity_only",    "LFC_salinity_only_wald",
  "salinity_added",   "LFC_salinity_added_wald",
  "combined",         "LFC_combined_wald",
  "additive",         "LFC_combined_wald",
  "interactive",      "LFC_interactive_only_wald"
)

contrast_levels <- c(
  "diesel_only",
  "diesel_added",
  "salinity_only",
  "salinity_added",
  "combined",
  "additive",
  "interactive"
)

contrast_labels <- c(
  diesel_only    = "Diesel",
  diesel_added   = "+ Diesel",
  salinity_only  = "Salinity",
  salinity_added = "+ Salinity",
  combined       = "Additive",
  additive       = "Additive",
  interactive    = "Interactive"
)

# 2. Helper functions  -----------------------------
pick_term_column <- function(df) {
  possible_cols <- c(
    "Representative_GO",
    "representative_Description_QuickGO",
    "Description_QuickGO",
    "Description",
    "ID"
  )

  term_col <- possible_cols[possible_cols %in% names(df)][1]

  if (is.na(term_col)) {
    stop("No suitable GO term column found.")
  }

  term_col
}

split_gene_column <- function(df, genes_col, split_pattern = "[,;/\\|\\s]+") {
  df %>%
    dplyr::mutate(
      genes = stringr::str_split(.data[[genes_col]], split_pattern)
    ) %>%
    tidyr::unnest(genes) %>%
    dplyr::mutate(
      gene = stringr::str_trim(genes)
    ) %>%
    dplyr::filter(!is.na(gene), gene != "") %>%
    dplyr::select(-genes)
}

make_lfc_map <- function(DE_results, lfc_col, gene_col = "gene") {
  if (!lfc_col %in% names(DE_results)) {
    stop(paste("Missing LFC column in DE_results:", lfc_col))
  }

  DE_results %>%
    dplyr::transmute(
      gene = .data[[gene_col]],
      lfc = as.numeric(.data[[lfc_col]])
    ) %>%
    dplyr::distinct()
}

make_enrichment_lfc_plotting_data <- function(
    enrichment_df,
    DE_results,
    genes_col,
    method_name,
    sig_cutoff = 0.05,
    gene_col = "gene",
    contrast_lookup = contrast_lfc_lookup
) {
  term_col <- pick_term_column(enrichment_df)

  if (!genes_col %in% names(enrichment_df)) {
    stop(paste("Missing gene-list column:", genes_col))
  }

  enrichment_sig <- enrichment_df %>%
    dplyr::filter(!is.na(p.adjust), p.adjust < sig_cutoff) %>%
    dplyr::mutate(
      term = as.character(.data[[term_col]]),
      original_GO_ID = as.character(ID),
      original_GO_term = as.character(Description),
      ontology = dplyr::case_when(
        "ontology" %in% names(.) ~ as.character(ontology),
        "GO" %in% names(.) ~ as.character(GO),
        TRUE ~ NA_character_
      )
    ) %>%
    dplyr::filter(!is.na(term), term != "")

  purrr::map_dfr(seq_len(nrow(contrast_lookup)), function(i) {
    contrast_i <- contrast_lookup$contrast[i]
    lfc_col_i  <- contrast_lookup$lfc_col[i]

    enrichment_i <- enrichment_sig %>%
      dplyr::filter(contrast == contrast_i)

    if (nrow(enrichment_i) == 0) {
      return(tibble::tibble())
    }

    if (!lfc_col_i %in% names(DE_results)) {
      warning(paste("Skipping", contrast_i, "- missing column:", lfc_col_i))
      return(tibble::tibble())
    }

    lfc_map <- make_lfc_map(
      DE_results = DE_results,
      lfc_col = lfc_col_i,
      gene_col = gene_col
    )

    enrichment_i %>%
      dplyr::select(
        contrast,
        ontology,
        term,
        original_GO_ID,
        original_GO_term,
        p.adjust,
        dplyr::all_of(genes_col)
      ) %>%
      split_gene_column(genes_col = genes_col) %>%
      dplyr::distinct(
        contrast,
        ontology,
        term,
        original_GO_ID,
        gene,
        .keep_all = TRUE
      ) %>%
      dplyr::left_join(lfc_map, by = "gene") %>%
      dplyr::mutate(method = method_name)
  })
}

# ORA plotting data  ---------------------
ORA_plotting_data <- make_enrichment_lfc_plotting_data(
  enrichment_df = all_ORA_results_semantic,
  DE_results = DE_results,
  genes_col = "geneID",
  method_name = "ORA",
  sig_cutoff = 0.05
)

ORA_plotting_data <- ORA_plotting_data %>%
  dplyr::filter(!is.na(lfc))

dplyr::glimpse(ORA_plotting_data)

# plot ORA -----------------
ora_term_levels <- ORA_plotting_data %>%
  dplyr::group_by(term) %>%
  dplyr::summarise(
    n_contrasts = dplyr::n_distinct(contrast),
    mean_abs_lfc = mean(abs(lfc), na.rm = TRUE),
    median_lfc = median(lfc, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::arrange(desc(n_contrasts), desc(mean_abs_lfc), median_lfc, term) %>%
  dplyr::pull(term)

ORA_plot_df <- ORA_plotting_data %>%
  dplyr::mutate(
    term = factor(term, levels = rev(ora_term_levels)),
    contrast = factor(contrast, levels = contrast_levels)
  ) %>%
  dplyr::filter(!is.na(contrast))

p_ORA_boxes <- ggplot(
  ORA_plot_df,
  aes(x = lfc, y = term)
) +
  geom_boxplot(
    outlier.size = 0.6,
    width = 0.6,
    colour = "grey30",
    fill = "grey88"
  ) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    linewidth = 0.6,
    colour = "grey60"
  ) +
  facet_grid(
    cols = vars(contrast),
    scales = "free_x",
    labeller = labeller(contrast = contrast_labels)
  ) +
  scale_y_discrete(
    labels = function(x) stringr::str_wrap(x, width = 80)
  ) +
  labs(
    x = "Shrunken log2 fold-change (LFC)",
    y = NULL
  ) +
  theme_bw(base_size = 8) +
  theme(
    strip.background = element_rect(fill = "white"),
    strip.text.x = element_text(face = "bold", size = 8),
    axis.text.y = element_text(size = 10, colour = "black"),
    axis.text.x = element_text(size = 6.5, colour = "black"),
    panel.grid.minor = element_blank(),
    panel.spacing.x = unit(0.5, "lines"),
    plot.margin = margin(5.5, 10, 5.5, 30)
  )

p_ORA_boxes

ggsave(
  "C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_3/figures/main figures/ORA_boxplot_semantic_representatives.png",
  p_ORA_boxes,
  width = 246,
  height = 159,
  units = "mm",
  dpi = 300
)

## GSEA plotting -----------

# Calculate whether each GSEA ridge peaks below or above zero -----------

GSEA_plotting_data <- make_enrichment_lfc_plotting_data(
  enrichment_df = all_GSEA_results_semantic,
  DE_results = DE_results,
  genes_col = "core_enrichment",
  method_name = "GSEA",
  sig_cutoff = 0.05
)

GSEA_plotting_data <- GSEA_plotting_data %>%
  dplyr::filter(!is.na(lfc))

dplyr::glimpse(GSEA_plotting_data)

gsea_term_levels <- GSEA_plotting_data %>%
  dplyr::group_by(term) %>%
  dplyr::summarise(
    n_contrasts = dplyr::n_distinct(contrast),
    mean_abs_lfc = mean(abs(lfc), na.rm = TRUE),
    median_lfc = median(lfc, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::arrange(desc(n_contrasts), desc(mean_abs_lfc), median_lfc, term) %>%
  dplyr::pull(term)


# Adjust these if you want more or less trimming
lfc_limits <- quantile(
  GSEA_plot_df$lfc,
  probs = c(0.01, 0.999),
  na.rm = TRUE
)

GSEA_plot_df_trimmed <- GSEA_plot_df %>%
  dplyr::filter(
    !is.na(lfc),
    lfc >= lfc_limits[1],
    lfc <= lfc_limits[2]
  )

get_density_peak <- function(x) {
  x <- x[!is.na(x)]

  if (length(unique(x)) < 2) {
    return(NA_real_)
  }

  d <- density(x)
  d$x[which.max(d$y)]
}

ridge_peak_df <- GSEA_plot_df_trimmed %>%
  dplyr::group_by(contrast, term) %>%
  dplyr::summarise(
    peak_lfc = get_density_peak(lfc),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    peak_direction = dplyr::case_when(
      is.na(peak_lfc) ~ "Peak unavailable",
      peak_lfc < 0   ~ "Peak < 0",
      peak_lfc > 0   ~ "Peak > 0",
      TRUE           ~ "Peak = 0"
    )
  )

GSEA_plot_df_coloured <- GSEA_plot_df_trimmed %>%
  dplyr::left_join(
    ridge_peak_df,
    by = c("contrast", "term")
  )

nes_direction_df <- all_GSEA_results_semantic %>%
  dplyr::select(
    contrast,
    term = Description,
    NES
  ) %>%
  dplyr::distinct() %>%
  dplyr::mutate(
    NES_direction = dplyr::case_when(
      NES > 0  ~ "Positive NES",
      NES < 0  ~ "Negative NES",
      NES == 0 ~ "NES = 0",
      TRUE     ~ "NES unavailable"
    )
  )

GSEA_plot_df_coloured <- GSEA_plot_df_trimmed %>%
  dplyr::left_join(
    nes_direction_df,
    by = c("contrast", "term")
  )

final_GSEA_ridges <- ggplot(
  GSEA_plot_df_coloured,
  aes(
    x = lfc,
    y = term,
    fill = NES_direction,
    colour = NES_direction
  )
) +
  ggridges::geom_density_ridges(
    scale = 0.9,
    rel_min_height = 0.01,
    linewidth = 0.3,
    alpha = 0.85,
    position = position_nudge(y = -0.25)
  ) +
  geom_vline(
    xintercept = 0,
    linetype = 2,
    linewidth = 0.35,
    colour = "grey30"
  ) +
  facet_grid(
    cols = vars(contrast),
    scales = "free_x",
    labeller = labeller(contrast = contrast_labels)
  ) +
  scale_y_discrete(
    labels = function(x) stringr::str_wrap(x, width = 100)
  ) +
  scale_x_continuous(
    breaks = scales::pretty_breaks(n = 5),
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  scale_fill_manual(
    values = c(
      "Negative NES" = "#4C78A8",
      "Positive NES" = "#E45756",
      "NES = 0" = "grey75",
      "NES unavailable" = "grey85"
    ),
    name = "NES direction"
  ) +
  scale_colour_manual(
    values = c(
      "Negative NES" = "#14467D",
      "Positive NES" = "#BD1F1E",
      "NES = 0" = "grey45",
      "NES unavailable" = "grey65"
    ),
    name = "NES direction"
  ) +
  labs(
    x = expression(log[2] * " fold-change (LFC, ashr)"),
    y = NULL
  ) +
  theme_bw(base_size = 10) +
  theme(
    strip.background = element_rect(fill = "grey95", colour = "black", linewidth = 0.4),
    strip.text.x = element_text(face = "bold", size = 10),
    axis.text.y = element_text(size = 9, colour = "black", lineheight = 0.9),
    axis.text.x = element_text(size = 8, colour = "black"),
    axis.title.x = element_text(size = 10, face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(size = 8, face = "bold"),
    legend.text = element_text(size = 10),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(colour = "grey85", linewidth = 0.25),
    panel.grid.major.x = element_line(colour = "grey88", linewidth = 0.25),
    panel.spacing.x = unit(0.5, "lines"),
    plot.margin = margin(5.5, 10, 5.5, 30)
  )

final_GSEA_ridges

ggsave(
  "C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_3/figures/main figures/GSEA_ridges_semantic_representatives.png",
  final_GSEA_ridges,
  width = 265,
  height = 180,
  units = "mm",
  dpi = 1200
)

getwd()
write.csv(GSEA_plot_df_trimmed, "GSEA_plot_df_trimmed.csv")
write.csv(GSEA_results, "GSEA_results.csv")

nes_lookup <- GSEA_results %>%
  dplyr::select(contrast, ID, NES) %>%
  distinct()

gsea_plot_df_joined <- GSEA_plot_df_trimmed %>%
  left_join(
    nes_lookup,
    by = join_by(contrast, original_GO_ID == ID),
    relationship = "many-to-one"
  )

# wc
# Should be the same number of rows as the original plot dataframe
nrow(GSEA_plot_df_trimmed)
nrow(gsea_plot_df_joined)

# Should be 0 for your current files
sum(is.na(gsea_plot_df_joined$NES))

# save
write_csv(gsea_plot_df_joined, "GSEA_plot_df_trimmed_with_NES.csv")

# Read joined file with NES already included ----
GSEA_plot_df_coloured <- read_csv(
  "GSEA_plot_df_trimmed_with_NES.csv",
  show_col_types = FALSE
) %>%
  dplyr::select(-any_of("...1")) %>%
  mutate(
    NES = as.numeric(NES),
    NES_direction = case_when(
      NES > 0  ~ "Positive NES",
      NES < 0  ~ "Negative NES",
      NES == 0 ~ "NES = 0",
      TRUE     ~ "NES unavailable"
    ),
    NES_direction = factor(
      NES_direction,
      levels = c("Positive NES", "Negative NES", "NES = 0", "NES unavailable")
    )
  )

GSEA_plot_df_coloured %>%
  summarise(
    n_rows = n(),
    missing_NES = sum(is.na(NES)),
    missing_NES_direction = sum(is.na(NES_direction)),
    positive_NES = sum(NES > 0, na.rm = TRUE),
    negative_NES = sum(NES < 0, na.rm = TRUE),
    zero_NES = sum(NES == 0, na.rm = TRUE)
  )

# Alphabetical ordering from top to bottom ----
term_levels_alpha <- GSEA_plot_df_coloured %>%
  dplyr::distinct(term) %>%
  dplyr::arrange(stringr::str_to_lower(term)) %>%
  dplyr::pull(term)

GSEA_plot_df_coloured_alpha <- GSEA_plot_df_coloured %>%
  dplyr::mutate(
    term_alpha = factor(
      term,
      levels = rev(term_levels_alpha)
    )
  )

# Set y-axis label wrapping width ----
y_label_width <- 50

# plot
final_GSEA_ridges <- ggplot(
  GSEA_plot_df_coloured_alpha,
  aes(
    x = lfc,
    y = term_alpha,
    fill = NES_direction,
    colour = NES_direction
  )
) +
  ggridges::geom_density_ridges(
    scale = 0.9,
    rel_min_height = 0.01,
    linewidth = 0.3,
    alpha = 0.85,
    position = position_nudge(y = -0.25)
  ) +
  geom_vline(
    xintercept = 0,
    linetype = 2,
    linewidth = 0.35,
    colour = "grey30"
  ) +
  facet_grid(
    cols = vars(contrast),
    scales = "free_x",
    labeller = labeller(contrast = contrast_labels)
  ) +
  scale_y_discrete(
  labels = function(x) stringr::str_wrap(x, width = y_label_width)
  ) +
  scale_x_continuous(
    breaks = scales::pretty_breaks(n = 5),
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  scale_fill_manual(
    values = c(
      "Negative NES" = "#4C78A8",
      "Positive NES" = "#E45756",
      "NES = 0" = "grey75",
      "NES unavailable" = "grey85"
    ),
    name = "NES direction"
  ) +
  scale_colour_manual(
    values = c(
      "Negative NES" = "#14467D",
      "Positive NES" = "#BD1F1E",
      "NES = 0" = "grey45",
      "NES unavailable" = "grey65"
    ),
    name = "NES direction"
  ) +
  labs(
    x = expression(log[2] * " fold-change (LFC, ashr)"),
    y = NULL
  ) +
  theme_bw(base_size = 10) +
  theme(
    strip.background = element_rect(fill = "grey95", colour = "black", linewidth = 0.4),
    strip.text.x = element_text(face = "bold", size = 10),
    axis.text.y = element_text(size = 9, colour = "black", lineheight = 0.9),
    axis.text.x = element_text(size = 8, colour = "black"),
    axis.title.x = element_text(size = 10, face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(size = 8, face = "bold"),
    legend.text = element_text(size = 10),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(colour = "grey85", linewidth = 0.25),
    panel.grid.major.x = element_line(colour = "grey88", linewidth = 0.25),
    panel.spacing.x = unit(0.5, "lines"),
    plot.margin = margin(5.5, 10, 5.5, 30)
  )

final_GSEA_ridges

ggsave(
  "C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_3/figures/main figures/final_GSEA_ridges.png",
  final_GSEA_ridges,
  width = 280,
  height = 280,
  units = "mm",
  dpi = 600
)

# Jittered ORA dotplot -----
# ORA dotplot with jittered points
# All contrasts/contexts on one panel
# Points coloured by contrast
library(dplyr)
library(ggplot2)
library(stringr)

# Order GO terms by how many contrasts they appear in, then by mean absolute LFC
ora_term_levels <- ORA_plotting_data %>%
  filter(!is.na(lfc)) %>%
  group_by(term) %>%
  summarise(
    n_contrasts = n_distinct(contrast),
    mean_abs_lfc = mean(abs(lfc), na.rm = TRUE),
    median_lfc = median(lfc, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(n_contrasts), desc(mean_abs_lfc), median_lfc, term) %>%
  pull(term)

ORA_dotplot_df <- ORA_plotting_data %>%
  filter(!is.na(lfc)) %>%
  mutate(
    term = factor(term, levels = rev(ora_term_levels)),
    contrast = factor(
      contrast,
      levels = c(
        "diesel_only",
        "diesel_added",
        "salinity_only",
        "salinity_added",
        "combined",
        "additive",
        "interactive"
      )
    )
  ) %>%
  filter(!is.na(contrast))

contrast_labels <- c(
  diesel_only    = "Diesel",
  diesel_added   = "+ Diesel",
  salinity_only  = "Salinity",
  salinity_added = "+ Salinity",
  combined       = "Additive",
  additive       = "Additive",
  interactive    = "Interactive"
)

lfc_breaks <- function(x) {
  rng <- range(x, na.rm = TRUE)
  max_abs <- max(abs(rng))

  step <- dplyr::case_when(
    max_abs <= 2  ~ 0.5,
    max_abs <= 5  ~ 1,
    max_abs <= 10 ~ 2,
    max_abs <= 20 ~ 5,
    TRUE          ~ 10
  )

  breaks <- seq(
    from = -ceiling(max_abs / step) * step,
    to   =  ceiling(max_abs / step) * step,
    by   = step
  )

  breaks[breaks >= rng[1] & breaks <= rng[2]]
}

contrast_order <- c(
  "diesel_only",
  "diesel_added",
  "salinity_only",
  "salinity_added",
  "additive",
  "combined",
  "interactive"
)

term_order <- ORA_dotplot_df %>%
  dplyr::mutate(
    contrast = factor(contrast, levels = contrast_order)
  ) %>%
  dplyr::arrange(contrast, term) %>%
  dplyr::pull(term) %>%
  unique()

ORA_dotplot_df <- ORA_dotplot_df %>%
  dplyr::mutate(
    term = factor(term, levels = rev(term_order)),
    LFC_direction = dplyr::case_when(
      lfc > 0  ~ "Positive LFC",
      lfc < 0  ~ "Negative LFC",
      lfc == 0 ~ "LFC = 0",
      TRUE     ~ "LFC unavailable"
    ),
    LFC_direction = factor(
      LFC_direction,
      levels = c("Positive LFC", "Negative LFC", "LFC = 0", "LFC unavailable")
    )
  )

p_ORA_dotplot <- ggplot(
  ORA_dotplot_df,
  aes(
    x = lfc,
    y = term,
    colour = LFC_direction
  )
) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    linewidth = 0.90,
    colour = "grey35"
  ) +
  geom_point(
    position = position_jitter(
      width = 0,
      height = 0.18,
      seed = 123
    ),
    size = 2.5,
    alpha = 0.65
  ) +
  scale_colour_manual(
    values = c(
      "Positive LFC" = "#E45756",
      "Negative LFC" = "#4C78A8",
      "LFC = 0" = "grey60",
      "LFC unavailable" = "grey85"
    ),
    name = "LFC direction"
  ) +
  scale_y_discrete(
    labels = function(x) stringr::str_wrap(x, width = 80)
  ) +
  scale_x_continuous(
    breaks = lfc_breaks,
    expand = expansion(mult = c(0.1, 0.1))
  ) +
  labs(
    x = "Shrunken log2 fold-change (LFC, ashr)",
    y = NULL
  ) +
  facet_grid(
    ~contrast,
    scales = "free",
    labeller = labeller(contrast = contrast_labels)
  ) +
  theme_bw(base_size = 8) +
  theme(
    axis.text.y = element_text(size = 10, colour = "black"),
    axis.text.x = element_text(size = 8, colour = "black"),
    axis.title.x = element_text(size = 10, face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    strip.text.x = element_text(size = 12, face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(colour = "grey88", linewidth = 0.25),
    panel.grid.major.x = element_line(colour = "grey90", linewidth = 0.25),
    plot.margin = margin(5.5, 10, 5.5, 30)
  )

p_ORA_dotplot

ggsave(
  "C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_3/figures/main figures/ORA_dotplot_semantic_representatives.png",
  p_ORA_dotplot,
  width = 280,
  height = 170,
  units = "mm",
  dpi = 600
)

write.csv(ORA_dotplot_df, "ORA_dotplot_df")
# 3 panel dotplot? ----
# ============================================================
# ORA semantic dotplot: 3 panels
# Diesel treatments | Salinity treatments | Multi-stress contexts
# ============================================================

library(dplyr)
library(ggplot2)
library(stringr)
library(scales)

ORA_dotplot_df <- ORA_plotting_data %>%
  filter(!is.na(lfc)) %>%
  mutate(
    context_group = case_when(
      contrast %in% c("diesel_only", "diesel_added") ~ "Diesel treatments",
      contrast %in% c("salinity_only", "salinity_added") ~ "Salinity treatments",
      contrast %in% c("combined", "additive", "interactive") ~ "Multi-stress contexts",
      TRUE ~ "Other"
    ),
    context_group = factor(
      context_group,
      levels = c(
        "Diesel treatments",
        "Salinity treatments",
        "Multi-stress contexts"
      )
    ),
    contrast = factor(
      contrast,
      levels = c(
        "diesel_only",
        "diesel_added",
        "salinity_only",
        "salinity_added",
        "combined",
        "additive",
        "interactive"
      )
    )
  ) %>%
  filter(!is.na(context_group), !is.na(contrast))

ora_term_levels <- ORA_dotplot_df %>%
  group_by(term) %>%
  summarise(
    n_contexts = n_distinct(contrast),
    mean_abs_lfc = mean(abs(lfc), na.rm = TRUE),
    median_lfc = median(lfc, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(n_contexts), desc(mean_abs_lfc), median_lfc, term) %>%
  pull(term)

ORA_dotplot_df <- ORA_dotplot_df %>%
  mutate(
    term = factor(term, levels = rev(ora_term_levels))
  )

p_ORA_dotplot_3panel <- ggplot(
  ORA_dotplot_df,
  aes(
    x = lfc,
    y = term,
    colour = contrast
  )
) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    linewidth = 0.5,
    colour = "grey55"
  ) +
  geom_point(
    position = position_jitter(
      width = 0,
      height = 0.18,
      seed = 123
    ),
    size = 3,
    alpha = 0.25
  ) +
 # facet_wrap(
 #   ~ context_group,
 #   nrow = 1,
 #   scales = "free_y"
 # ) +
    facet_grid(
    ~ context_group
  ) +
  scale_colour_manual(
    values = c(
      diesel_only    = "#1B9E77",
      diesel_added   = "#377EB8",
      salinity_only  = "#E41A1C",
      salinity_added = "#FF7F00",
      combined       = "#4DAF4A",
      additive       = "#4DAF4A",
      interactive    = "#984EA3"
    ),
    labels = c(
      diesel_only    = "Diesel",
      diesel_added   = "+ Diesel",
      salinity_only  = "Salinity",
      salinity_added = "+ Salinity",
      combined       = "Additive",
      additive       = "Additive",
      interactive    = "Interactive"
    ),
    name = "Context"
  ) +
  scale_y_discrete(
    labels = function(x) stringr::str_wrap(x, width = 45)
  ) +
  scale_x_continuous(
    breaks = seq(-10, 10, by = 5),
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  coord_cartesian(xlim = c(-10, 10)) +
  labs(
    x = "Shrunken log2 fold-change (LFC, ashr)",
    y = NULL
  ) +
  theme_bw(base_size = 8) +
  theme(
    strip.background = element_rect(fill = "grey95", colour = "black", linewidth = 0.4),
    strip.text.x = element_text(face = "bold", size = 8),
    axis.text.y = element_text(size = 9, colour = "black"),
    axis.text.x = element_text(size = 9, colour = "black"),
    axis.title.x = element_text(size = 10, face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 10),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(colour = "grey88", linewidth = 0.25),
    panel.grid.major.x = element_line(colour = "grey90", linewidth = 0.25),
    panel.spacing.x = unit(1.2, "lines"),
    plot.margin = margin(5.5, 10, 5.5, 20)
  )

p_ORA_dotplot_3panel

ggsave(
  "C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_3/figures/main figures/ORA_dotplot_semantic_representatives.png",
  p_ORA_dotplot_3panel,
  width = 246,
  height = 200,
  units = "mm",
  dpi = 300
)

```

# Venn diagram of GO terms

```{r venn_of_enrichments}
# ============================================================
# Combine ORA and GSEA results + make Venn diagrams
# ============================================================

library(readr)
library(dplyr)
library(stringr)
library(purrr)
library(ggplot2)

# Install if needed:
# install.packages("ggVennDiagram")
library(ggVennDiagram)

# ------------------------------------------------------------
# 1. File paths
# ------------------------------------------------------------

gsea_file <- "GSEA_results.csv"
ora_file  <- "ORA_dotplot_df.csv"

out_dir <- "combined_ORA_GSEA_outputs"

if (!dir.exists(out_dir)) {
  dir.create(out_dir, recursive = TRUE)
}

# ------------------------------------------------------------
# 2. Contrast naming
# ------------------------------------------------------------
# additive and combined are treated as the same contrast.

contrast_order <- c(
  "diesel",
  "+ diesel",
  "salinity",
  "+ salinity",
  "combined",
  "interactive"
)

clean_contrast <- function(x) {
  dplyr::case_when(
    x %in% c("diesel_only", "diesel") ~ "diesel",
    x %in% c("diesel_added", "+ diesel") ~ "+ diesel",
    x %in% c("salinity_only", "salinity") ~ "salinity",
    x %in% c("salinity_added", "+ salinity") ~ "+ salinity",
    x %in% c("combined", "additive") ~ "combined",
    x %in% c("interactive", "interaction") ~ "interactive",
    TRUE ~ as.character(x)
  )
}

# ------------------------------------------------------------
# 3. Read files
# ------------------------------------------------------------

GSEA_results <- read_csv(gsea_file, show_col_types = FALSE) %>%
  select(-any_of("...1"), -any_of("Unnamed: 0"))

ORA_results <- read_csv(ora_file, show_col_types = FALSE) %>%
  select(-any_of("...1"), -any_of("Unnamed: 0"))

# ------------------------------------------------------------
# 4. Standardise GSEA results
# ------------------------------------------------------------

GSEA_combined <- GSEA_results %>%
  mutate(
    method = "GSEA",
    contrast_original = contrast,
    contrast = clean_contrast(contrast),
    
    original_GO_ID = ID,
    original_GO_term = Description,
    
    # GSEA file does not currently contain a separate semantic
    # representative term column, so Description is used here.
    semantic_GO_term = Description,
    
    p_adjust = p.adjust,
    statistic = NES,
    direction = case_when(
      NES > 0 ~ "positive",
      NES < 0 ~ "negative",
      NES == 0 ~ "zero",
      TRUE ~ NA_character_
    )
  ) %>%
  select(
    method,
    contrast,
    contrast_original,
    original_GO_ID,
    original_GO_term,
    semantic_GO_term,
    p_adjust,
    statistic,
    direction,
    everything()
  )

# ------------------------------------------------------------
# 5. Standardise ORA results
# ------------------------------------------------------------

ORA_combined <- ORA_results %>%
  mutate(
    method = "ORA",
    contrast_original = contrast,
    contrast = clean_contrast(contrast),
    
    original_GO_ID = original_GO_ID,
    original_GO_term = original_GO_term,
    
    # In your ORA file, `term` is the representative semantic term.
    semantic_GO_term = term,
    
    p_adjust = p.adjust,
    statistic = lfc,
    direction = case_when(
      lfc > 0 ~ "positive",
      lfc < 0 ~ "negative",
      lfc == 0 ~ "zero",
      TRUE ~ NA_character_
    )
  ) %>%
  select(
    method,
    contrast,
    contrast_original,
    original_GO_ID,
    original_GO_term,
    semantic_GO_term,
    p_adjust,
    statistic,
    direction,
    everything()
  )

# ------------------------------------------------------------
# 6. Combine into one dataframe
# ------------------------------------------------------------

combined_GO_results <- bind_rows(
  GSEA_combined,
  ORA_combined
) %>%
  mutate(
    contrast = factor(contrast, levels = contrast_order),
    method = factor(method, levels = c("ORA", "GSEA"))
  )

write_csv(
  combined_GO_results,
  file.path(out_dir, "combined_ORA_GSEA_results.csv")
)

# Optional QC
combined_GO_results %>%
  count(method, contrast, name = "n_rows")

combined_GO_results %>%
  distinct(method, contrast, original_GO_ID, original_GO_term, semantic_GO_term) %>%
  count(method, contrast, name = "n_unique_GO_terms")

# ------------------------------------------------------------
# 7. Helper function for contrast Venn sets
# ------------------------------------------------------------
library(tibble)
make_venn_sets <- function(df, value_col) {
  df %>%
    filter(!is.na(contrast)) %>%
    filter(contrast %in% contrast_order) %>%
    mutate(contrast = as.character(contrast)) %>%
    distinct(contrast, {{ value_col }}) %>%
    filter(!is.na({{ value_col }})) %>%
    group_by(contrast) %>%
    summarise(
      values = list(unique({{ value_col }})),
      .groups = "drop"
    ) %>%
    deframe() %>%
    .[contrast_order] %>%
    purrr::discard(is.null)
}

# ------------------------------------------------------------
# 8. Venn diagram: original GO terms
# ------------------------------------------------------------

venn_original_GO <- make_venn_sets(
  combined_GO_results,
  original_GO_ID
)

p_venn_original_GO <- ggVennDiagram(
  venn_original_GO,
  label_alpha = 0
) +
  scale_fill_gradient(low = "white", high = "grey60") +
  labs(
    title = "Overlap of original GO terms across contrasts",
    fill = "Count"
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "right"
  )

p_venn_original_GO

ggsave(
  file.path(out_dir, "venn_original_GO_terms.png"),
  p_venn_original_GO,
  width = 220,
  height = 180,
  units = "mm",
  dpi = 600
)

# ------------------------------------------------------------
# 9. Venn diagram: semantic representative GO terms
# ------------------------------------------------------------

venn_semantic_GO <- make_venn_sets(
  combined_GO_results,
  semantic_GO_term
)

p_venn_semantic_GO <- ggVennDiagram(
  venn_semantic_GO,
  label_alpha = 0
) +
  scale_fill_gradient(low = "white", high = "grey60") +
  labs(
    title = "Overlap of semantic GO terms across contrasts",
    fill = "Count"
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "right"
  )

p_venn_semantic_GO

ggsave(
  file.path(out_dir, "venn_semantic_GO_terms.png"),
  p_venn_semantic_GO,
  width = 220,
  height = 180,
  units = "mm",
  dpi = 600
)
```

# Upset plot of enrichment data

```{r upset}
# ============================================================
# UpSet plots instead of Venn diagrams
# ============================================================

# Install if needed:
# install.packages("ComplexUpset")

library(ComplexUpset)
library(tidyr)
library(dplyr)
library(ggplot2)
library(readr)

# ------------------------------------------------------------
# 1. Contrast labels for plotting
# ------------------------------------------------------------

contrast_order <- c(
  "diesel",
  "+ diesel",
  "salinity",
  "+ salinity",
  "combined",
  "interactive"
)

contrast_plot_labels <- c(
  diesel        = "Diesel",
  `+ diesel`   = "+ Diesel",
  salinity      = "Salinity",
  `+ salinity` = "+ Salinity",
  combined      = "Combined",
  interactive   = "Interactive"
)

# ------------------------------------------------------------
# 2. Helper function to make UpSet input dataframe
# ------------------------------------------------------------

make_upset_input <- function(df, value_col, value_name = "GO_term") {
  
  upset_df <- df %>%
    mutate(
      contrast = as.character(contrast),
      contrast = dplyr::case_when(
        contrast == "additive" ~ "combined",
        TRUE ~ contrast
      )
    ) %>%
    filter(contrast %in% contrast_order) %>%
    mutate(
      contrast_label = contrast_plot_labels[contrast],
      GO_value = as.character({{ value_col }})
    ) %>%
    filter(
      !is.na(GO_value),
      GO_value != ""
    ) %>%
    distinct(GO_value, contrast_label) %>%
    mutate(present = TRUE) %>%
    tidyr::pivot_wider(
      names_from = contrast_label,
      values_from = present,
      values_fill = FALSE
    )
  
  # Ensure all expected contrast columns exist, even if a contrast has no terms
  expected_cols <- unname(contrast_plot_labels[contrast_order])
  missing_cols <- setdiff(expected_cols, names(upset_df))
  
  if (length(missing_cols) > 0) {
    for (col in missing_cols) {
      upset_df[[col]] <- FALSE
    }
  }
  
  upset_df %>%
    select(
      GO_value,
      all_of(expected_cols)
    ) %>%
    rename(!!value_name := GO_value)
}

# ------------------------------------------------------------
# 3. Original GO term UpSet plot
# ------------------------------------------------------------

upset_original_GO <- make_upset_input(
  combined_GO_results,
  original_GO_ID,
  value_name = "original_GO_ID"
)

contrast_cols <- unname(contrast_plot_labels[contrast_order])

p_upset_original_GO <- ComplexUpset::upset(
  upset_original_GO,
  intersect = contrast_cols,
  name = "Contrast",
  min_size = 1,
  base_annotations = list(
    "Intersection size" = ComplexUpset::intersection_size(
      counts = TRUE
    )
  ),
  set_sizes = ComplexUpset::upset_set_size()
) +
  labs(
    title = "Overlap of original GO terms across contrasts"
  ) +
  theme(
    plot.title = element_text(
      face = "bold",
      hjust = 0.5
    )
  )

p_upset_original_GO

ggsave(
  file.path(out_dir, "upset_original_GO_terms.png"),
  p_upset_original_GO,
  width = 220,
  height = 160,
  units = "mm",
  dpi = 600
)

# ------------------------------------------------------------
# 4. Semantic representative GO term UpSet plot
# ------------------------------------------------------------

upset_semantic_GO <- make_upset_input(
  combined_GO_results,
  semantic_GO_term,
  value_name = "semantic_GO_term"
)

p_upset_semantic_GO <- ComplexUpset::upset(
  upset_semantic_GO,
  intersect = contrast_cols,
  name = "Contrast",
  min_size = 1,
  base_annotations = list(
    "Intersection size" = ComplexUpset::intersection_size(
      counts = TRUE
    )
  ),
  set_sizes = ComplexUpset::upset_set_size()
) +
  labs(
    title = "Overlap of semantic representative GO terms across contrasts"
  ) +
  theme(
    plot.title = element_text(
      face = "bold",
      hjust = 0.5
    )
  )

p_upset_semantic_GO


# Making plots look nicer ----
# ============================================================
# ORA + GSEA UpSet plots for original and semantic GO terms
# ============================================================

# Load packages ----
check_packages <- function(pkg_list) {
  for (pkg in pkg_list) {
    if (!require(pkg, character.only = TRUE)) {
      message(pkg, " is not installed")
      
      if (pkg %in% c("ComplexHeatmap")) {
        if (!requireNamespace("BiocManager", quietly = TRUE)) {
          install.packages("BiocManager")
        }
        BiocManager::install(pkg)
      } else {
        install.packages(pkg)
      }
      
      suppressMessages(library(pkg, character.only = TRUE))
    } else {
      suppressMessages(library(pkg, character.only = TRUE))
    }
  }
}

pkg_list <- c(
  "readr",
  "dplyr",
  "tidyr",
  "stringr",
  "purrr",
  "ComplexHeatmap",
  "grid",
  "tibble"
)

check_packages(pkg_list)

# ------------------------------------------------------------
# File paths
# ------------------------------------------------------------

gsea_file <- "GSEA_results.csv"
ora_file  <- "ORA_dotplot_df.csv"

out_dir <- "combined_ORA_GSEA_outputs"

if (!dir.exists(out_dir)) {
  dir.create(out_dir, recursive = TRUE)
}

# ------------------------------------------------------------
# Read files
# ------------------------------------------------------------

GSEA_results <- readr::read_csv(gsea_file, show_col_types = FALSE) %>%
  dplyr::select(-dplyr::any_of(c("...1", "Unnamed: 0")))

ORA_results <- readr::read_csv(ora_file, show_col_types = FALSE) %>%
  dplyr::select(-dplyr::any_of(c("...1", "Unnamed: 0")))

# ------------------------------------------------------------
# Contrast name harmonisation
# ------------------------------------------------------------
# additive and combined are treated as the same contrast.

clean_contrast <- function(x) {
  dplyr::case_when(
    x %in% c("diesel_only", "diesel") ~ "Diesel",
    x %in% c("salinity_only", "salinity") ~ "Salinity",
    x %in% c("diesel_added", "+ diesel", "+ Diesel") ~ "+ Diesel",
    x %in% c("salinity_added", "+ salinity", "+ Salinity") ~ "+ Salinity",
    x %in% c("combined", "additive") ~ "D + S + D*S",
    x %in% c("interactive", "interaction", "interactive_only") ~ "D*S",
    TRUE ~ as.character(x)
  )
}

set_order <- c(
  "Diesel",
  "Salinity",
  "+ Diesel",
  "+ Salinity",
  "D + S + D*S",
  "D*S"
)

# ------------------------------------------------------------
# Standardise GSEA
# ------------------------------------------------------------
# GSEA file has:
# original GO ID   = ID
# original GO term = Description
#
# It does not currently have a separate semantic representative column,
# so Description is used as semantic_GO_term for GSEA.

GSEA_combined <- GSEA_results %>%
  dplyr::mutate(
    method = "GSEA",
    contrast_original = contrast,
    contrast = clean_contrast(contrast),
    original_GO_ID = ID,
    original_GO_term = Description,
    semantic_GO_term = Description,
    p_adjust = p.adjust,
    statistic = NES,
    direction = dplyr::case_when(
      NES > 0  ~ "positive",
      NES < 0  ~ "negative",
      NES == 0 ~ "zero",
      TRUE     ~ NA_character_
    )
  ) %>%
  dplyr::select(
    method,
    contrast,
    contrast_original,
    original_GO_ID,
    original_GO_term,
    semantic_GO_term,
    p_adjust,
    statistic,
    direction,
    dplyr::everything()
  )

# ------------------------------------------------------------
# Standardise ORA
# ------------------------------------------------------------
# ORA file has:
# original GO ID   = original_GO_ID
# original GO term = original_GO_term
# semantic term    = term

ORA_combined <- ORA_results %>%
  dplyr::mutate(
    method = "ORA",
    contrast_original = contrast,
    contrast = clean_contrast(contrast),
    semantic_GO_term = term,
    p_adjust = p.adjust,
    statistic = lfc,
    direction = dplyr::case_when(
      lfc > 0  ~ "positive",
      lfc < 0  ~ "negative",
      lfc == 0 ~ "zero",
      TRUE     ~ NA_character_
    )
  ) %>%
  dplyr::select(
    method,
    contrast,
    contrast_original,
    original_GO_ID,
    original_GO_term,
    semantic_GO_term,
    p_adjust,
    statistic,
    direction,
    dplyr::everything()
  )

# ------------------------------------------------------------
# Combine
# ------------------------------------------------------------

combined_GO_results <- dplyr::bind_rows(
  ORA_combined,
  GSEA_combined
) %>%
  dplyr::mutate(
    contrast = factor(contrast, levels = set_order),
    method = factor(method, levels = c("ORA", "GSEA"))
  )

readr::write_csv(
  combined_GO_results,
  file.path(out_dir, "combined_ORA_GSEA_results.csv")
)

# QC
combined_GO_results %>%
  dplyr::count(method, contrast, name = "n_rows")

# ============================================================
# Functions for UpSet plotting
# ============================================================

make_upset_sets <- function(df, value_col, term_id_col = "GO_term") {
  
  upset_sets <- df %>%
    dplyr::filter(!is.na(contrast)) %>%
    dplyr::filter(contrast %in% set_order) %>%
    dplyr::mutate(
      contrast = as.character(contrast),
      GO_value = as.character({{ value_col }})
    ) %>%
    dplyr::filter(
      !is.na(GO_value),
      GO_value != ""
    ) %>%
    dplyr::distinct(GO_value, contrast) %>%
    dplyr::mutate(present = TRUE) %>%
    tidyr::pivot_wider(
      names_from = contrast,
      values_from = present,
      values_fill = FALSE
    )
  
  # Ensure all expected contrast columns exist
  missing_sets <- setdiff(set_order, names(upset_sets))
  
  if (length(missing_sets) > 0) {
    for (set_name in missing_sets) {
      upset_sets[[set_name]] <- FALSE
    }
  }
  
  upset_sets <- upset_sets %>%
    dplyr::select(
      GO_value,
      dplyr::all_of(set_order)
    )
  
  # Convert to logical matrix for ComplexHeatmap
  mat <- upset_sets %>%
    dplyr::select(dplyr::all_of(set_order)) %>%
    as.matrix()
  
  storage.mode(mat) <- "logical"
  
  rownames(mat) <- upset_sets$GO_value
  
  mat <- mat[rowSums(mat, na.rm = TRUE) > 0, , drop = FALSE]
  
  return(mat)
}

make_intersection_table <- function(comb_mat) {
  
  comb_codes <- ComplexHeatmap::comb_name(comb_mat)
  set_names  <- ComplexHeatmap::set_name(comb_mat)
  
  intersection_table <- tibble::tibble(
    comb_code = comb_codes,
    n_contrasts = ComplexHeatmap::comb_degree(comb_mat),
    n_terms = ComplexHeatmap::comb_size(comb_mat),
    intersection = vapply(
      comb_codes,
      function(code) {
        hits <- set_names[as.logical(as.integer(strsplit(code, "")[[1]]))]
        paste(hits, collapse = " & ")
      },
      character(1)
    )
  ) %>%
    dplyr::filter(n_terms > 0) %>%
    dplyr::select(intersection, n_contrasts, n_terms) %>%
    dplyr::arrange(
      dplyr::desc(n_terms),
      dplyr::desc(n_contrasts)
    )
  
  intersection_table
}

make_go_upset_plot <- function(
    upset_mat,
    output_file,
    intersection_table_file,
    width_px = 3250,
    height_px = 2000,
    res = 300
) {
  
  # Make combination matrix
  comb_mat <- ComplexHeatmap::make_comb_mat(upset_mat)
  
  # Remove empty combinations
  comb_mat <- comb_mat[
    ComplexHeatmap::comb_degree(comb_mat) > 0 &
      ComplexHeatmap::comb_size(comb_mat) > 0
  ]
  
  # Save intersection table
  intersection_table <- make_intersection_table(comb_mat)
  
  readr::write_csv(
    intersection_table,
    intersection_table_file
  )
  
  # Colours matched to your previous contrast aesthetics
  bg_cols <- c(
    "Diesel"        = "#98c2d9",
    "Salinity"      = "#f7e9a1",
    "+ Diesel"      = "#98c2d9",
    "+ Salinity"    = "#f7e9a1",
    "D + S + D*S"   = "#cc9593",
    "D*S"           = "#cc4551"
  )
  
  png(
    output_file,
    width = width_px,
    height = height_px,
    res = res
  )
  
  ComplexHeatmap::UpSet(
    comb_mat,
    
    set_order = set_order,
    
    comb_order = order(
      ComplexHeatmap::comb_size(comb_mat),
      decreasing = TRUE
    ),
    
    top_annotation = ComplexHeatmap::upset_top_annotation(
      comb_mat,
      add_numbers    = TRUE,
      numbers_gp     = grid::gpar(cex = 0.6),
      numbers_offset = grid::unit(1, "mm"),
      numbers_rot    = 0,
      gp             = grid::gpar(fill = "grey50", col = "grey30"),
      bar_width      = grid::unit(5, "mm"),
      ylim           = c(0, max(ComplexHeatmap::comb_size(comb_mat))),
      axis_param     = list(
        at     = pretty(c(0, max(ComplexHeatmap::comb_size(comb_mat)))),
        labels = pretty(c(0, max(ComplexHeatmap::comb_size(comb_mat)))),
        gp     = grid::gpar(cex = 0.75)
      ),
      height = grid::unit(40, "mm")
    ),
    
    right_annotation = ComplexHeatmap::upset_right_annotation(
      comb_mat,
      add_numbers = TRUE,
      numbers_gp  = grid::gpar(cex = 0.7),
      gp          = grid::gpar(fill = "grey75", col = "grey30"),
      width       = grid::unit(4, "cm"),
      bar_width   = grid::unit(6, "mm")
    ),
    
    comb_col = "black",
    bg_col = unname(bg_cols[rev(set_order)]),
    bg_pt_col = "white",
    pt_size = grid::unit(3, "mm"),
    lwd = 1.2,
    row_names_gp = grid::gpar(fontsize = 12)
  )
  
  dev.off()
  
  return(
    list(
      comb_mat = comb_mat,
      intersection_table = intersection_table
    )
  )
}

# Required packages ----
library(dplyr)
library(tidyr)
library(ComplexHeatmap)
library(grid)

# Contrast order ----
set_order <- c(
  "Diesel",
  "Salinity",
  "+ Diesel",
  "+ Salinity",
  "D + S + D*S",
  "D*S"
)

# Make logical matrix for ComplexHeatmap UpSet ----
make_upset_sets <- function(df, value_col) {
  
  upset_sets <- df %>%
    dplyr::filter(!is.na(contrast)) %>%
    dplyr::filter(contrast %in% set_order) %>%
    dplyr::mutate(
      contrast = as.character(contrast),
      GO_value = as.character({{ value_col }})
    ) %>%
    dplyr::filter(
      !is.na(GO_value),
      GO_value != ""
    ) %>%
    dplyr::distinct(GO_value, contrast) %>%
    dplyr::mutate(present = TRUE) %>%
    tidyr::pivot_wider(
      names_from = contrast,
      values_from = present,
      values_fill = FALSE
    )
  
  missing_sets <- setdiff(set_order, names(upset_sets))
  
  if (length(missing_sets) > 0) {
    for (set_name in missing_sets) {
      upset_sets[[set_name]] <- FALSE
    }
  }
  
  upset_sets <- upset_sets %>%
    dplyr::select(
      GO_value,
      dplyr::all_of(set_order)
    )
  
  upset_mat <- upset_sets %>%
    dplyr::select(dplyr::all_of(set_order)) %>%
    as.matrix()
  
  storage.mode(upset_mat) <- "logical"
  rownames(upset_mat) <- upset_sets$GO_value
  
  upset_mat <- upset_mat[rowSums(upset_mat, na.rm = TRUE) > 0, , drop = FALSE]
  
  return(upset_mat)
}

# Make ComplexHeatmap UpSet object ----
make_go_upset_object <- function(upset_mat) {
  
  comb_mat <- ComplexHeatmap::make_comb_mat(upset_mat)
  
  comb_mat <- comb_mat[
    ComplexHeatmap::comb_degree(comb_mat) > 0 &
      ComplexHeatmap::comb_size(comb_mat) > 0
  ]
  
  bg_cols <- c(
    "Diesel"      = "#98c2d9",
    "Salinity"    = "#f7e9a1",
    "+ Diesel"    = "#98c2d9",
    "+ Salinity"  = "#f7e9a1",
    "D + S + D*S" = "#cc9593",
    "D*S"         = "#cc4551"
  )
  
  ComplexHeatmap::UpSet(
    comb_mat,
    
    set_order = set_order,
    
    comb_order = order(
      ComplexHeatmap::comb_size(comb_mat),
      decreasing = TRUE
    ),
    
    top_annotation = ComplexHeatmap::upset_top_annotation(
      comb_mat,
      add_numbers    = TRUE,
      numbers_gp     = grid::gpar(cex = 0.6),
      numbers_offset = grid::unit(1, "mm"),
      numbers_rot    = 0,
      gp             = grid::gpar(fill = "grey50", col = "grey30"),
      bar_width      = grid::unit(5, "mm"),
      ylim           = c(0, max(ComplexHeatmap::comb_size(comb_mat))),
      axis_param     = list(
        at     = pretty(c(0, max(ComplexHeatmap::comb_size(comb_mat)))),
        labels = pretty(c(0, max(ComplexHeatmap::comb_size(comb_mat)))),
        gp     = grid::gpar(cex = 0.75)
      ),
      height = grid::unit(40, "mm")
    ),
    
    right_annotation = ComplexHeatmap::upset_right_annotation(
      comb_mat,
      add_numbers = TRUE,
      numbers_gp  = grid::gpar(cex = 0.7),
      gp          = grid::gpar(fill = "grey75", col = "grey30"),
      width       = grid::unit(4, "cm"),
      bar_width   = grid::unit(6, "mm")
    ),
    
    comb_col = "black",
    bg_col = unname(bg_cols[rev(set_order)]),
    bg_pt_col = "white",
    pt_size = grid::unit(3, "mm"),
    lwd = 1.2,
    row_names_gp = grid::gpar(fontsize = 12)
  )
}

# ============================================================
# Original GO term UpSet plot
# ============================================================

upset_original_GO_mat <- make_upset_sets(
  combined_GO_results,
  original_GO_ID
)

original_GO_upset <- make_go_upset_plot(
  upset_mat = upset_original_GO_mat,
  output_file = file.path(out_dir, "upset_original_GO_terms.png"),
  intersection_table_file = file.path(out_dir, "upset_original_GO_terms_intersections.csv")
)

original_GO_upset <- make_go_upset_object(
  upset_mat = upset_original_GO_mat
)

grid::grid.newpage()
ComplexHeatmap::draw(original_GO_upset)

# ============================================================
# Semantic representative GO term UpSet plot
# ============================================================

upset_semantic_GO_mat <- make_upset_sets(
  combined_GO_results,
  semantic_GO_term
)

semantic_GO_upset <- make_go_upset_object(
  upset_mat = upset_semantic_GO_mat
)

grid::grid.newpage()
ComplexHeatmap::draw(semantic_GO_upset)

# plot together
# ------------------------------------------------------------
# Plot both in one plotting space: A top, B bottom
# ------------------------------------------------------------

grid::grid.newpage()

pushViewport(
  viewport(
    layout = grid.layout(
      nrow = 2,
      ncol = 1,
      heights = unit(c(1, 1), "null")
    )
  )
)

# Panel A: Original GO terms ----
pushViewport(
  viewport(
    layout.pos.row = 1,
    layout.pos.col = 1
  )
)

ComplexHeatmap::draw(
  original_GO_upset,
  newpage = FALSE
)

grid::grid.text(
  "A",
  x = unit(2, "mm"),
  y = unit(1, "npc") - unit(2, "mm"),
  just = c("left", "top"),
  gp = gpar(
    fontsize = 18,
    fontface = "bold"
  )
)

popViewport()

# Panel B: Semantic GO terms ----
pushViewport(
  viewport(
    layout.pos.row = 2,
    layout.pos.col = 1
  )
)

ComplexHeatmap::draw(
  semantic_GO_upset,
  newpage = FALSE
)

grid::grid.text(
  "B",
  x = unit(2, "mm"),
  y = unit(1, "npc") - unit(2, "mm"),
  just = c("left", "top"),
  gp = gpar(
    fontsize = 18,
    fontface = "bold"
  )
)

popViewport()

popViewport()

# save plot
{
png(
  file.path("upset_original_and_semantic_GO_terms_combined.png"),
  width = 3250,
  height = 4000,
  res = 450
)

grid::grid.newpage()

pushViewport(
  viewport(
    layout = grid.layout(
      nrow = 2,
      ncol = 1,
      heights = unit(c(1, 1), "null")
    )
  )
)

# A: Original GO terms
pushViewport(viewport(layout.pos.row = 1, layout.pos.col = 1))

ComplexHeatmap::draw(
  original_GO_upset,
  newpage = FALSE
)

grid::grid.text(
  "A",
  x = unit(2, "mm"),
  y = unit(1, "npc") - unit(2, "mm"),
  just = c("left", "top"),
  gp = gpar(fontsize = 18, fontface = "bold")
)

popViewport()

# B: Semantic GO terms
pushViewport(viewport(layout.pos.row = 2, layout.pos.col = 1))

ComplexHeatmap::draw(
  semantic_GO_upset,
  newpage = FALSE
)

grid::grid.text(
  "B",
  x = unit(2, "mm"),
  y = unit(1, "npc") - unit(2, "mm"),
  just = c("left", "top"),
  gp = gpar(fontsize = 18, fontface = "bold")
)

popViewport()

popViewport()

dev.off()
}
```

# supplementary table
```{r supplementary_table}
# ============================================================
# Combined supplementary GO semantic clustering table
# ORA + GSEA
# ============================================================

library(dplyr)
library(stringr)
library(tibble)

# -----------------------------
# 1. Helper function
# -----------------------------

# ============================================================
# Safe supplementary table builder
# ORA + GSEA semantic enrichment results
# ============================================================

library(dplyr)
library(tibble)

safe_col <- function(df, col, default = NA) {
  if (col %in% names(df)) {
    df[[col]]
  } else {
    rep(default, nrow(df))
  }
}

make_supplementary_go_table <- function(df, method_name) {
  
  df <- as_tibble(df)
  
  # Ontology fallback
  ontology_col <- if ("ontology" %in% names(df)) {
    df$ontology
  } else if ("GO" %in% names(df)) {
    df$GO
  } else {
    rep(NA_character_, nrow(df))
  }
  
  # Representative semantic GO fallback
  representative_term <- if ("Representative_GO" %in% names(df)) {
    df$Representative_GO
  } else if ("representative_Description_QuickGO" %in% names(df)) {
    df$representative_Description_QuickGO
  } else {
    df$Description
  }
  
  representative_id <- if ("representative_ID" %in% names(df)) {
    df$representative_ID
  } else {
    df$ID
  }
  
  # Gene-list column differs between ORA and GSEA
  genes <- if ("geneID" %in% names(df)) {
    df$geneID
  } else if ("core_enrichment" %in% names(df)) {
    df$core_enrichment
  } else {
    rep(NA_character_, nrow(df))
  }
  
  df %>%
    transmute(
      Method = method_name,
      Contrast = safe_col(df, "contrast", NA_character_),
      Ontology = ontology_col,
      
      `Original GO ID` = ID,
      `Original GO term` = Description,
      
      `Semantic cluster number` = safe_col(df, "semantic_cluster", NA_integer_),
      `Semantic cluster ID` = safe_col(df, "semantic_cluster_id", NA_character_),
      `Representative semantic GO ID` = representative_id,
      `Representative semantic GO term` = representative_term,
      
      `p-value` = safe_col(df, "pvalue", NA_real_),
      `Adjusted p-value` = safe_col(df, "p.adjust", NA_real_),
      qvalue = safe_col(df, "qvalue", NA_real_),
      
      # ORA-specific columns
      GeneRatio = safe_col(df, "GeneRatio", NA_real_),
      BgRatio = safe_col(df, "BgRatio", NA_real_),
      RichFactor = safe_col(df, "RichFactor", NA_real_),
      FoldEnrichment = safe_col(df, "FoldEnrichment", NA_real_),
      zScore = safe_col(df, "zScore", NA_real_),
      Count = safe_col(df, "Count", NA_integer_),
      
      # GSEA-specific columns
      setSize = safe_col(df, "setSize", NA_integer_),
      enrichmentScore = safe_col(df, "enrichmentScore", NA_real_),
      NES = safe_col(df, "NES", NA_real_),
      rank = safe_col(df, "rank", NA_real_),
      leading_edge = safe_col(df, "leading_edge", NA_character_),
      
      Genes = as.character(genes)
    )
}


combined_GO_semantic_supplementary_table <- bind_rows(
  make_supplementary_go_table(
    all_ORA_results_semantic,
    method_name = "ORA"
  ),
  make_supplementary_go_table(
    all_GSEA_results_semantic,
    method_name = "GSEA"
  )
) %>%
  arrange(
    Ontology,
    `Representative semantic GO term`,
    Method,
    Contrast,
    `Adjusted p-value`
  )

combined_GO_semantic_supplementary_table_sig <- combined_GO_semantic_supplementary_table %>%
  filter(`Adjusted p-value` < 0.05)

combined_GO_semantic_supplementary_table_simple <- combined_GO_semantic_supplementary_table_sig %>%
  dplyr::select(
    Method,
    Contrast,
    Ontology,
    `Original GO ID`,
    `Original GO term`,
    `Semantic cluster ID`,
    `Representative semantic GO ID`,
    `Representative semantic GO term`,
    `Adjusted p-value`,
    GeneRatio,
    Count,
    setSize,
    FoldEnrichment,
    NES,
    Genes
  ) %>%
  arrange(
    Ontology,
    `Representative semantic GO term`,
    Method,
    Contrast,
    `Adjusted p-value`
  )

write.csv(
  combined_GO_semantic_supplementary_table_simple,
  "C:/Users/r02hw22/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/chapter_3/results/combined_GO_semantic_supplementary_table_simple.csv",
  row.names = FALSE
)
```
