---
title: "Differential Expression Analysis C3"
author: "Hamish Williams"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose of this Markdown

I think for the 3rd - 4th time I am re-creating a github repository so that I can consolidate and refine my code into a more digestable format so it is easier to work with.

This specific RMD will process and perform all of the analysis for fo chapter 3's omics side of things. This will not focus on making the plots. 

Instead, at the end I will have created an image of the environment which should b able to be found in the ../RData directory of this chapters directory which I will load individuall in seperate scripts to make the figures or refine results later in other scripts. 

# Load packages

```{r load_packages}
## Make Function to Install and Load Package list
check_packages <- function(pkg_list) {
  for (pkg in pkg_list) {
    if (!require(pkg, character.only = TRUE)) {
      print(paste(pkg, "is not installed, installing package..."))
      install.packages(pkg) & suppressMessages(library(pkg, character.only = TRUE))
    } else {
      suppressMessages(library(pkg, character.only = TRUE))
      print(paste(pkg, "is installed and loaded"))
    }
  }
}

## Generate a dataframe with a list of packages required
pkg_list<-c("DESeq2",
            "ggplot2",
            "dplyr",
            "tidyr",
            "ggrepel",
            "patchwork",
            "forcats",
            "purrr",
            "ComplexUpset",
            "grid",
            "ComplexHeatmap",
            "ggVennDiagram",
            "patchwork")

## Run Function::
check_packages(pkg_list)
```

# Load environment from end
```{r enviro_load}
load("MS_DEA_Envrionment.RData")  
```

# Initial Data processing

## Load data

```{r load_data}
# COunt Data
counts <- as.matrix(read.table("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 3/Data/Processed/A_Equina_Counts_unstranded_copy.txt", header = T, sep = '\t'))

countData <- as.matrix(subset(counts, select = c(-Chr,-Start,-End,-Strand,-Length,-X2)))
{apply(countData, 2, as.numeric)
  sapply(countData, as.numeric)
  class(countData) <- "numeric"
  storage.mode(countData) <- "numeric"}

# Load colData
colData <- read.table("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 3/Data/MS_colData.txt", header = T, sep = '\t', 
                      stringsAsFactors = TRUE)

# remove outlier "X2" from colData
colData <- subset(colData[c(1:24,26:32),])

reorder_idx <- match(colnames(countData), rownames(colData))
if (anyNA(reorder_idx)) {
  stop("Name mismatch after match(); check for typos/duplicates in sample names.\n",
       "Problem samples: ",
       paste(colnames(countData)[is.na(reorder_idx)], collapse=", "))
}
colData <- colData[reorder_idx, , drop = FALSE]

## Make the factors explicit and set "N" as the reference:
colData$Diesel   <- factor(colData$Diesel,   levels = c("N","Y"))
colData$Salinity <- factor(colData$Salinity, levels = c("N","Y"))
```

## calculate parameers and statistics

```{r parameters_and_statistics}
## CPM
cpm <- {apply(subset(counts, select = c(-Chr,-Start,-End,-Strand,-Length)), 2, 
              function(x) as.numeric(x)/sum(as.numeric(x)) * 10^6)}
counts_rownames <- as.vector(row.names(counts))
rownames(cpm) <- counts_rownames

## Genelengths
geneLengths <- as.vector(subset(counts, select = c(Length)))
geneLengths <- as.vector(as.numeric(geneLengths))

## RPKM
rpkm <- {apply(X = subset(counts, select = c(-Chr,-Start,-End,-Strand,-Length)),
               MARGIN = 2, 
               FUN = function(x) {
                 10^9 * as.numeric(x) / geneLengths / sum(as.numeric(x))
               })
}
rownames(rpkm) <- counts_rownames

## RPK
rpk <- {apply(subset(counts, select = c(-Chr,-Start,-End,-Strand,-Length)), 2, 
              function(x) as.numeric(x)/(geneLengths/1000))}
rownames(rpk) <- counts_rownames

## TPM
tpm <- apply(rpk, 2, function(x) as.numeric(x) / sum(as.numeric(x)) * 10^6)
rownames(tpm) <- counts_rownames

## Variance
V <- apply(tpm, 1, var) # Can make Heatmaps from these values ... 

## Select top 100 genes:
selectedGenes <- names(V[order(V, decreasing = T)][1:100])

## Check each step for errors:
colSums(cpm) # = 10^6 for ALL
colSums(rpkm) # Different values
colSums(tpm) # = 10^6 for ALL

pheatmap(tpm[selectedGenes,], scale = 'row', 
         show_rownames = FALSE, 
         annotation_col = colData[c("Group")])

# Continue code for figure in new file, something like:
  # Code/Heatmap_plotting_code.R
```


# Differential Expression ANalyses

## LRT models ----

### Diesel only
```{r diesel_H0}
## Diesel ~ H0
# Subset data to only include samples with diesel or control treatments:
{
keep <- colData$Salinity == "N"

counData_keep <- countData[, keep]
colData_keep <- droplevels(colData[keep,])

dds <- DESeqDataSetFromMatrix(countData = countData[, keep],
                              colData   = droplevels(colData[keep,]),
                              design    = ~ Diesel)

# (Optional but harmless) prefilter very low counts:
dds <- dds[rowSums(counts(dds) >= 10) >= 2, ]

# LRT
dds_diesel_H0 <- DESeq(dds, test = "LRT",
                     reduced = ~ 1)

  # Shrinking LFC for later comparison
  # see ~/Code/LRT_and_Wald_LFC_Comparison_plots.R for plotting
res_diesel_H0_ashr <- lfcShrink(
  dds  = dds_diesel_H0,
  coef = "Diesel_Y_vs_N",   # replace with exact name from resultsNames(dds_diesel_H0)
  type = "ashr"
)

res_diesel_H0 <- results(dds_diesel_H0)
res_diesel_H0 <- res_diesel_H0[order(res_diesel_H0$padj),]
view <- as.data.frame(res_diesel_H0)
sig_diesel_H0 <- subset(res_diesel_H0, !is.na(padj) & padj < 0.05)
sig_diesel_H0
}
  # 0 significant DEGs
```

### Salinity only
```{r salinity_H0}
## Salinity ~ H0
# Subset data to only include samples with Salinity or control treatments:
{
keep <- colData$Diesel == "N"

counData_keep <- countData[, keep]
colData_keep <- droplevels(colData[keep,])

dds <- DESeqDataSetFromMatrix(countData = countData[, keep],
                              colData   = droplevels(colData[keep,]),
                              design    = ~ Salinity)

# (Optional but harmless) prefilter very low counts:
dds <- dds[rowSums(counts(dds) >= 10) >= 2, ]

# LRT
dds_Salinity_H0 <- DESeq(dds, test = "LRT",
                       reduced = ~ 1)

  # Shrinking LFC for later comparison
  # see ~/Code/LRT_and_Wald_LFC_Comparison_plots.R for plotting
    # shrink the Salinity coefficient using ashr for later
    res_salinity_H0_ashr <- lfcShrink(
      dds  = dds_Salinity_H0,
      coef = "Salinity_Y_vs_N",
      type = "ashr"
    )

res_Salinity_H0 <- results(dds_Salinity_H0)
res_Salinity_H0 <- res_Salinity_H0[order(res_Salinity_H0$padj),]
view <- as.data.frame(res_Salinity_H0)
sig_Salinity_H0 <- subset(res_Salinity_H0, !is.na(padj) & padj < 0.05)
sig_Salinity_H0
}
  # 5439 DEGs
```

### Diesl + Salinity + D*S
```{r D_and_S_H0}
## Diesel + Salinity + D*S ~ H0
{
dds <- DESeqDataSetFromMatrix(countData = countData,
                              colData   = colData,
                              design    = ~ Diesel + Salinity + Diesel*Salinity)

# (Optional but harmless) prefilter very low counts:
dds <- dds[rowSums(counts(dds) >= 10) >= 2, ]

# LRT
dds_full_model_H0 <- DESeq(dds, test = "LRT",
                        reduced = ~ 1)

res_full_model_H0 <- results(dds_full_model_H0)
res_full_model_H0 <- res_full_model_H0[order(res_full_model_H0$padj),]
view <- as.data.frame(res_full_model_H0)
sig_full_model_H0 <- subset(res_full_model_H0, !is.na(padj) & padj < 0.05)
sig_full_model_H0
}
# 10844 DEGs
```

### D*S interaction only
```{r LRT_interaction}
{
dds <- DESeqDataSetFromMatrix(countData = countData,
                              colData   = colData,
                              design    = ~ Diesel + Salinity + Diesel * Salinity)

# (Optional but harmless) prefilter very low counts:
dds <- dds[rowSums(counts(dds) >= 10) >= 2, ]

# LRT
dds_interactive_LRT <- DESeq(dds, test = "LRT",
                       reduced = ~ Diesel + Salinity)

res_interactive_LRT <- results(dds_interactive_LRT)
res_interactive_LRT <- res_interactive_LRT[order(res_interactive_LRT$padj),]
res_interactive_LRT <- as.data.frame(res_interactive_LRT)
sig_interactive_LRT <- subset(res_interactive_LRT, !is.na(padj) & padj < 0.05)
sig_interactive_LRT
}

# 57 DEGs
```

## Final LRT results:
- Diesel only   :    0 
- Salinity Only :    5439
- D + S + D*S   :    10844
- D * S only    :    57


### Combine results

```{r combine_LRT_results}
# combine all results
as_prefixed_df <- function(res, prefix) {
  stopifnot(!is.null(rownames(res)))
  df <- as.data.frame(res)              # from DESeqResults/DataFrame -> data.frame
  df$gene <- rownames(res)
  
  keep_cols <- intersect(c("baseMean","log2FoldChange","lfcSE","stat","pvalue","padj"),
                         names(df))
  df <- df[, c("gene", keep_cols), drop = FALSE]
  
  # Coerce any Rle/list-like columns to plain vectors
  for (nm in names(df)) df[[nm]] <- as.vector(df[[nm]])
  
  # Deduplicate genes (keep first)
  df <- df[!duplicated(df$gene), , drop = FALSE]
  
  # Prefix non-key columns
  if (ncol(df) > 1) names(df)[-1] <- paste0(prefix, "_", names(df)[-1])
  df
}

# Build prefixed tables
diesel_df   <- as_prefixed_df(res_diesel_H0,     "dieselLRT")
salinity_df <- as_prefixed_df(res_Salinity_H0,   "salinityLRT")
full_df     <- as_prefixed_df(res_full_model_H0, "fullLRT")

# Merge by gene (full outer join) using base merge to avoid tibble/S4 quirks
merged_lrt <- Reduce(function(x, y) merge(x, y, by = "gene", all = TRUE),
                     list(diesel_df, salinity_df, full_df))
```


## Wald Models ----

### perform analysis:
```{r Wald_DEA}
dds <- DESeqDataSetFromMatrix(countData = countData,
                              colData   = colData,
                              design    = ~ Diesel + Salinity + Diesel:Salinity)

# (Optional but harmless) prefilter very low counts:
dds <- dds[rowSums(counts(dds) >= 10) >= 2, ] 
    # !!! realising this wasn't applied to all Wald methods, so some 
      # changes to results are coming as a result...

dds_wald <- DESeq(dds, test = "Wald")
resultsNames(dds_wald)
# You'll likely see names like:
#   "Diesel_Y_vs_N"
#   "Salinity_Y_vs_N"
#   "DieselY.SalinityY"   # the interaction coefficient
```

### Diesel ~ H0
```{r Diesel_only_Wald}
## Diesel ~ H0
# Do not need to subset the samples because interactions are modelled in the Wald
# design, and we only specify for diesel effects when salinity is = "N"
res_Diesel_at_SalN <- lfcShrink(dds_wald, coef = "Diesel_Y_vs_N", type = "ashr")
res_Diesel_at_SalN <- as.data.frame(res_Diesel_at_SalN)
sum(res_Diesel_at_SalN$padj < 0.05, na.rm = T)

## UNsrhunk results for comparison
res_Diesel_at_SalN_no_shrink <- results(dds_wald, name = "Diesel_Y_vs_N")
```
### Salinity ~ H0
```{r Salinity_only_Wald}
## Salinity ~ H0
res_Sal_at_DieselN <- lfcShrink(dds_wald, coef = "Salinity_Y_vs_N", type = "ashr")
res_Sal_at_DieselN <- as.data.frame(res_Sal_at_DieselN)
sum(res_Sal_at_DieselN$padj < 0.05, na.rm = T)

# unshrunk results for comparison
res_Sal_at_DieselN_no_shrink <- results(dds_wald, name = "Salinity_Y_vs_N")
```

### Diesel + Salinity + D*S
```{r full_additive_Wald}
# Diesel + Salinity + D*S ~ H0
res_combined_additive <- lfcShrink(dds_wald, 
                                   contrast = list(c("Diesel_Y_vs_N", 
                                                     "Salinity_Y_vs_N",
                                                     "DieselY.SalinityY")), 
                                   type = "ashr")

res_combined_additive <- as.data.frame(res_combined_additive)

sum(res_combined_additive$padj < 0.05, na.rm = T)
```

### Diesel * Salinity Interaction only
```{r interaction_only_Wald}
res_interactive_Wald <- lfcShrink(dds_wald, 
                                  coef = "DieselY.SalinityY",
                                  type = "ashr") 

res_interactive_Wald <- as.data.frame(res_interactive_Wald)

sum(res_interactive_Wald$padj < 0.05, na.rm = T)
```
### + Diesel Effect
```{r add_diesel_wald}
res_Sal_at_DieselY <- lfcShrink(dds_wald, 
                                contrast = list(c("Diesel_Y_vs_N", 
                                                  "DieselY.SalinityY")), 
                                type = "ashr")

res_Sal_at_DieselY <- as.data.frame(res_Sal_at_DieselY)

sum(res_Sal_at_DieselY$padj < 0.05, na.rm = T)
```
### + Salinity Effect
```{r add_salinity_wald}
res_Diesel_at_SalY <- lfcShrink(dds_wald, 
                                contrast = list(c("Salinity_Y_vs_N", 
                                                  "DieselY.SalinityY")), 
                                type = "ashr")

res_Diesel_at_SalY <- as.data.frame(res_Diesel_at_SalY)

sum(res_Diesel_at_SalY$padj < 0.05, na.rm = T)
```

## Wald Final results ----
diesel only       : 7
salinity only     : 6412
D + S + D*S       : 7610
D*S               : 125
+ D               : 1625
+ S               : 7643

# Merge and Save results

```{r save results}
# capture gene names to merge
df <- data.frame(gene = rownames(res_Diesel_at_SalN))

# reduce results to essential elements for merging
diesel_only_merge_results <- as.data.frame(subset(res_Diesel_at_SalN[, c("log2FoldChange", "padj")]))
diesel_only_merge_results$gene <- row.names(diesel_only_merge_results) 

# and change column names to be able to identify them later
names(diesel_only_merge_results)[names(diesel_only_merge_results) == "log2FoldChange"] <- "LFC_diesel_only_wald"
names(diesel_only_merge_results)[names(diesel_only_merge_results) == "padj"] <- "padj_diesel_only_wald"

# repeat for other results...
salinity_only_merge_results <- as.data.frame(subset(res_Sal_at_DieselN[, c("log2FoldChange", "padj")]))
salinity_only_merge_results$gene <- row.names(salinity_only_merge_results) 
names(salinity_only_merge_results)[names(salinity_only_merge_results) == "log2FoldChange"] <- "LFC_salinity_only_wald"
names(salinity_only_merge_results)[names(salinity_only_merge_results) == "padj"] <- "padj_salinity_only_wald"

diesel_added_merge_results <- as.data.frame(subset(res_Sal_at_DieselY[, c("log2FoldChange", "padj")]))
diesel_added_merge_results$gene <- row.names(diesel_added_merge_results) 
names(diesel_added_merge_results)[names(diesel_added_merge_results) == "log2FoldChange"] <- "LFC_diesel_added_wald"
names(diesel_added_merge_results)[names(diesel_added_merge_results) == "padj"] <- "padj_diesel_added_wald"

salinity_added_merge_results <- as.data.frame(subset(res_Diesel_at_SalY[, c("log2FoldChange", "padj")]))
salinity_added_merge_results$gene <- row.names(salinity_added_merge_results) 
names(salinity_added_merge_results)[names(salinity_added_merge_results) == "log2FoldChange"] <- "LFC_salinity_added_wald"
names(salinity_added_merge_results)[names(salinity_added_merge_results) == "padj"] <- "padj_salinity_added_wald"

interactive_only_merge_results <- as.data.frame(subset(res_interactive_Wald[, c("log2FoldChange", "padj")]))
interactive_only_merge_results$gene <- row.names(interactive_only_merge_results) 
names(interactive_only_merge_results)[names(interactive_only_merge_results) == "log2FoldChange"] <- "LFC_interactive_only_wald"
names(interactive_only_merge_results)[names(interactive_only_merge_results) == "padj"] <- "padj_interactive_only_wald"

combined_merge_results <- as.data.frame(subset(res_combined_additive[, c("log2FoldChange", "padj")]))
combined_merge_results$gene <- row.names(combined_merge_results) 
names(combined_merge_results)[names(combined_merge_results) == "log2FoldChange"] <- "LFC_combined_wald"
names(combined_merge_results)[names(combined_merge_results) == "padj"] <- "padj_combined_wald"

# and repeat for all of the LRT results too ...
diesel_H0_merge_results <- as.data.frame(subset(res_diesel_H0[, c("log2FoldChange", "padj")]))
diesel_H0_merge_results$gene <- row.names(diesel_H0_merge_results) 
names(diesel_H0_merge_results)[names(diesel_H0_merge_results) == "log2FoldChange"] <- "LFC_diesel_only_LRT"
names(diesel_H0_merge_results)[names(diesel_H0_merge_results) == "padj"] <- "padj_diesel_only_LRT"

salinity_H0_merge_results <- as.data.frame(subset(res_Salinity_H0[, c("log2FoldChange", "padj")]))
salinity_H0_merge_results$gene <- row.names(salinity_H0_merge_results) 
names(salinity_H0_merge_results)[names(salinity_H0_merge_results) == "log2FoldChange"] <- "LFC_salinity_only_LRT"
names(salinity_H0_merge_results)[names(salinity_H0_merge_results) == "padj"] <- "padj_salinity_only_LRT"

full_model_H0_merge_results <- as.data.frame(subset(res_full_model_H0[, c("log2FoldChange", "padj")]))
full_model_H0_merge_results$gene <- row.names(full_model_H0_merge_results) 
names(full_model_H0_merge_results)[names(full_model_H0_merge_results) == "log2FoldChange"] <- "LFC_full_model_LRT"
names(full_model_H0_merge_results)[names(full_model_H0_merge_results) == "padj"] <- "padj_full_model_LRT"

interactive_LRT_merge_results <- as.data.frame(subset(res_interactive_LRT[, c("log2FoldChange", "padj")]))
interactive_LRT_merge_results$gene <- row.names(interactive_LRT_merge_results) 
names(interactive_LRT_merge_results)[names(interactive_LRT_merge_results) == "log2FoldChange"] <- "LFC_interactive_only_LRT"
names(interactive_LRT_merge_results)[names(interactive_LRT_merge_results) == "padj"] <- "padj_interactive_only_LRT"


# make list
dfs <- list(df,
            diesel_only_merge_results,
            salinity_only_merge_results,
            diesel_added_merge_results,
            salinity_added_merge_results,
            interactive_only_merge_results,
            combined_merge_results,
            diesel_H0_merge_results,
            salinity_H0_merge_results,
            full_model_H0_merge_results,
            interactive_LRT_merge_results)

library(purrr)
library(dplyr)
# use "purrr" and "dplyr" to full_join
merged_df <- purrr::reduce(dfs, dplyr::full_join, by = "gene")

# save and export file
write.csv(merged_df, "C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 3/Results/Multi_Stressor_all_Differential_Expression_Analysis_results.csv", row.names = FALSE)
```


# Prepare evironment to save result
 Need to unload (rm()) files which are large and bulky and won't be used later:
 
```{r clean-up}
rm(colData_keep,
   combined_merge_results,
   counData_keep,
   countData,
   counts,
   Counts_only,
   cpm,
   dds,
   dds_diesel_H0,
   dds_full_model_H0,
   dds_interactive_LRT,
   dds_Salinity_H0,
   dds_wald,
   df,
   dfs,
   diesel_added_merge_results,
   diesel_H0_merge_results,
   diesel_only_merge_results,
   full_model_H0_merge_results,
   interactive_LRT_merge_results,
   interactive_only_merge_results,
   res_combined_additive,
   res_Diesel_at_SalN,
   res_Diesel_at_SalN_DEGs,
   res_Diesel_at_SalN_no_shrink,
   res_Diesel_at_SalY,
   res_diesel_H0,
   res_diesel_H0_ashr,
   res_full_model_H0,
   res_interactive_LRT,
   res_interactive_Wald,
   res_Sal_at_DieselN,
   res_Sal_at_DieselY,
   res_Sal_at_DieselN_no_shrink,
   res_Salinity_H0,
   res_salinity_H0_ashr,
   rpk,
   rpkm,
   salinity_added_merge_results,
   salinity_H0_merge_results,
   salinity_only_merge_results,
   sig_diesel_H0,
   sig_full_model_H0,
   sig_interactive_LRT,
   sig_Salinity_H0,
   view,
   counts_rownames,
   geneLengths,
   keep,
   pkg_list,
   reorder_idx,
   check_packages)

# Save environment
save.image("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 3/RData/MS_DEA_Envrionment.RData") 
  # load environment
load("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 3/RData/MS_DEA_Envrionment.RData")  
```