---
title: "Chapter 3 - Enrichment Analyses"
author: "Hamish Williams"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

RMarkdown Script for the enrichment analyses portion of this chapter, allows me to make easier comments on whats going on

# Load Packages
```{r packages}
# Load packages ----
check_packages <- function(pkg_list) {
  for (pkg in pkg_list) {
    if (!require(pkg, character.only = TRUE)) {
      print(paste(pkg, "is not installed, installing package..."))
      install.packages(pkg)
    } else {
      suppressMessages(library(pkg, character.only = TRUE))
      print(paste(pkg, "is installed and loaded"))
    }
  }
}

# Generate a dataframe with a list of packages required
pkg_list<-c("tidyverse", 
                            "tibble", 
                            "stats",
                            "ggplot2", 
                            "dplyr", 
                            "tidyr", 
                            "ggfortify", 
                            "knitr",
                            "clusterProfiler",
                            #"AnnotationForge",
                            "enrichplot",
                            "data.table",
                            "stats",
                            "plotly",
                            "ggVennDiagram",
                            "VennDiagram",
                            "limma",
                            "ggridges",
                            # "pathview",
                            "KEGGREST",
                            "stringr",
            "forcats",
            "ComplexHeatmap",
            "grid",
            "viridisLite",
            "RColorBrewer",
            "circlize")

# Run Function::
check_packages(pkg_list)

# Generate Functions ----
convertFractionToNumeric <- function(fraction) {
  fraction_parts <- strsplit(fraction, "/")[[1]]
  numerator <- as.numeric(fraction_parts[1])
  denominator <- as.numeric(fraction_parts[2])
  numeric_value <- numerator / denominator
  return(numeric_value)
}
```

## Load Data
```{r data}
getwd()
# Load the data
BP_MF_Terms <- read.table("../Data/Processed/UNIPROT_results.txt", 
                          header = T, 
                          sep = '\t', 
                          stringsAsFactors = FALSE)

# Import the blastp (matching) file
matching_file <- read.table("../Data/Large_Data_Files/genome_matching_file.txt", 
                            header = F, 
                            sep = '\t', 
                            stringsAsFactors = FALSE)

# Data manipulations
# Change the name of the columns to be identical
{
names(matching_file)[names(matching_file) == "V13"] <- "ID"
names(BP_MF_Terms)[names(BP_MF_Terms) == "From"] <- "ID"
names(BP_MF_Terms)[names(BP_MF_Terms) == "Gene.Names"] <- "GENENAMES"
names(BP_MF_Terms)[names(BP_MF_Terms) == "Gene.Ontology..biological.process."] <- "GO:BP"
names(BP_MF_Terms)[names(BP_MF_Terms) == "Gene.Ontology..molecular.function."] <- "GO:MF"
}

# Import DEA results
DE_results <- read.table("../Results/Multi_Stressor_all_Differential_Expression_Analysis_results.csv",
                           header = T, 
                           sep = ',', 
                           stringsAsFactors = TRUE)

## Perform naming transformation to make gene IDs matching
DE_results$gene <- gsub("\\.TU", "", DE_results$gene) # Removing "TU" part of names
DE_results$gene <- gsub("path", "mrna", DE_results$gene) # Replacing "path" -> "mrna"
```

## Generate DEGs lists and ranked lists
```{r DEGs_and_Ranked_lists}
# Diesel_only
diesel_only_geneList = DE_results[,2]
names(diesel_only_geneList) = as.character(DE_results[,1])
diesel_only_geneList = sort(diesel_only_geneList, decreasing = TRUE)

# Salinity only
salinity_only_geneList = DE_results[,4]
names(salinity_only_geneList) = as.character(DE_results[,1])
salinity_only_geneList = sort(salinity_only_geneList, decreasing = TRUE)

# Interactive
interactive_only_geneList = DE_results[,10]
names(interactive_only_geneList) = as.character(DE_results[,1])
interactive_only_geneList = sort(interactive_only_geneList, decreasing = TRUE)

# Diesel Added
diesel_added_geneList = DE_results[,6]
names(diesel_added_geneList) = as.character(DE_results[,1])
diesel_added_geneList = sort(diesel_added_geneList, decreasing = TRUE)

# Salinity added
salinity_added_geneList = DE_results[,8]
names(salinity_added_geneList) = as.character(DE_results[,1])
salinity_added_geneList = sort(salinity_added_geneList, decreasing = TRUE)

# Additive
combined_geneList = DE_results[,12]
names(combined_geneList) = as.character(DE_results[,1])
combined_geneList = sort(combined_geneList, decreasing = TRUE)

# Generate DEGs lists for each contrast ----
# diesel only
DEGs <- subset(DE_results[, c("gene", "padj_diesel_only_wald")])
DEGs <- DEGs[(DEGs$padj_diesel_only_wald < 0.05),]
DEGs <- DEGs[complete.cases(DEGs),]
diesel_only_Wald_DEGs <- as.vector(DEGs[,1])

# salinity only
DEGs <- subset(DE_results[, c("gene", "padj_salinity_only_wald")])
DEGs <- DEGs[(DEGs$padj_salinity_only_wald < 0.05),]
DEGs <- DEGs[complete.cases(DEGs),]
salinity_only_Wald_DEGs <- as.vector(DEGs[,1])

# Interactive only
DEGs <- subset(DE_results[, c("gene", "padj_interactive_only_wald")])
DEGs <- DEGs[(DEGs$padj_interactive_only_wald < 0.05),]
DEGs <- DEGs[complete.cases(DEGs),]
interactive_only_Wald_DEGs <- as.vector(DEGs[,1])

# Diesel Added
DEGs <- subset(DE_results[, c("gene", "padj_diesel_added_wald")])
DEGs <- DEGs[(DEGs$padj_diesel_added_wald < 0.05),]
DEGs <- DEGs[complete.cases(DEGs),]
diesel_added_Wald_DEGs <- as.vector(DEGs[,1])

# salinity Added
DEGs <- subset(DE_results[, c("gene", "padj_salinity_added_wald")])
DEGs <- DEGs[(DEGs$padj_salinity_added_wald < 0.05),]
DEGs <- DEGs[complete.cases(DEGs),]
salinity_added_Wald_DEGs <- as.vector(DEGs[,1])

# Additive
DEGs <- subset(DE_results[, c("gene", "padj_combined_wald")])
DEGs <- DEGs[(DEGs$padj_combined_wald < 0.05),]
DEGs <- DEGs[complete.cases(DEGs),]
combined_Wald_DEGs <- as.vector(DEGs[,1])
```

## Create GO term files
```{r GO_terms}
# Merge DEG Files:
merged_df <- merge(matching_file, BP_MF_Terms, by = "ID")

# Subset Dataframe
merged_df <- subset(merged_df[, c("V1","ID","GO:BP","GO:MF")])

# Change Variable names
names(merged_df)[names(merged_df) == "V1"] <- "GENE"
# Select BP column:
BP <- merged_df[,c("GENE","ID","GO:BP")]
BP <- BP[!apply(BP == "", 1, any), ]
BP <- BP[!duplicated(BP), ]
# Extract GO terms:
GENES_BP <- BP
GENES_BP$GO <- ""
bracketed_rows <- grepl("\\[.*\\]", GENES_BP$`GO:BP`)
GENES_BP$GO[bracketed_rows] <- gsub(".*\\[(.*)\\].*", "\\1", GENES_BP$`GO:BP`[bracketed_rows])
GENES_BP <- GENES_BP[, c("GO","GENE")]
GENES_BP <- GENES_BP[!duplicated(GENES_BP), ]
GENES_BP <- GENES_BP[complete.cases(GENES_BP),]

## Extract Terms:
TERMs_BP <- BP
TERMs_BP$GO <- ""
bracketed_rows <- grepl("\\[.*\\]", TERMs_BP$`GO:BP`)
TERMs_BP$GO[bracketed_rows] <- gsub(".*\\[(.*)\\].*", "\\1", TERMs_BP$`GO:BP`[bracketed_rows])
TERMs_BP$`GO:BP`[bracketed_rows] <- gsub("\\[.*\\]", "", TERMs_BP$`GO:BP`[bracketed_rows])
TERMs_BP <- TERMs_BP[, c("GO","GO:BP")]
TERMs_BP <- TERMs_BP[!duplicated(TERMs_BP), ]
names(TERMs_BP)[names(TERMs_BP) == "GO:BP"] <- "GENENAMES_BP"
TERMs_BP <- TERMs_BP[complete.cases(TERMs_BP),]

# Select MF column:
MF <- merged_df[,c("GENE","ID","GO:MF")]
MF <- MF[!apply(MF == "", 1, any), ]
MF <- MF[!duplicated(MF), ]
# Extract GO terms:
GENES_MF <- MF
GENES_MF$GO <- ""
bracketed_rows <- grepl("\\[.*\\]", GENES_MF$`GO:MF`)
GENES_MF$GO[bracketed_rows] <- gsub(".*\\[(.*)\\].*", "\\1", GENES_MF$`GO:MF`[bracketed_rows])
GENES_MF <- GENES_MF[, c("GO","GENE")]
GENES_MF <- GENES_MF[!duplicated(GENES_MF), ]
GENES_MF <- GENES_MF[complete.cases(GENES_MF),]

## Extract Terms:
TERMs_MF <- MF
TERMs_MF$GO <- ""
bracketed_rows <- grepl("\\[.*\\]", TERMs_MF$`GO:MF`)
TERMs_MF$GO[bracketed_rows] <- gsub(".*\\[(.*)\\].*", "\\1", TERMs_MF$`GO:MF`[bracketed_rows])
TERMs_MF$`GO:MF`[bracketed_rows] <- gsub("\\[.*\\]", "", TERMs_MF$`GO:MF`[bracketed_rows])
TERMs_MF <- TERMs_MF[, c("GO","GO:MF")]
TERMs_MF <- TERMs_MF[!duplicated(TERMs_MF), ]
names(TERMs_MF)[names(TERMs_MF) == "GO:MF"] <- "GENENAMES_MF"
TERMs_MF <- TERMs_MF[complete.cases(TERMs_MF),]

# Tidy-up Environment:
# Tidy up environment..
rm(MF,
   BP,
   BP_MF_Terms,
   matching_file,
   DEGs,
   merged_df,
   bracketed_rows
)

# Save Environment
save.image(file = "C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 3/RData/ORA_prep.RData")

```

# Overrepresentation Analysis

```{r ORA}
  # Load environment for analysis
  load("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 3/RData/ORA_prep.RData")

## Looping script to perform all ORAs ----
run_ORA_loop <- function(groups, out_dir = "../Results", assign_to_env = TRUE, env = .GlobalEnv) {
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
  res_list <- list()
  
  safe_enricher <- function(degs, TERM2GENE, TERM2NAME, label) {
    tryCatch({
      set.seed(100)
      obj <- enricher(degs, TERM2GENE = TERM2GENE, TERM2NAME = TERM2NAME, minGSSize = 1)
      res <- obj@result
      if (!is.null(res) && nrow(res) > 0) res$GO <- label
      res
    }, error = function(e) {
      warning(sprintf("ORA failed for '%s': %s", label, conditionMessage(e)))
      NULL
    })
  }
  safe_convert <- function(x) tryCatch(convertFractionToNumeric(x), error = function(e) NA_real_)
  
  for (g in groups) {
    tryCatch({
      deg_var <- paste0(g, "_DEGs")
      if (!exists(deg_var)) {
        warning(sprintf("Skipping '%s' — object '%s' not found.", g, deg_var))
        res_list[[g]] <- NULL
        next
      }
      degs <- get(deg_var, inherits = TRUE)
      
      bp_res <- safe_enricher(degs, GENES_BP, TERMs_BP, "biological_processes")
      mf_res <- safe_enricher(degs, GENES_MF, TERMs_MF, "molecular_functions")
      pieces <- Filter(Negate(is.null), list(bp_res, mf_res))
      if (length(pieces) == 0) {
        warning(sprintf("No ORA results produced for '%s'.", g))
        res_list[[g]] <- NULL
        next
      }
      
      res <- tryCatch(do.call(rbind, pieces), error = function(e) { warning(sprintf("Combine fail '%s': %s", g, e$message)); NULL })
      if (is.null(res) || nrow(res) == 0) { res_list[[g]] <- NULL; next }
      
      if ("p.adjust" %in% names(res)) res <- res[!is.na(res$p.adjust) & res$p.adjust < 0.05, , drop = FALSE]
      if (nrow(res) == 0) { warning(sprintf("No significant terms for '%s'.", g)); res_list[[g]] <- res; next }
      
      if ("GeneRatio" %in% names(res)) res$GeneRatio <- sapply(res$GeneRatio, safe_convert)
      if ("BgRatio"   %in% names(res)) res$BgRatio   <- sapply(res$BgRatio,   safe_convert)
      
      outfile <- file.path(out_dir, sprintf("%s_ORA_results.csv", g))
      tryCatch(write.csv(res, outfile, row.names = FALSE),
               error = function(e) warning(sprintf("Write fail '%s': %s", g, e$message)))
      
      res_list[[g]] <- res
      
      if (assign_to_env) {
        obj_name <- sprintf("%s_ORA_results", g)
        assign(obj_name, res, envir = env)
      }
    }, error = function(e) {
      warning(sprintf("Unexpected error in group '%s': %s", g, conditionMessage(e)))
      res_list[[g]] <- NULL
    })
  }
  
  return(res_list)
}
groups <- c("diesel_only_Wald",
            "salinity_only_Wald",
            "interactive_only_Wald",
            "diesel_added_Wald",
            "salinity_added_Wald",
            "combined_Wald") 

res <- run_ORA_loop(groups, assign_to_env = TRUE)
```


# Gene Set Enrichment Analysis
```{r GSEA}
# Make loop
run_GSEA_loop <- function(groups,
                          out_dir = "../Results",
                          p_cutoff = 0.05,
                          assign_to_env = TRUE,
                          env = .GlobalEnv) {
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
  res_list <- list()
  
  # Safe GSEA runner -> returns a list(GSEA_obj = <enrichResult>, df = <data.frame>|NULL)
  safe_gsea <- function(geneList, TERM2GENE, TERM2NAME) {
    tryCatch({
      set.seed(100)
      g <- GSEA(geneList, TERM2GENE = TERM2GENE, TERM2NAME = TERM2NAME, pvalueCutoff = p_cutoff)
      df <- g@result
      list(GSEA_obj = g, df = df)
    }, error = function(e) {
      warning(sprintf("GSEA failed: %s", conditionMessage(e)))
      list(GSEA_obj = NULL, df = NULL)
    })
  }
  
  for (g in groups) {
    tryCatch({
      gl_var <- paste0(g, "_geneList")
      if (!exists(gl_var)) {
        warning(sprintf("Skipping '%s' — object '%s' not found.", g, gl_var))
        res_list[[g]] <- NULL
        next
      }
      geneList <- get(gl_var, inherits = TRUE)
      
      # Run BP and MF
      bp <- safe_gsea(geneList, GENES_BP, TERMs_BP)
      mf <- safe_gsea(geneList, GENES_MF, TERMs_MF)
      
      bp_df <- bp$df
      if (!is.null(bp_df) && nrow(bp_df) > 0) bp_df$GO <- "biological_processes"
      mf_df <- mf$df
      if (!is.null(mf_df) && nrow(mf_df) > 0) mf_df$GO <- "molecular_functions"
      
      # Merge results (skip NULLs)
      pieces <- Filter(function(x) !is.null(x) && nrow(x) > 0, list(bp_df, mf_df))
      if (length(pieces) == 0) {
        warning(sprintf("No GSEA results produced for '%s'.", g))
        res_list[[g]] <- NULL
        next
      }
      merged <- tryCatch({
        do.call(rbind, pieces)
      }, error = function(e) {
        warning(sprintf("Failed to combine GSEA results for '%s': %s", g, conditionMessage(e)))
        NULL
      })
      
      if (is.null(merged) || nrow(merged) == 0) {
        warning(sprintf("Empty combined GSEA results for '%s'.", g))
        res_list[[g]] <- NULL
        next
      }
      
      # Filter significant results
      if ("p.adjust" %in% names(merged)) {
        merged <- merged[!is.na(merged$p.adjust) & merged$p.adjust < p_cutoff, , drop = FALSE]
      } else {
        warning(sprintf("'p.adjust' column missing for '%s'; skipping significance filter.", g))
      }
      
      # Package like your example: list(BP, MF, Results)
      group_list <- list(
        BP = bp$GSEA_obj,
        MF = mf$GSEA_obj,
        Results = merged
      )
      
      # Write CSV
      outfile <- file.path(out_dir, sprintf("%s_GSEA_results.csv", g))
      tryCatch({
        write.csv(group_list$Results, outfile, row.names = FALSE)
      }, error = function(e) {
        warning(sprintf("Failed to write CSV for '%s' to '%s': %s", g, outfile, conditionMessage(e)))
      })
      
      res_list[[g]] <- group_list
      
      # Optional: assign to environment for easy inspection
      if (assign_to_env) {
        assign(sprintf("%s_GSEA", g), group_list, envir = env)
        assign(sprintf("%s_GSEA_results", g), group_list$Results, envir = env)
      }
    }, error = function(e) {
      warning(sprintf("Unexpected error in group '%s': %s", g, conditionMessage(e)))
      res_list[[g]] <- NULL
    })
  }
  
  return(res_list)  # visible return
}

# Run loop
groups <- c("diesel_only","salinity_only","interactive_only","diesel_added","salinity_added","combined")
set.seed(100)
gsea_out <- run_GSEA_loop(groups, p_cutoff = 0.05, assign_to_env = TRUE)

# Save image
save.image(file = "C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 3/RData/ORA_and_GSEA_Wald_completed.RData")
```

# Combine and Save results files ----

## ORA
```{r combined_save_ORA}
# results files names:
  # diesel_only_Wald_ORA_results
  # diesel_added_Wald_ORA_results
  # salinity_only_Wald_ORA_results
  # salinity_added_Wald_ORA_results
  # combined_Wald_ORA_results
  # interactive_only_Wald_ORA_results

# Add contrast column so differentiate between contrasts in the same file:
diesel_only_Wald_ORA_results$contrast <- "diesel_only"
diesel_added_Wald_ORA_results$contrast <- "diesel_added"
salinity_only_Wald_ORA_results$contrast <- "salinity_only"
salinity_added_Wald_ORA_results$contrast <- "salinity_added"
combined_Wald_ORA_results$contrast <- "combined"
interactive_only_Wald_ORA_results$contrast <- "interactive"

# merge
all_ORA_results <- bind_rows(
  diesel_only_Wald_ORA_results,
  diesel_added_Wald_ORA_results,
  salinity_only_Wald_ORA_results,
  salinity_added_Wald_ORA_results,
  combined_Wald_ORA_results,
  interactive_only_Wald_ORA_results
)

# Write.csv
write.csv(all_ORA_results, "../Results/MS_all_ORA_results.csv", row.names = F)

# load QuickGO version of the file
ORA_results <- read.csv("../Results/MS_all_ORA_results_QuickGO.csv", header = T)
```

## GSEA
```{r combined_save_GSEA}
# Adding contrast column
# diesel_only_GSEA_results$contrast <- "diesel_only" # No Results
salinity_only_GSEA_results$contrast <- "salinity_only"
interactive_only_GSEA_results$contrast <- "interactive"
diesel_added_GSEA_results$contrast <- "diesel_added"
salinity_added_GSEA_results$contrast <- "salinity_added"
combined_GSEA_results$contrast <- "additive"

# Combine
all_GSEA_results <- bind_rows(#diesel_only_GSEA_results,
                         salinity_only_GSEA_results,
                         interactive_only_GSEA_results,
                         diesel_added_GSEA_results,
                         salinity_added_GSEA_results,
                         combined_GSEA_results
)

# Save
write.csv(all_GSEA_results, "../Results/MS_all_GSEA_results.csv", row.names = F)

# Load QuickGO version
GSEA_results <- read.csv("../Results/MS_all_GSEA_results_QuickGO.csv", header = T)
```

# Plots

## ORA plots

### Attach LFC data to GO data
```{r ORA_plots}
# only significant results
ORA_results <- subset(ORA_results[ORA_results$p.adjust < 0.05,])
ORA_results <- ORA_results[!apply(is.na(ORA_results), 1, all), ]

# empty df to add data to:
ORA_plotting_data <- tibble(
  term = character(),
  gene = character(),
  lfc = numeric(),
  contrast = factor()
)

# Diesel only
{
data_temp <- subset(DE_results[, c("gene", "padj_diesel_only_wald","LFC_diesel_only_wald")])

  # change these if your column names differ
  lfc_col        <- "LFC_diesel_only_wald"  # shrunken LFC column in DEA file
  gene_col       <- "gene"            
  
  # ORA columns 
  ora_df         <- subset(ORA_results[ORA_results$contrast == "diesel_only",]) #                     # Change this to new contrast
  ora_term_col   <- if ("Description_QuickGO" %in% names(ora_df)) "Description_QuickGO" else "ID"
  ora_padj_col   <- if ("p.adjust"   %in% names(ora_df)) "p.adjust"   else "pvalue"  # fallback
  ora_genes_col  <- "geneID"         # nested gene list in each row
  

  lfc_map <- data_temp %>%
    transmute(gene = .data[[gene_col]],
              lfc  = .data[[lfc_col]]) %>%
    distinct()
  

  attach_lfc <- function(df, term_col, genes_col, lfc_tbl,
                         split_pattern = "[,;/\\|\\s]+") {
    out <- df %>%
      transmute(term  = .data[[term_col]],
                genes = .data[[genes_col]]) %>%
      mutate(gene = str_split(genes, split_pattern)) %>%
      select(-genes) %>%
      unnest(gene) %>%
      mutate(gene = str_trim(gene)) %>%
      filter(!is.na(gene), gene != "") %>%
      distinct(term, gene) %>%  
      left_join(lfc_tbl, by = "gene")
    out
  }
  

  ora_long  <- attach_lfc(ora_df,  ora_term_col,  ora_genes_col,  lfc_map)

  ora_terms  <- ora_df  %>% arrange(.data[[ora_padj_col]]) %>% pull(!!ora_term_col)
  
  ora_plotdf  <- ora_long  %>% filter(term %in% ora_terms) %>% mutate(contrast = "diesel_only")   # Change contrast name here

  
  # order terms inside each panel by median LFC (you can change to FDR ordering)
  ora_levels  <- ora_plotdf  %>% group_by(term)  %>% summarise(med = median(lfc, na.rm = TRUE))  %>%
    arrange(med) %>% pull(term)

  
  ora_plotdf$term  <- factor(ora_plotdf$term,  levels = ora_levels)

  ORA_plotting_data <- bind_rows(ora_plotdf, ORA_plotting_data)
}

# salinity only
{
data_temp <- subset(DE_results[, c("gene", "padj_salinity_only_wald","LFC_salinity_only_wald")])

  # change these if your column names differ
  lfc_col        <- "LFC_salinity_only_wald"  # shrunken LFC column in DEA file
  gene_col       <- "gene"            
  
  # ORA columns 
  ora_df         <- subset(ORA_results[ORA_results$contrast == "salinity_only",]) #                     # Change this to new contrast
  ora_term_col   <- if ("Description_QuickGO" %in% names(ora_df)) "Description_QuickGO" else "ID"
  ora_padj_col   <- if ("p.adjust"   %in% names(ora_df)) "p.adjust"   else "pvalue"  # fallback
  ora_genes_col  <- "geneID"         # nested gene list in each row
  

  lfc_map <- data_temp %>%
    transmute(gene = .data[[gene_col]],
              lfc  = .data[[lfc_col]]) %>%
    distinct()
  

  attach_lfc <- function(df, term_col, genes_col, lfc_tbl,
                         split_pattern = "[,;/\\|\\s]+") {
    out <- df %>%
      transmute(term  = .data[[term_col]],
                genes = .data[[genes_col]]) %>%
      mutate(gene = str_split(genes, split_pattern)) %>%
      select(-genes) %>%
      unnest(gene) %>%
      mutate(gene = str_trim(gene)) %>%
      filter(!is.na(gene), gene != "") %>%
      distinct(term, gene) %>%  
      left_join(lfc_tbl, by = "gene")
    out
  }
  

  ora_long  <- attach_lfc(ora_df,  ora_term_col,  ora_genes_col,  lfc_map)

  ora_terms  <- ora_df  %>% arrange(.data[[ora_padj_col]]) %>% pull(!!ora_term_col)
  
  ora_plotdf  <- ora_long  %>% filter(term %in% ora_terms) %>% mutate(contrast = "salinity_only")   # Change contrast name here

  
  # order terms inside each panel by median LFC (you can change to FDR ordering)
  ora_levels  <- ora_plotdf  %>% group_by(term)  %>% summarise(med = median(lfc, na.rm = TRUE))  %>%
    arrange(med) %>% pull(term)

  
  ora_plotdf$term  <- factor(ora_plotdf$term,  levels = ora_levels)

  ORA_plotting_data <- bind_rows(ORA_plotting_data, ora_plotdf)
}

# Diesel added
{
data_temp <- subset(DE_results[, c("gene", "padj_diesel_added_wald","LFC_diesel_added_wald")])

  # change these if your column names differ
  lfc_col        <- "LFC_diesel_added_wald"  # shrunken LFC column in DEA file
  gene_col       <- "gene"            
  
  # ORA columns 
  ora_df         <- subset(ORA_results[ORA_results$contrast == "diesel_added",]) #                     # Change this to new contrast
  ora_term_col   <- if ("Description_QuickGO" %in% names(ora_df)) "Description_QuickGO" else "ID"
  ora_padj_col   <- if ("p.adjust"   %in% names(ora_df)) "p.adjust"   else "pvalue"  # fallback
  ora_genes_col  <- "geneID"         # nested gene list in each row
  

  lfc_map <- data_temp %>%
    transmute(gene = .data[[gene_col]],
              lfc  = .data[[lfc_col]]) %>%
    distinct()
  

  attach_lfc <- function(df, term_col, genes_col, lfc_tbl,
                         split_pattern = "[,;/\\|\\s]+") {
    out <- df %>%
      transmute(term  = .data[[term_col]],
                genes = .data[[genes_col]]) %>%
      mutate(gene = str_split(genes, split_pattern)) %>%
      select(-genes) %>%
      unnest(gene) %>%
      mutate(gene = str_trim(gene)) %>%
      filter(!is.na(gene), gene != "") %>%
      distinct(term, gene) %>%  
      left_join(lfc_tbl, by = "gene")
    out
  }
  

  ora_long  <- attach_lfc(ora_df,  ora_term_col,  ora_genes_col,  lfc_map)

  ora_terms  <- ora_df  %>% arrange(.data[[ora_padj_col]]) %>% pull(!!ora_term_col)
  
  ora_plotdf  <- ora_long  %>% filter(term %in% ora_terms) %>% mutate(contrast = "diesel_added")   # Change contrast name here

  
  # order terms inside each panel by median LFC (you can change to FDR ordering)
  ora_levels  <- ora_plotdf  %>% group_by(term)  %>% summarise(med = median(lfc, na.rm = TRUE))  %>%
    arrange(med) %>% pull(term)

  
  ora_plotdf$term  <- factor(ora_plotdf$term,  levels = ora_levels)

  ORA_plotting_data <- bind_rows(ora_plotdf, ORA_plotting_data)
}

# salinity added
{
data_temp <- subset(DE_results[, c("gene", "padj_salinity_added_wald","LFC_salinity_added_wald")])

  # change these if your column names differ
  lfc_col        <- "LFC_salinity_added_wald"  # shrunken LFC column in DEA file
  gene_col       <- "gene"            
  
 # ORA columns
ora_df         <- subset(ORA_results[ORA_results$contrast == "salinity_added",])
ora_term_col   <- if ("Description_QuickGO" %in% names(ora_df)) "Description_QuickGO" else "ID"
ora_padj_col   <- if ("p.adjust"   %in% names(ora_df)) "p.adjust"   else "pvalue"  # fallback
ora_genes_col  <- "geneID"         # nested gene list in each row

  
lfc_map <- data_temp %>%
  transmute(gene = .data[[gene_col]],
            lfc  = .data[[lfc_col]]) %>%
  distinct()


attach_lfc <- function(df, term_col, genes_col, lfc_tbl,
                       split_pattern = "[,;/\\|\\s]+") {
  out <- df %>%
    transmute(term  = .data[[term_col]],
              genes = .data[[genes_col]]) %>%
    mutate(gene = str_split(genes, split_pattern)) %>%
    select(-genes) %>%
    unnest(gene) %>%
    mutate(gene = str_trim(gene)) %>%
    filter(!is.na(gene), gene != "") %>%
    distinct(term, gene) %>%                      # drop duplicates within a term
    left_join(lfc_tbl, by = "gene")
  out
}


ora_long  <- attach_lfc(ora_df,  ora_term_col,  ora_genes_col,  lfc_map)
  

  ora_long  <- attach_lfc(ora_df,  ora_term_col,  ora_genes_col,  lfc_map)

  ora_terms  <- ora_df  %>% arrange(.data[[ora_padj_col]]) %>% pull(!!ora_term_col)
  
  ora_plotdf  <- ora_long  %>% filter(term %in% ora_terms) %>% mutate(contrast = "salinity_added")   # Change contrast name here

  
  # order terms inside each panel by median LFC (you can change to FDR ordering)
  ora_levels  <- ora_plotdf  %>% group_by(term)  %>% summarise(med = median(lfc, na.rm = TRUE))  %>%
    arrange(med) %>% pull(term)

  
  ora_plotdf$term  <- factor(ora_plotdf$term,  levels = ora_levels)

  ORA_plotting_data <- bind_rows(ORA_plotting_data, ora_plotdf)
}

# combined
{
data_temp <- subset(DE_results[, c("gene", "padj_combined_wald","LFC_combined_wald")])

  # change these if your column names differ
  lfc_col        <- "LFC_combined_wald"  # shrunken LFC column in DEA file
  gene_col       <- "gene"            
  
 # ORA columns
ora_df         <- subset(ORA_results[ORA_results$contrast == "combined",])
ora_term_col   <- if ("Description_QuickGO" %in% names(ora_df)) "Description_QuickGO" else "ID"
ora_padj_col   <- if ("p.adjust"   %in% names(ora_df)) "p.adjust"   else "pvalue"  # fallback
ora_genes_col  <- "geneID"         # nested gene list in each row

  
lfc_map <- data_temp %>%
  transmute(gene = .data[[gene_col]],
            lfc  = .data[[lfc_col]]) %>%
  distinct()


attach_lfc <- function(df, term_col, genes_col, lfc_tbl,
                       split_pattern = "[,;/\\|\\s]+") {
  out <- df %>%
    transmute(term  = .data[[term_col]],
              genes = .data[[genes_col]]) %>%
    mutate(gene = str_split(genes, split_pattern)) %>%
    select(-genes) %>%
    unnest(gene) %>%
    mutate(gene = str_trim(gene)) %>%
    filter(!is.na(gene), gene != "") %>%
    distinct(term, gene) %>%                      # drop duplicates within a term
    left_join(lfc_tbl, by = "gene")
  out
}


ora_long  <- attach_lfc(ora_df,  ora_term_col,  ora_genes_col,  lfc_map)
  

  ora_long  <- attach_lfc(ora_df,  ora_term_col,  ora_genes_col,  lfc_map)

  ora_terms  <- ora_df  %>% arrange(.data[[ora_padj_col]]) %>% pull(!!ora_term_col)
  
  ora_plotdf  <- ora_long  %>% filter(term %in% ora_terms) %>% mutate(contrast = "combined")   # Change contrast name here

  
  # order terms inside each panel by median LFC (you can change to FDR ordering)
  ora_levels  <- ora_plotdf  %>% group_by(term)  %>% summarise(med = median(lfc, na.rm = TRUE))  %>%
    arrange(med) %>% pull(term)

  
  ora_plotdf$term  <- factor(ora_plotdf$term,  levels = ora_levels)

  ORA_plotting_data <- bind_rows(ORA_plotting_data, ora_plotdf)
}

# interactive
{
data_temp <- subset(DE_results[, c("gene", "padj_interactive_only_wald","LFC_interactive_only_wald")])

  # change these if your column names differ
  lfc_col        <- "LFC_interactive_only_wald"  # shrunken LFC column in DEA file
  gene_col       <- "gene"            
  
 # ORA columns
ora_df         <- subset(ORA_results[ORA_results$contrast == "interactive",])
ora_term_col   <- if ("Description_QuickGO" %in% names(ora_df)) "Description_QuickGO" else "ID"
ora_padj_col   <- if ("p.adjust"   %in% names(ora_df)) "p.adjust"   else "pvalue"  # fallback
ora_genes_col  <- "geneID"         # nested gene list in each row

  
lfc_map <- data_temp %>%
  transmute(gene = .data[[gene_col]],
            lfc  = .data[[lfc_col]]) %>%
  distinct()


attach_lfc <- function(df, term_col, genes_col, lfc_tbl,
                       split_pattern = "[,;/\\|\\s]+") {
  out <- df %>%
    transmute(term  = .data[[term_col]],
              genes = .data[[genes_col]]) %>%
    mutate(gene = str_split(genes, split_pattern)) %>%
    select(-genes) %>%
    unnest(gene) %>%
    mutate(gene = str_trim(gene)) %>%
    filter(!is.na(gene), gene != "") %>%
    distinct(term, gene) %>%                      # drop duplicates within a term
    left_join(lfc_tbl, by = "gene")
  out
}


ora_long  <- attach_lfc(ora_df,  ora_term_col,  ora_genes_col,  lfc_map)
  

  ora_long  <- attach_lfc(ora_df,  ora_term_col,  ora_genes_col,  lfc_map)

  ora_terms  <- ora_df  %>% arrange(.data[[ora_padj_col]]) %>% pull(!!ora_term_col)
  
  ora_plotdf  <- ora_long  %>% filter(term %in% ora_terms) %>% mutate(contrast = "interactive")   # Change contrast name here

  
  # order terms inside each panel by median LFC (you can change to FDR ordering)
  ora_levels  <- ora_plotdf  %>% group_by(term)  %>% summarise(med = median(lfc, na.rm = TRUE))  %>%
    arrange(med) %>% pull(term)

  
  ora_plotdf$term  <- factor(ora_plotdf$term,  levels = ora_levels)

  ORA_plotting_data <- bind_rows(ORA_plotting_data, ora_plotdf)
}
```


### Plot ORA

```{r image_load_ORA}
load("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 3/RData/MS_Enrichment_Analyses.RData")
```

```{r plot_ORA_results}
# 
library(dplyr)
library(ggplot2)
library(ggridges)
library(stringr)
library(forcats)
library(scales)

# Options
plot_df <- ORA_plotting_data %>%
  group_by(term) %>%
  summarise(
    n_contrasts = n_distinct(contrast),
    mean_abs_lfc = mean(abs(lfc), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(n_contrasts), desc(mean_abs_lfc), term) %>%
  pull(term) -> term_levels

plot_df <- ORA_plotting_data %>%
  mutate(
    term = factor(term, levels = rev(term_levels)),
    contrast = factor(
      contrast,
      levels = c(
        "diesel_only",
        "diesel_added",
        "salinity_only",
        "salinity_added",
        "combined",
        "interactive"
      )
    )
  )

# code to modify:
  p_boxes <- ggplot(plot_df %>% filter(!is.na(lfc)),
                    aes(x = lfc, 
                        y = term, 
                        #fill = contrast, 
                        #color = contrast
                        )) +
    geom_boxplot(outlier.size = 0.6, width = 0.6, 
                 color = "grey30", 
                 fill = "grey88"
                 ) +
    geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.6, col = "grey") +
    labs(x = "Shrunken log2 fold-change (LFC)", y = NULL) +
    facet_grid(
      cols = vars(contrast),
      scales = "free",
      labeller = labeller(
        contrast = c(
          diesel_only = "Diesel",
          diesel_added = "+ Diesel",
          salinity_only = "Salinity",
          salinity_added = "+ Salinity",
          additive = "Additive",
          interactive = "Interactive"
        )
      )
    ) +
    #xlim(-25,25) +
    theme_bw(base_size = 8) +
    theme(strip.background = element_rect(fill = "white"),
          axis.text.y = element_text(size = 6))

p_boxes

# Save plot
ggsave(
  "C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 3/Figures/Main Figures/ORA_boxplot.png",
  p_boxes,
  width = 246,
  height = 159,
  units = "mm",
  dpi = 300
)
```

## GSEA plots

## attach lfc data

```{r GSEA_lfc_data}
# only significant results
GSEA_results <- subset(GSEA_results[GSEA_results$p.adjust < 0.05,])
GSEA_results <- GSEA_results[!apply(is.na(GSEA_results), 1, all), ]

# empty df to add data to:
GSEA_plotting_data <- tibble(
  term = character(),
  gene = character(),
  lfc = numeric(),
  contrast = factor()
)

# No diesel only results, skip...

# salinity only
{
data_temp <- subset(DE_results[, c("gene", "padj_salinity_only_wald","LFC_salinity_only_wald")])

  # change these if your column names differ
  lfc_col        <- "LFC_salinity_only_wald"  # shrunken LFC column in DEA file
  gene_col       <- "gene"            
  
  # GSEA columns 
  GSEA_df         <- subset(GSEA_results[GSEA_results$contrast == "salinity_only",]) #              
  GSEA_term_col   <- if ("Description_QuickGO" %in% names(GSEA_df)) "Description_QuickGO" else "ID"
  GSEA_padj_col   <- if ("p.adjust"   %in% names(GSEA_df)) "p.adjust"   else "pvalue"  # fallback
  GSEA_genes_col <- "core_enrichment"  # nested gene list (leading edge)
  

  lfc_map <- data_temp %>%
    transmute(gene = .data[[gene_col]],
              lfc  = .data[[lfc_col]]) %>%
    distinct()
  

  attach_lfc <- function(df, term_col, genes_col, lfc_tbl,
                         split_pattern = "[,;/\\|\\s]+") {
    out <- df %>%
      transmute(term  = .data[[term_col]],
                genes = .data[[genes_col]]) %>%
      mutate(gene = str_split(genes, split_pattern)) %>%
      select(-genes) %>%
      unnest(gene) %>%
      mutate(gene = str_trim(gene)) %>%
      filter(!is.na(gene), gene != "") %>%
      distinct(term, gene) %>%  
      left_join(lfc_tbl, by = "gene")
    out
  }
  

  GSEA_long  <- attach_lfc(GSEA_df,  GSEA_term_col,  GSEA_genes_col,  lfc_map)

  GSEA_terms  <- GSEA_df  %>% arrange(.data[[GSEA_padj_col]]) %>% pull(!!GSEA_term_col)
  
  GSEA_plotdf  <- GSEA_long  %>% filter(term %in% GSEA_terms) %>% mutate(contrast = "salinity_only")   # Change contrast name here

  
  # order terms inside each panel by median LFC (you can change to FDR ordering)
  GSEA_levels  <- GSEA_plotdf  %>% group_by(term)  %>% summarise(med = median(lfc, na.rm = TRUE))  %>%
    arrange(med) %>% pull(term)

  
  GSEA_plotdf$term  <- factor(GSEA_plotdf$term,  levels = GSEA_levels)

  GSEA_plotting_data <- bind_rows(GSEA_plotting_data, GSEA_plotdf)
}

# Diesel added
{
data_temp <- subset(DE_results[, c("gene", "padj_diesel_added_wald","LFC_diesel_added_wald")])

  # change these if your column names differ
  lfc_col        <- "LFC_diesel_added_wald"  # shrunken LFC column in DEA file
  gene_col       <- "gene"            
  
  # GSEA columns 
  GSEA_df         <- subset(GSEA_results[GSEA_results$contrast == "diesel_added",]) #                     # Change this to new contrast
  GSEA_term_col   <- if ("Description_QuickGO" %in% names(GSEA_df)) "Description_QuickGO" else "ID"
  GSEA_padj_col   <- if ("p.adjust"   %in% names(GSEA_df)) "p.adjust"   else "pvalue"  # fallback
  GSEA_genes_col <- "core_enrichment"  # nested gene list (leading edge)
  

  lfc_map <- data_temp %>%
    transmute(gene = .data[[gene_col]],
              lfc  = .data[[lfc_col]]) %>%
    distinct()
  

  attach_lfc <- function(df, term_col, genes_col, lfc_tbl,
                         split_pattern = "[,;/\\|\\s]+") {
    out <- df %>%
      transmute(term  = .data[[term_col]],
                genes = .data[[genes_col]]) %>%
      mutate(gene = str_split(genes, split_pattern)) %>%
      select(-genes) %>%
      unnest(gene) %>%
      mutate(gene = str_trim(gene)) %>%
      filter(!is.na(gene), gene != "") %>%
      distinct(term, gene) %>%  
      left_join(lfc_tbl, by = "gene")
    out
  }
  

  GSEA_long  <- attach_lfc(GSEA_df,  GSEA_term_col,  GSEA_genes_col,  lfc_map)

  GSEA_terms  <- GSEA_df  %>% arrange(.data[[GSEA_padj_col]]) %>% pull(!!GSEA_term_col)
  
  GSEA_plotdf  <- GSEA_long  %>% filter(term %in% GSEA_terms) %>% mutate(contrast = "diesel_added")   # Change contrast name here

  
  # order terms inside each panel by median LFC (you can change to FDR ordering)
  GSEA_levels  <- GSEA_plotdf  %>% group_by(term)  %>% summarise(med = median(lfc, na.rm = TRUE))  %>%
    arrange(med) %>% pull(term)

  
  GSEA_plotdf$term  <- factor(GSEA_plotdf$term,  levels = GSEA_levels)

  GSEA_plotting_data <- bind_rows(GSEA_plotdf, GSEA_plotting_data)
}

# salinity added
{
data_temp <- subset(DE_results[, c("gene", "padj_salinity_added_wald","LFC_salinity_added_wald")])

  # change these if your column names differ
  lfc_col        <- "LFC_salinity_added_wald"  # shrunken LFC column in DEA file
  gene_col       <- "gene"            
  
 # GSEA columns
GSEA_df         <- subset(GSEA_results[GSEA_results$contrast == "salinity_added",])
GSEA_term_col   <- if ("Description_QuickGO" %in% names(GSEA_df)) "Description_QuickGO" else "ID"
GSEA_padj_col   <- if ("p.adjust"   %in% names(GSEA_df)) "p.adjust"   else "pvalue"  # fallback
GSEA_genes_col <- "core_enrichment"  # nested gene list (leading edge)

  
lfc_map <- data_temp %>%
  transmute(gene = .data[[gene_col]],
            lfc  = .data[[lfc_col]]) %>%
  distinct()


attach_lfc <- function(df, term_col, genes_col, lfc_tbl,
                       split_pattern = "[,;/\\|\\s]+") {
  out <- df %>%
    transmute(term  = .data[[term_col]],
              genes = .data[[genes_col]]) %>%
    mutate(gene = str_split(genes, split_pattern)) %>%
    select(-genes) %>%
    unnest(gene) %>%
    mutate(gene = str_trim(gene)) %>%
    filter(!is.na(gene), gene != "") %>%
    distinct(term, gene) %>%                      # drop duplicates within a term
    left_join(lfc_tbl, by = "gene")
  out
}


GSEA_long  <- attach_lfc(GSEA_df,  GSEA_term_col,  GSEA_genes_col,  lfc_map)
  

  GSEA_long  <- attach_lfc(GSEA_df,  GSEA_term_col,  GSEA_genes_col,  lfc_map)

  GSEA_terms  <- GSEA_df  %>% arrange(.data[[GSEA_padj_col]]) %>% pull(!!GSEA_term_col)
  
  GSEA_plotdf  <- GSEA_long  %>% filter(term %in% GSEA_terms) %>% mutate(contrast = "salinity_added")   # Change contrast name here

  
  # order terms inside each panel by median LFC (you can change to FDR ordering)
  GSEA_levels  <- GSEA_plotdf  %>% group_by(term)  %>% summarise(med = median(lfc, na.rm = TRUE))  %>%
    arrange(med) %>% pull(term)

  
  GSEA_plotdf$term  <- factor(GSEA_plotdf$term,  levels = GSEA_levels)

  GSEA_plotting_data <- bind_rows(GSEA_plotting_data, GSEA_plotdf)
}

# additive
{
data_temp <- subset(DE_results[, c("gene", "padj_combined_wald","LFC_combined_wald")]) # slight mis-match in names: additive = combined

  # change these if your column names differ
  lfc_col        <- "LFC_combined_wald"  # shrunken LFC column in DEA file
  gene_col       <- "gene"            
  
 # GSEA columns
GSEA_df         <- subset(GSEA_results[GSEA_results$contrast == "additive",])
GSEA_term_col   <- if ("Description_QuickGO" %in% names(GSEA_df)) "Description_QuickGO" else "ID"
GSEA_padj_col   <- if ("p.adjust"   %in% names(GSEA_df)) "p.adjust"   else "pvalue"  # fallback
GSEA_genes_col <- "core_enrichment"  # nested gene list (leading edge)

  
lfc_map <- data_temp %>%
  transmute(gene = .data[[gene_col]],
            lfc  = .data[[lfc_col]]) %>%
  distinct()


attach_lfc <- function(df, term_col, genes_col, lfc_tbl,
                       split_pattern = "[,;/\\|\\s]+") {
  out <- df %>%
    transmute(term  = .data[[term_col]],
              genes = .data[[genes_col]]) %>%
    mutate(gene = str_split(genes, split_pattern)) %>%
    select(-genes) %>%
    unnest(gene) %>%
    mutate(gene = str_trim(gene)) %>%
    filter(!is.na(gene), gene != "") %>%
    distinct(term, gene) %>%                      # drop duplicates within a term
    left_join(lfc_tbl, by = "gene")
  out
}


GSEA_long  <- attach_lfc(GSEA_df,  GSEA_term_col,  GSEA_genes_col,  lfc_map)
  

  GSEA_long  <- attach_lfc(GSEA_df,  GSEA_term_col,  GSEA_genes_col,  lfc_map)

  GSEA_terms  <- GSEA_df  %>% arrange(.data[[GSEA_padj_col]]) %>% pull(!!GSEA_term_col)
  
  GSEA_plotdf  <- GSEA_long  %>% filter(term %in% GSEA_terms) %>% mutate(contrast = "additive")   # Change contrast name here

  
  # order terms inside each panel by median LFC (you can change to FDR ordering)
  GSEA_levels  <- GSEA_plotdf  %>% group_by(term)  %>% summarise(med = median(lfc, na.rm = TRUE))  %>%
    arrange(med) %>% pull(term)

  
  GSEA_plotdf$term  <- factor(GSEA_plotdf$term,  levels = GSEA_levels)

  GSEA_plotting_data <- bind_rows(GSEA_plotting_data, GSEA_plotdf)
}

# interactive
{
data_temp <- subset(DE_results[, c("gene", "padj_interactive_only_wald","LFC_interactive_only_wald")])

  # change these if your column names differ
  lfc_col        <- "LFC_interactive_only_wald"  # shrunken LFC column in DEA file
  gene_col       <- "gene"            
  
 # GSEA columns
GSEA_df         <- subset(GSEA_results[GSEA_results$contrast == "interactive",])
GSEA_term_col   <- if ("Description_QuickGO" %in% names(GSEA_df)) "Description_QuickGO" else "ID"
GSEA_padj_col   <- if ("p.adjust"   %in% names(GSEA_df)) "p.adjust"   else "pvalue"  # fallback
GSEA_genes_col <- "core_enrichment"  # nested gene list (leading edge)

  
lfc_map <- data_temp %>%
  transmute(gene = .data[[gene_col]],
            lfc  = .data[[lfc_col]]) %>%
  distinct()


attach_lfc <- function(df, term_col, genes_col, lfc_tbl,
                       split_pattern = "[,;/\\|\\s]+") {
  out <- df %>%
    transmute(term  = .data[[term_col]],
              genes = .data[[genes_col]]) %>%
    mutate(gene = str_split(genes, split_pattern)) %>%
    select(-genes) %>%
    unnest(gene) %>%
    mutate(gene = str_trim(gene)) %>%
    filter(!is.na(gene), gene != "") %>%
    distinct(term, gene) %>%                      # drop duplicates within a term
    left_join(lfc_tbl, by = "gene")
  out
}


GSEA_long  <- attach_lfc(GSEA_df,  GSEA_term_col,  GSEA_genes_col,  lfc_map)
  

  GSEA_long  <- attach_lfc(GSEA_df,  GSEA_term_col,  GSEA_genes_col,  lfc_map)

  GSEA_terms  <- GSEA_df  %>% arrange(.data[[GSEA_padj_col]]) %>% pull(!!GSEA_term_col)
  
  GSEA_plotdf  <- GSEA_long  %>% filter(term %in% GSEA_terms) %>% mutate(contrast = "interactive")   # Change contrast name here

  
  # order terms inside each panel by median LFC (you can change to FDR ordering)
  GSEA_levels  <- GSEA_plotdf  %>% group_by(term)  %>% summarise(med = median(lfc, na.rm = TRUE))  %>%
    arrange(med) %>% pull(term)

  
  GSEA_plotdf$term  <- factor(GSEA_plotdf$term,  levels = GSEA_levels)

  GSEA_plotting_data <- bind_rows(GSEA_plotting_data, GSEA_plotdf)
}
```

### Plot GSEA

```{r image_load_ORA}
load("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 3/RData/MS_Enrichment_Analyses.RData")
```

```{r plot_ORA_results}
library(dplyr)
library(ggplot2)
library(ggridges)
library(stringr)
library(forcats)
library(scales)

plot_df <- GSEA_plotting_data %>%
  group_by(term) %>%
  summarise(
    n_contrasts = n_distinct(contrast),
    mean_abs_lfc = mean(abs(lfc), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(n_contrasts), desc(mean_abs_lfc), term) %>%
  pull(term) -> term_levels

plot_df <- GSEA_plotting_data %>%
  mutate(
    term = factor(term, levels = rev(term_levels)),
    contrast = factor(
      contrast,
      levels = c(
        "diesel_only",
        "diesel_added",
        "salinity_only",
        "salinity_added",
        "additive",
        "interactive"
      )
    )
  )

x_lim <- max(abs(plot_df$lfc), na.rm = TRUE)
hline_df <- data.frame(y = seq_along(levels(plot_df$term)))

final_GSEA_ridges <- ggplot(plot_df, aes(x = lfc, y = term)) +
  
  geom_density_ridges(
    scale = 0.9,
    rel_min_height = 0.01,
    fill = "grey85",
    colour = "grey20",
    linewidth = 0.3,
    position = position_nudge(y = -0.25)
  ) +
  
  geom_vline(
    xintercept = 0,
    linetype = 2,
    linewidth = 0.35,
    colour = "grey40"
  ) +
  
  # geom_hline(
  #   data = hline_df,
  #   aes(yintercept = y),
  #   inherit.aes = FALSE,
  #   colour = "grey85",
  #   linewidth = 0.25
  # ) +
  
  facet_grid(
    cols = vars(contrast),
    scales = "free",
    labeller = labeller(
      contrast = c(
        diesel_only = "Diesel",
        diesel_added = "+ Diesel",
        salinity_only = "Salinity",
        salinity_added = "+ Salinity",
        additive = "Additive",
        interactive = "Interactive"
      )
    )
  ) +
  
  scale_y_discrete(
    labels = function(x) stringr::str_wrap(x, width = 120)
  ) +
  
  scale_x_continuous(
    #limits = c(-x_lim, x_lim),
    breaks = pretty_breaks(n = 5),
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  
  labs(
    x = expression(log[2] * " fold-change (LFC, ashr)"),
    y = NULL
  ) +
  
  theme_bw(base_size = 10) +
  theme(
    strip.background = element_rect(fill = "grey95", colour = "black", linewidth = 0.4),
    strip.text.x = element_text(face = "bold", size = 8),
    axis.text.y = element_text(size = 5, colour = "black", lineheight = 0.9),
    axis.text.x = element_text(size = 6.5, colour = "black"),
    axis.title.x = element_text(size = 8, face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(colour = "grey85", linewidth = 0.25),
    panel.grid.major.x = element_line(colour = "grey88", linewidth = 0.25),
    panel.spacing.x = unit(0.5, "lines"),
    plot.margin = margin(5.5, 10, 5.5, 30)
  )

final_GSEA_ridges

ggsave(
  "C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 3/Figures/Main Figures/final_GSEA_ridges.png",
  final_GSEA_ridges,
  width = 265,
  height = 159,
  units = "mm",
  dpi = 1200
)
```


# Clean-up for saving environment
```{r clean_up}
rm(BP_MF_Terms,
   data_temp,
   DE_results,
   diesel_added_Wald_ORA_results,
   diesel_only_Wald_ORA_results,
   GENES_BP,
   GENES_MF,
   GSEA_df,
   GSEA_plotdf,
   matching_file,
   ora_df,
   ora_long,
   ora_plotdf,
   TERMs_BP,
   TERMs_MF,
   tpm
)

save.image("C:/Users/hamis/OneDrive/Documents/PhD/PhD---omics-analysis-portfolio/Chapter 3/RData/MS_Enrichment_Analyses.RData")
```